% =============================================================================
%  Chapter 4 — Gauss-Point to Nodal Projection Methods
% =============================================================================

\chapter{Gauss-Point to Nodal Projection}
\label{ch:projection}

In displacement-based finite elements, the primary unknowns are nodal.  The
derived fields of engineering interest\,---\,strains, stresses, and internal
variables\,---\,are instead evaluated at Gauss points.  For VTK/ParaView
visualisation, those material fields must be transferred to the mesh nodes.

\texttt{fall\_n} currently implements three nodal transfer operators:
\begin{enumerate}[nosep]
  \item lumped $L^2$ projection;
  \item polynomial patch recovery (the production fallback for higher-order
        simplex elements);
  \item volume-weighted averaging (a deliberately low-order but very robust
        explicit fallback).
\end{enumerate}

The adaptive default is now:
\begin{center}
  \fbox{\parbox{0.85\linewidth}{
  use lumped $L^2$ only when every local nodal lump is strictly positive;
  otherwise use polynomial patch recovery.}}
\end{center}


% ─────────────────────────────────────────────────────────────────────────
\section{Lumped \texorpdfstring{$L^2$}{L2} Projection}
\label{sec:l2_projection}

\subsection{Consistent \texorpdfstring{$L^2$}{L2} Projection}

Let $\sigma^h$ be the field known at Gauss points and let
$\tilde{\sigma}^h = \sum_I \tilde{\sigma}_I N_I$ be its nodal projection.
The consistent $L^2$ projection solves:
\begin{equation}
  \mathbf{M}\,\tilde{\bsigma} = \mathbf{f},
  \label{eq:consistent_L2}
\end{equation}
with
\begin{equation}
  M_{IJ} = \int_\Omega N_I N_J \,\dd\Omega,
  \qquad
  f_I = \int_\Omega N_I \sigma^h \,\dd\Omega.
\end{equation}

This is mathematically clean but operationally unattractive for
post-processing: it introduces a global solve whose cost is disproportionate
to the visualisation task.

\subsection{Row-Sum Lumping}
\label{sec:lumping}

The classical shortcut is to lump the mass matrix by row sum:
\begin{equation}
  \tilde{M}_{II}
  = \sum_J M_{IJ}
  = \int_\Omega N_I \sum_J N_J \,\dd\Omega
  = \int_\Omega N_I \,\dd\Omega.
  \label{eq:lumped_mass}
\end{equation}
Using quadrature,
\begin{equation}
  \tilde{\sigma}_I
  \approx
  \frac{\displaystyle\sum_{e \in \mathcal{E}_I}
        \sum_{g=1}^{n_g} N_I(\bxi_g)\, w_g\, |J_g|\, \sigma_g}
       {\displaystyle\sum_{e \in \mathcal{E}_I}
        \sum_{g=1}^{n_g} N_I(\bxi_g)\, w_g\, |J_g|}.
  \label{eq:lumped_L2_final}
\end{equation}

This behaves as a nodal average only if every denominator is strictly
positive.  For higher-order simplex elements that condition may fail even
under all-positive quadrature, because the exact lumped nodal measure
$\int_\Omega N_I \,\dd\Omega$ can itself be signed.  The TET10 vertex nodes
are precisely such a case, as shown in \cref{ch:adaptive_projection}.

\paragraph{Status in \texttt{fall\_n}.}
Lumped $L^2$ remains the preferred projection when it is mathematically sound.
For element families such as HEX27, where the local lumped nodal measures stay
strictly positive under the chosen quadrature, it yields the smoothest and
least diffusive nodal field at essentially zero additional complexity.


% ─────────────────────────────────────────────────────────────────────────
\section{Polynomial Patch Recovery}
\label{sec:projection_patch}

\subsection{Motivation}

When lumped $L^2$ stops being a true average, the right replacement is not a
coarser averaging operator but a \emph{local reconstruction} operator.  The
classical Superconvergent Patch Recovery (SPR) method of Zienkiewicz and
Zhu~\cite{ZienkiewiczZhu1992a,ZienkiewiczZhu1992b} fits a polynomial over the
patch of elements surrounding a node and evaluates that polynomial at the
node.  For \texttt{TET10} this is the correct architectural move: the Gauss
point data are smooth, but the nodal transfer operator must stop behaving like
signed lumping.

\subsection{Patch Definition}

For a node $I$, define its patch as the union of elements that contain that
node:
\begin{equation}
  \mathcal{P}_I = \{\, e \in \mathcal{E} \mid I \in e \,\}.
\end{equation}
All Gauss points of all elements in $\mathcal{P}_I$ are used as sample sites.
If the patch contributes observations
$\{(\bx_g, \sigma_g, \omega_g)\}_{g=1}^{n_p}$, we seek a polynomial
$p_I(\bx)$ that best fits the data in weighted least squares.

\subsection{Local Polynomial Basis}

To improve conditioning, the implementation uses centered and scaled local
coordinates around the evaluation node $\bx_I$:
\begin{equation}
  \hat{\bx}_g = \frac{\bx_g - \bx_I}{h_I},
  \qquad
  h_I = \max_{g \in \mathcal{P}_I} \|\bx_g - \bx_I\|_2.
\end{equation}

For three-dimensional continuum elements the quadratic basis is
\begin{equation}
  \mathbf{p}(\hat{\bx}) =
  \begin{bmatrix}
  1 &
  \hat{x} &
  \hat{y} &
  \hat{z} &
  \hat{x}^2 &
  \hat{x}\hat{y} &
  \hat{x}\hat{z} &
  \hat{y}^2 &
  \hat{y}\hat{z} &
  \hat{z}^2
  \end{bmatrix}^{\!\mathsf{T}}.
\end{equation}
The implementation uses the corresponding reduced bases in 1D and 2D.

\subsection{Weighted Least Squares}

The coefficients $\ba_I$ are obtained from
\begin{equation}
  \ba_I
  = \arg\min_{\ba}
    \sum_{g=1}^{n_p} \omega_g
    \left(
      \mathbf{p}(\hat{\bx}_g)^{\mathsf{T}}\ba - \sigma_g
    \right)^2.
\end{equation}
The normal equations are
\begin{equation}
  \left(
    \sum_g \omega_g\, \mathbf{p}_g \mathbf{p}_g^{\mathsf{T}}
  \right)\ba_I
  =
  \sum_g \omega_g\, \mathbf{p}_g \sigma_g.
\end{equation}

In the current implementation, the weights are
\begin{equation}
  \omega_g = |w_g J_g|,
\end{equation}
that is, the \emph{magnitude} of the quadrature measure.  This is intentional:
for patch fitting the weight is a confidence/importance measure for a sample,
not a signed coefficient of an integral approximation.  Therefore negative
quadrature weights must not flip the sign of the least-squares metric.

\subsection{Adaptive Degree Fallback}

Boundary patches or coarse patches may not contain enough independent samples
to support a quadratic fit.  The exporter therefore applies a strict
hierarchy:
\begin{enumerate}[nosep]
  \item try a quadratic polynomial;
  \item if the patch is rank-deficient, try a linear polynomial;
  \item if the patch is still insufficient, use a weighted average of the
        patch samples.
\end{enumerate}

This fallback is important for robustness near boundaries and corners.  It
keeps the high-order path where the local data support it, without making the
visualisation pipeline brittle.

\subsection{Evaluation at the Node}

Because the polynomial basis is centered at $\bx_I$, the recovered nodal value
is simply the constant coefficient:
\begin{equation}
  \tilde{\sigma}_I = p_I(\bx_I) = a_{I,0}.
\end{equation}
This is why the implementation can recover the nodal value directly after the
solve of the local least-squares system.


% ─────────────────────────────────────────────────────────────────────────
\section{Volume-Weighted Averaging}
\label{sec:projection_volume_average}

Volume-weighted averaging is retained as an explicit low-order fallback:
\begin{equation}
  \tilde{\sigma}_I
  = \frac{\displaystyle\sum_{e \in \mathcal{E}_I} V_e\, \bar{\sigma}_e}
         {\displaystyle\sum_{e \in \mathcal{E}_I} V_e},
\end{equation}
where
\begin{equation}
  \bar{\sigma}_e
  = \frac{\sum_g w_g |J_g| \sigma_g}{\sum_g w_g |J_g|}.
\end{equation}

This operator is deliberately diffusive.  Its purpose is no longer to serve as
the default production path, but to remain available when a user wants the
most conservative possible smoothing or when investigating pathological
patches.


% ─────────────────────────────────────────────────────────────────────────
\section{Implementation in \texttt{VTKModelExporter}}
\label{sec:projection_impl}

The implementation lives in \texttt{src/post-processing/VTK/VTKModelExporter.hh}.
The key pieces are:
\begin{description}
  \item[\texttt{l2\_project\_to\_nodes()}]
    Implements lumped $L^2$ projection.

  \item[\texttt{patch\_recover\_to\_nodes()}]
    Builds nodal patches from element-to-node connectivity, gathers Gauss-point
    samples, and performs centered/scaled least-squares recovery.

  \item[\texttt{spr\_average\_to\_nodes()}]
    Retained as the explicit volume-weighted fallback path.

  \item[\texttt{compute\_material\_fields()}]
    Resolves the strategy automatically:
    \begin{enumerate}[nosep]
      \item if every element has strictly positive local lumped nodal weights,
            use lumped $L^2$;
      \item otherwise use polynomial patch recovery.
    \end{enumerate}

  \item[\texttt{compute\_material\_fields\_patch()}]
    Exposes polynomial patch recovery explicitly.

  \item[\texttt{compute\_material\_fields\_spr()}]
    Preserves the old volume-average path as an explicit user choice.
\end{description}

\paragraph{Computational structure.}
The hot finite-element kernels are untouched.  All additional work is confined
to post-processing and uses only geometry already available in the element
wrappers.  This preserves the compile-time and inlining properties of the
assembly path while upgrading the quality of the visualisation path.


% ─────────────────────────────────────────────────────────────────────────
\section{Comparison of the Three Operators}
\label{sec:projection_comparison}

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{p{4.0cm}ccc}
  \toprule
  Property & Lumped $L^2$ & Patch recovery & Volume average \\
  \midrule
  Requires positive local lumps & yes & no & no \\
  Exactness on linear fields & yes & yes & no \\
  Exactness on quadratic fields & family-dependent & patch-dependent & no \\
  Continuity of rendered field & good when safe & good & fair / diffusive \\
  Sensitivity to signed simplex lumps & high & none & none \\
  Additional local solves & none & yes & none \\
  Intended use & best smooth path when safe & adaptive default for simplex & conservative fallback \\
  \bottomrule
\end{tabular}
\end{center}


% ─────────────────────────────────────────────────────────────────────────
\section{Verification}
\label{sec:projection_verification}

The implementation was verified at two levels.

\subsection{Algebraic regression}

The simplex test suite now checks:
\begin{itemize}[nosep]
  \item TET10 retains signed lumped nodal measures even under the
        positive-weight Hammer--Marlowe--Stroud rule and under the positive
        Stroud conical-product rule;
  \item polynomial patch recovery reproduces an exact linear field on a single
        TET10, which exercises the quadratic$\rightarrow$linear fallback;
  \item polynomial patch recovery reproduces an exact quadratic scalar field
        on a patch of multiple TET10 elements.
\end{itemize}

\subsection{Application-level behaviour}

On the cantilever benchmark:
\begin{itemize}[nosep]
  \item \textbf{HEX27}: the adaptive resolver chooses lumped $L^2$, and the
        resulting field remains smooth as before.
  \item \textbf{TET10}: the adaptive resolver now chooses polynomial patch
        recovery instead of volume averaging, eliminating the signed-lumping
        artefact without introducing the excessive diffusion of the old
        fallback.
\end{itemize}

This is the important practical conclusion:
\begin{quote}
The constitutive update and the element kinematics were already correct.  The
remaining ParaView artefact came from the nodal transfer operator, and the
appropriate cure was a better local reconstruction, not a change in the
stiffness quadrature.
\end{quote}
