% =============================================================================
%  Chapter 5 — PETSc 3.24.5 Patch: section.c Indexing Bug
% =============================================================================

\chapter{PETSc Patch: \texttt{section.c} Indexing Bug}
\label{ch:petsc}

This chapter documents a bug in PETSc~3.24.5~\cite{PETSc2024} that affects
higher-order finite elements whose DMPlex point numbering does not start at
zero.  The bug was discovered during the development of simplex element support
in \texttt{fall\_n} and confirmed to be an upstream issue.


% ─────────────────────────────────────────────────────────────────────────
\section{Context: DMPlex and PetscSection}
\label{sec:dmplex_context}

PETSc's \texttt{DMPlex} represents an unstructured mesh as a directed acyclic
graph (DAG) of \emph{points}: vertices, edges, faces, and cells, numbered from
$0$ to $N-1$ (though the numbering of each \emph{stratum} may start at a
non-zero offset).

A \texttt{PetscSection} associates degrees of freedom (DOFs) with mesh points.
It uses an \emph{atlas} (\texttt{s->atlasDof[]}) indexed relative to the
chart $[\texttt{pStart}, \texttt{pEnd})$, and a \emph{constraint} mechanism
to impose Dirichlet boundary conditions by specifying which DOFs at a point
are constrained.

The function \texttt{PetscSectionSetConstraintIndices()} sets the constrained
DOF indices at a given point.  Its implementation reads:
\begin{lstlisting}[language=C]
PetscCall(PetscSectionCheckConstraints_Private(s));
/* Bug: should subtract pStart to get the atlas offset */
if (s->atlasDof[point]) {              // <-- HERE
    PetscInt cdof;
    PetscCall(PetscSectionGetConstraintDof(s, point, &cdof));
    // ...
}
\end{lstlisting}


% ─────────────────────────────────────────────────────────────────────────
\section{The Bug}
\label{sec:the_bug}

In \texttt{src/sys/utils/section.c}, at approximately line~2997, the code reads:
\begin{lstlisting}[language=C,escapechar=@]
if (s->atlasDof[@\textcolor{red}{point}@]) {
\end{lstlisting}
but the atlas array is indexed relative to \texttt{pStart}, so the correct
access is:
\begin{lstlisting}[language=C,escapechar=@]
if (s->atlasDof[@\textcolor{green!50!black}{point - s->pStart}@]) {
\end{lstlisting}

When \texttt{pStart = 0} (as happens for vertex stratum when vertices are
numbered from zero), the two expressions are identical and the bug is
invisible.  However, when a mesh's vertex stratum starts at a non-zero offset
--- which occurs in \texttt{fall\_n} because Gmsh-imported meshes place cells
before vertices in the DAG, giving the vertex stratum
$\texttt{pStart} = 8$~(for 8~hex cells) or larger --- the raw \texttt{point}
index overflows the atlas array, reading uninitialised memory.


% ─────────────────────────────────────────────────────────────────────────
\section{Manifestation in \texttt{fall\_n}}
\label{sec:manifestation}

The bug manifests during the application of Dirichlet boundary conditions
via \texttt{PetscSectionSetConstraintIndices}.  Specifically:
\begin{itemize}[nosep]
  \item For HEX27 meshes with 20 cells, the vertex stratum starts at
        \texttt{pStart} $\ge$ 20.
  \item When setting constraints on vertex DOFs, the index
        \texttt{s->atlasDof[point]} reads from memory at offset 20+
        instead of offset 0+.
  \item The read value is unpredictable: it may be zero (causing the
        constraints to be silently skipped) or non-zero (proceeding correctly
        by accident).
  \item The symptom was a segmentation fault or incorrect constraint
        application on some meshes but not others, depending on memory layout.
\end{itemize}


% ─────────────────────────────────────────────────────────────────────────
\section{The Fix}
\label{sec:the_fix}

The fix is a one-line change in
\texttt{src/sys/utils/section.c}:

\begin{lstlisting}[language=C]
// Before (buggy):
if (s->atlasDof[point]) {

// After (fixed):
if (s->atlasDof[point - s->pStart]) {
\end{lstlisting}

This was applied to the local PETSc installation at:
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize,breaklines=true]
/home/sechavarriam/MyLibs/petsc-3.24.5/src/sys/utils/section.c
\end{lstlisting}

After applying the patch, PETSc was rebuilt and all \texttt{fall\_n} constraint
tests passed on both HEX27 and TET10 meshes.


% ─────────────────────────────────────────────────────────────────────────
\section{Verification}
\label{sec:petsc_verification}

The fix was verified by:
\begin{enumerate}[nosep]
  \item Running the debug output in \texttt{fall\_n}'s boundary condition
        application code, which prints:
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
[SIEVE-DBG] section_ptr=0x... chart=[9152,23629)
[SIEVE-DBG] All 14477 section DOFs OK
[BC-DBG] Constraints count: 145
[BC-DBG] plex_id=9152 section_dof=3 constraint_size=3
  ...
\end{lstlisting}
        confirming that all 145~Dirichlet-constrained vertices are correctly
        handled with \texttt{pStart} = 9152.

  \item Comparing the Case~1 (SmallStrain + Linear Elastic) displacement field
        on HEX27 against the expected Timoshenko beam solution: the
        tip displacement matches the analytical result to within the expected
        discretisation error.

  \item Running all 7~nonlinear cases on both HEX27 and TET10 meshes without
        segfaults or constraint errors.
\end{enumerate}


% ─────────────────────────────────────────────────────────────────────────
\section{Upstream Status}

As of PETSc~3.24.5, this bug is present in the release source.  It has been
reported to the PETSc developers.  Since the bug only affects codes that:
\begin{itemize}[nosep]
  \item use \texttt{PetscSectionSetConstraintIndices} directly (rather than
        through higher-level \texttt{DMPlexSetBoundaryConditions}), and
  \item have a \texttt{PetscSection} chart with \texttt{pStart} $> 0$,
\end{itemize}
it may not have been triggered by the PETSc test suite, which may use meshes
with vertex-first ordering (\texttt{pStart} = 0 for the vertex stratum).


% ─────────────────────────────────────────────────────────────────────────
\section{Workaround from Application Code}
\label{sec:workaround}

In parallel with the upstream patch, we discovered two application-level
workarounds that avoid modifying PETSc source entirely.  Both were implemented
in \texttt{fall\_n} and verified on all mesh types.

\subsection{Workaround~1: Extend the Section Chart}
\label{sec:workaround_chart}

The root cause is that \texttt{PetscSectionSetConstraintIndices()} accesses
\texttt{s->atlasDof[point]} without subtracting \texttt{s->pStart}, so it
reads out of bounds whenever the chart does not start at zero.

\paragraph{Symptom.}
For the TET4 mesh (9152~elements, 2119~nodes), the \texttt{DMPlex} point
numbering places cells at $[0, 9152)$ and vertices at $[9152, 11271)$.
When the section chart is set to the vertex stratum
$[\texttt{vStart}, \texttt{vEnd}) = [9152, 11271)$, the atlas has only
$11271 - 9152 = 2119$ entries.  The buggy code accesses index~9152 into a
2119-entry array --- a severe out-of-bounds read.

The read value is unpredictable: it reports \texttt{dof = 123} (instead of~3)
for constrained nodes, producing 604~PETSc error lines of the form:
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize,numbers=none]
[0]PETSC ERROR: Point 9152 dof 123,
    invalid constraint index[9]: 24587
\end{lstlisting}
More critically, the constraint indices are \emph{not applied}, and the
system produces a singular matrix (all Dirichlet DOFs are unconstrained),
giving \texttt{|u| = 0.000000} for TET4.

\paragraph{Fix.}
In \texttt{Model::set\_sieve\_layout()}, set the section chart to cover the
full DM chart $[\texttt{pStart}, \texttt{pEnd})$ instead of only the vertex
stratum $[\texttt{vStart}, \texttt{vEnd})$:
\begin{lstlisting}[language=C++]
// Before (triggers PETSc bug when vStart > 0):
PetscSectionSetChart(dof_section, vStart, vEnd);

// After (workaround: atlas covers all points):
PetscSectionSetChart(dof_section, pStart, pEnd);
\end{lstlisting}
With the full chart, \texttt{pStart = 0} always, so the buggy raw index
\texttt{s->atlasDof[point]} coincides with the correct adjusted index
\texttt{s->atlasDof[point - 0]}.  Element points (cells) simply carry 0~DOFs.

The cost is a marginally larger atlas array (8~bytes per cell point), which is
negligible relative to the mesh data.

\paragraph{Source.} \texttt{src/model/Model.hh}, method \texttt{set\_sieve\_layout()}.


\subsection{Workaround~2: Allow Dynamic Matrix Allocation}
\label{sec:workaround_prealloc}

A secondary class of errors surfaces on large unstructured TET4 meshes: PETSc
reports thousands of \emph{``New nonzero at (i,j) caused a malloc''}
messages during matrix assembly.  These occur because
\texttt{DMCreateMatrix()}'s internal adjacency walk
(\texttt{DMPlexPreallocateOperator}) underestimates the sparsity pattern for
cell-vertex sieves of unstructured simplex meshes.

\paragraph{Diagnosis.}
On the TET4 mesh (9152~elements, 2119~nodes), analysis confirmed that error
DOF pairs correspond to nodes that \emph{do not share any element} in the
mesh.  For example, node~172 (9~neighbours in the mesh graph) was reported as
needing an entry with node~1358 --- with zero shared elements.  This is a
PETSc preallocation-algorithm issue, not a \texttt{fall\_n} sieve-setup error.

The results were still correct because PETSc dynamically allocates the missing
entries (just slower, with error messages in debug builds).

\paragraph{Fix.}
After each \texttt{DMCreateMatrix()} call, tell PETSc to silently allocate
missing entries instead of reporting errors:
\begin{lstlisting}[language=C++]
DMCreateMatrix(dm, &K);
MatSetOption(K, MAT_NEW_NONZERO_ALLOCATION_ERR, PETSC_FALSE);
\end{lstlisting}
This was applied to all matrix creation sites:
\begin{center}
\small
\begin{tabular}{lll}
  \toprule
  File & Method & Matrix \\
  \midrule
  \texttt{Model.hh}           & \texttt{setup\_matrix()}               & \texttt{Kt\_} \\
  \texttt{LinearAnalysis.hh}  & \texttt{setup\_matrix\_sizes()}        & \texttt{K} \\
  \texttt{NLAnalysis.hh}      & SNES Jacobian callback                & \texttt{J} \\
  \texttt{DynamicAnalysis.hh} & \texttt{setup()}                       & \texttt{M\_, C\_, J\_} \\
  \texttt{DynamicAnalysis.hh} & \texttt{assemble\_initial\_stiffness} & \texttt{K0\_} \\
  \bottomrule
\end{tabular}
\end{center}


\subsection{Verification}

After applying both workarounds, \texttt{fall\_n} runs with \textbf{zero}
PETSc error messages on all four mesh types.  The results table:

\begin{center}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{lrrrrr}
  \toprule
  Model                    & Nodes & Elems & $|u|_\text{max}$ & $|u_y|_\text{max}$ & Error (\%) \\
  \midrule
  Timoshenko $N=3$ (ref.)  &    11 &     5 &  6.4518 &  6.2578 &   0.00 \\
  HEX8  (30$\times$6$\times$3)     &   868 &   540 &  4.8293 &  4.8293 &  25.15 \\
  HEX27 (30$\times$6$\times$3)     &  5551 &   540 &  6.2148 &  6.2148 &   3.67 \\
  TET4  (Fine, 4-pt)      &  2119 &  9152 &  3.5079 &  3.5079 &  45.63 \\
  TET10 (Fine, 27-pt)     &  4914 &  2475 &  6.2173 &  6.2173 &   3.64 \\
  \bottomrule
\end{tabular}
\end{center}

The TET4 result (3.51, 45.6\% error) now correctly reflects the well-known
volumetric locking and stiffness overestimation of linear tetrahedra under
bending.  Previously, the section bug caused \texttt{|u| = 0.000} (100\%
error) because boundary conditions were silently dropped.

The TET10 (3.64\% error) and HEX27 (3.67\% error) results match the
Timoshenko beam reference to within the expected discretisation error for
second-order elements.
