% ==========================================================================
%  Chapter 106 -- Energy Line Search, TAO Polish, and the FE2 Probe
% ==========================================================================

\chapter{Energy Line Search, TAO Polish, and the FE\textsuperscript{2} Probe}
\label{ch:energy-linesearch-fe2}

\paragraph{Normative status.}
Second remediation round after the Phase-I closure
(Cap.~\ref{ch:driver-integration}). Two solver-level remedies proposed
from the Phase-I evidence are implemented and measured, an energy
backtracking line search inside the LM continuation and an LM polish
stage after the hybrid TAO step, and the FE\textsuperscript{2} route
(material information passing between a beam macro and a KoBathe RVE) is
probed as a modelling-level alternative. Everything remains
environment-gated and OFF by default, and the gates-off campaign was
re-verified byte-identical to \path{data/output/g0_baseline} after every
solver change of this round.

\section{Energy line search inside the LM continuation}
\label{sec:ls-theory}

The Phase-I autopsy of deflation (Cap.~\ref{ch:driver-integration})
showed that near a reversal the spurious basins are dense, so
\emph{avoiding} a known root is not selection. The hybrid TAO step
selects by \emph{descending}, and the line search brings that principle
into the LM step itself at almost no cost. With
$\mathbf K + \mu\mathbf I$ positive definite the LM direction satisfies
\[
  \mathbf R^{\top}\,\mathrm d\mathbf u
  = -\,\mathbf R^{\top}(\mathbf K+\mu\mathbf I)^{-1}\mathbf R \;<\; 0,
\]
and since $\mathbf R = \nabla\Pi$ within a frozen-state step,
$\mathrm d\mathbf u$ is a \emph{descent direction of the incremental
potential}. A full step that raises $\Pi$ is therefore not descending
the energy landscape, it is jumping over a barrier into another basin
--- the reversal-spike mechanism. The gated rule
(\texttt{KOBATHE\_LM\_LINESEARCH=1}) is
%
\begin{enumerate}
  \item accept the step only if the residual decreases (the original
        rule) \emph{and} $\Pi$ does not increase;
  \item if $\Pi$ increases, halve $\alpha$ (up to
        \texttt{KOBATHE\_LM\_LS\_BACKTRACKS}, default 4);
  \item if no $\alpha$ lowers $\Pi$, \emph{reject} the step so $\mu$
        grows. The large-$\mu$ limit
        $\mathrm d\mathbf u \to -\mu^{-1}\nabla\Pi$ is gradient descent
        on $\Pi$, which cannot terminate at a maximum or saddle.
\end{enumerate}
%
The energy comes from the same three-point quadrature backend of
Cap.~\ref{ch:driver-integration} (the LM solver in the driver is now
instantiated with \texttt{NLAnalysisEnergyBackend}, arithmetically
identical to the plain adapter when the gate is off). The unit test
\texttt{regularized\_newton\_energy\_linesearch} isolates the mechanism
on $\Pi=(u^2-1)^2$, whose residual root $u=0$ is a \emph{maximum} of
$\Pi$. Plain LM from $u_0=0.15$ converges to that maximum and reports
success; with the gate on, the solver refuses it (every energy-raising
step is rejected), descent inside a well is unimpeded, the cubic
benchmark of the deflation test still lands on its energy-downhill
root, and with the gate off the energy-capable backend reproduces the
plain backend bit-exactly.

\paragraph{Campaign.} With only the line search on
(\path{data/output/p1_linesearch}), the dominant spike of the baseline
collapses, step~117 falls from $0.0477$ to $0.0107$\,MN, \emph{below}
the pre-reversal envelope, step~116 from $0.0281$ to $0.0194$, and the
$-200$\,mm reversal calms from $0.0369/0.0393/0.0241$ to
$0.0260/0.0107/0.0017$\,MN. Step 119 converges where the baseline did
not. Globally \texttt{noconv} drops 80~$\to$~78 and the peak from
$0.0508$ to $0.0493$\,MN. The cost is a handful of extra residual
evaluations per Newton iteration, only while the gate is on.

\section{LM polish after the hybrid TAO step}
\label{sec:tao-polish}

Phase-I left TAO selecting the right basin without closing the
residual under heavy cracking. The polish stage
(\texttt{KOBATHE\_TAO\_POLISH=1}) restores the pre-step state after the
TAO solve, seeds the LM continuation with the TAO iterate through
\texttt{set\_solution}, and lets the local monotone-residual iteration
finish the job through the single surviving commit. The monotone
acceptance cannot climb back over the $\lVert\mathbf R\rVert$ ridge
into the spurious basin, and a residual guard reproduces the plain TAO
step when the polish does not improve.

\paragraph{Campaign.} (\path{data/output/p2_tao_polish}) The polish
engaged on 18 of the 24 hybrid steps and did its local job, the
$+150$\,mm-reversal residuals improve about threefold
($0.137/0.126 \to 0.044/0.041$), three polished steps land near or
below $10^{-4}$ and one closes fully (7 hybrid steps converged against
6 without polish). The global outcome is mixed, \texttt{noconv} rises
to 83 because the slightly different committed states shift the
trajectory of the heavily damaged tail. The polish is therefore an
honest refinement of the hybrid step, not a global win on this mesh.

\section{Combined gates}
\label{sec:combo}

The full stack (\texttt{KOBATHE\_TAO\_REVERSAL} +
\texttt{KOBATHE\_TAO\_POLISH} + \texttt{KOBATHE\_LM\_LINESEARCH},
\path{data/output/p2_tao_ls_combo}) ties the best convergence count
(\texttt{noconv}=77) and keeps every reversal spike tamed (step 117 at
$0.0202$\,MN, well under the envelope bar). Its global $|V|$ maximum,
$0.0554$\,MN, sits at steps 162--163, the $+187.5/+200$\,mm
\emph{loading} extremes of the last cycle, not at a reversal. A
higher force at the pushover extreme reflects a different, less-damaged
committed trajectory carrying more load, which is not the spurious-basin
signature the spike detector targets; the reversal-window metric is the
one the extension is accountable for, and there the combination holds.

\section{The FE\textsuperscript{2} probe}
\label{sec:fe2-probe}

The modelling-level question was whether passing material information
between models (one-way or two-way) smooths the cyclic response. The
repository already had the production FE\textsuperscript{2} engine
(\texttt{MultiscaleAnalysis} + \texttt{BeamMacroBridge} +
\texttt{NonlinearSubModelEvolver}, previously wired only to building
drivers), and the probe added the missing artifact, a single-column
driver (\path{main_rc_column_multiscale.cpp}) with the reference-column
geometry, a fiber Timoshenko macro (eight 2-node elements with reduced
integration, after a shear-locking autopsy against the validated
cantilever benchmark) and one KoBathe RVE at the base hinge, driven by
the exact 196-target protocol of the monolithic driver. A minimal
additive toggle in the evolver selects the homogenisation operator
(boundary reactions, the production path, versus the volume-average
comparison path).

\paragraph{One-way.} The macro trajectory is identical to the
uncoupled fiber column by construction, and it is trivially smooth
(196/196 steps, zero non-converged rows, peak $|V| = 0.0237$\,MN). The
smoothness is bought by never letting the macro see the continuum
pathology, and the price is constitutive fidelity, the fiber section
carries half the monolithic peak shear.

\paragraph{Two-way, the pathological oscillation and its autopsy.}
The first full two-way runs oscillated violently ($|V|$ flipping sign
step to step, up to $\pm 0.06$\,MN near zero drift) while reporting
\texttt{macro\_converged=1} on every row. The audit traced a
self-reinforcing defect in the hybrid observation window of
\texttt{MultiscaleAnalysis} (the recovery path that advances with the
last converged feedback when the staggered loop does not meet its 3\%
tolerance). That path committed \emph{only the macro}, it never
re-solved nor committed the local models, and the injected feedback
refreshed \emph{only} on strictly converged steps. In a run whose only
strict steps were the four axial-preload steps, the RVE stayed frozen
at a near-virgin state while the macro cycled to $\pm 50$\,mm, every
step imposed the full accumulated section kinematics on that stale
state in one jump, the re-cracking response came out chaotic or
diverged outright, the staggered force residual stayed pinned at
$0.4$--$1.0$, and every step fell back to the hybrid window with an
affine anchor ($\mathbf f_0$, $\boldsymbol\varepsilon_{\mathrm{ref}}$)
extrapolating from the preload state. The fix is the opt-in
\texttt{evolve\_locals\_in\_hybrid} of
\texttt{TwoWayFailureRecoveryPolicy}: after the hybrid macro commit the
locals are re-solved under the final macro state with the same
machinery as the one-way path, their material history is committed,
and the feedback anchor is refreshed from the committed state. The
effect is night and day, the staggered residual drops to
$0.02$--$0.07$, strict two-way steps reappear throughout the protocol,
and the committed shear unloads smoothly through the reversals that
previously oscillated.

\paragraph{The homogenisation operator.} Three independent
observations mark the operator as the real lever. With the frozen
(defective) feedback, the volume-average operator converged strictly
where boundary reactions could not and survived eleven steps further.
With the fix, the roles invert, the boundary-reaction operator (whose
condensed tangent is \emph{consistent} with its forces) runs stably,
while the volume-average toggle, which replaces only the force vector
and keeps the boundary-reaction tangent
(\texttt{forces\_consistent\_with\_tangent=false}), aborts at step~40:
once the feedback is live, the inconsistent affine pair
$(\mathbf f_{\mathrm{vol}}, \mathbf D_{\mathrm{bnd}})$ of a cracked RVE
makes the macro unsolvable at the reversal. A consistent volume-average
operator needs its own tangent (finite differences of the averaged
forces), noted as the follow-up.

\paragraph{Two-way with the fix, full protocol.}
With \texttt{evolve\_locals\_in\_hybrid} the boundary-reaction two-way
completes all 196 targets with \emph{zero} non-converged rows, 20
strictly converged steps and the rest in the (now well-fed) hybrid
window with staggered force residuals of $0.02$--$0.07$. One noisy
transient remains during the initial cracking of the RVE (steps
$\sim$5--19, peak $|V|=0.064$\,MN) while the first committed local
histories catch up with the macro; from the $\pm 50$\,mm cycle onward
the loops are smooth, the critical reversals unload monotonically
($0.027 \to 0.013 \to 0.007$\,MN at the $+150$\,mm reversal that
carries the monolithic's dominant spike), and the large-amplitude
cycles track the monolithic loop shape, pinching included, far more
faithfully than the fiber macro. Deferring the coupling start past the
initial-cracking window, or ramping the staggered relaxation during
it, is the obvious next refinement.

\section{Updated evidence matrix}
\label{sec:round2-matrix}

Both remediation levels, on the same $\pm$50/100/150/200\,mm protocol.
The monolithic rows are directly comparable to
Cap.~\ref{ch:driver-integration}; the FE\textsuperscript{2} rows are a
\emph{different model} of the same column and are compared on
smoothness and completion, not on force values.

\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
& \texttt{noconv} & spike (step 117) & note \\
\midrule
\multicolumn{4}{@{}l}{\emph{monolithic continuum (solver remedies)}} \\
baseline G0             & 80 & 0.0477 & peak 0.0508\,MN \\
LM + energy line search & 78 & 0.0107 & near-zero cost \\
hybrid TAO              & 77 & 0.0189 & peak 0.0377\,MN \\
TAO + polish + LS       & 77 & 0.0202 & higher pushover extreme \\
\midrule
\multicolumn{4}{@{}l}{\emph{FE\textsuperscript{2} column (modelling level)}} \\
fiber macro / one-way   & 0/196 & --- & smooth, half the strength \\
two-way boundary + fix  & 0/196 & smooth reversals & one onset transient \\
two-way volume + fix    & aborts (40) & --- & inconsistent $(\mathbf f,\mathbf D)$ pair \\
\bottomrule
\end{tabular}
\end{center}

The reading across both levels is consistent with the Phase-I
conclusion, selection needs \emph{descent} or \emph{consistent
homogenised feedback}, not avoidance. At the solver level the energy
line search is the cheapest sustained win and composes with TAO; at the
modelling level the two-way FE\textsuperscript{2} with evolving local
histories completes the protocol with zero non-converged steps and
smooth large-amplitude loops, at the cost of one RVE solve chain per
step. Everything remains environment- or flag-gated, and the gates-off
monolithic path stays byte-identical to \path{data/output/g0_baseline}.
