% ==========================================================================
%  Chapter 105 -- Driver Integration of the Branch-Selection Extension (Phase I)
% ==========================================================================

\chapter{Driver Integration of the Branch-Selection Extension}
\label{ch:driver-integration}

\paragraph{Normative status.}
Phase~I of the branch-selection extension
(Caps.~\ref{ch:branch-selection-reversals}--\ref{ch:extension-architecture}).
The three remedies --- deflation
(Cap.~\ref{ch:deflated-continuation}), energy descent by TAO
(Cap.~\ref{ch:energy-tao-continuation}) and the Cultural-Algorithm tuner
(Cap.~\ref{ch:ca-tuner-solver}) --- are wired into the validated cyclic
driver \path{src/validation/ReducedRCColumnContinuumBaseline.cpp},
following the control flow fixed in Cap.~\ref{ch:extension-architecture}
(Fig.~\ref{fig:ext-retry-flow}). Every remedy is environment-gated and
\textbf{OFF by default}. The non-regression contract is byte-level
equality of the LM hysteresis CSV against the archived baseline
\path{data/output/g0_baseline} when all gates are off, and it is verified
by a fresh campaign in Section~\ref{sec:drv-evidence}.

\section{Bit-identity by construction}
\label{sec:drv-bitident}

The driver instantiates the continuation solver with the deflation policy
compiled in,
\begin{center}
\ttfamily RegularizedNewtonContinuation<Backend, LevenbergMarquardt,
ShermanMorrisonDeflation>,
\end{center}
even when the gate is off. This is safe for a reason established in
Cap.~\ref{ch:deflated-continuation} and worth restating. Every deflated
branch inside \texttt{advance\_to} is guarded twice, first by
\texttt{if constexpr (Defl::enabled)} and then by the runtime check
\texttt{!roots\_.empty()}. A solver that never receives a registered root
therefore executes \emph{exactly} the floating-point operations of the
\texttt{NoDeflation} instantiation. The gated logic of this chapter only
ever registers roots inside the retry loop, which itself sits behind the
\texttt{KOBATHE\_LM\_DEFLATE} gate. The same discipline applies to the
CSV writer. The historical nine columns are streamed by the same code
path and the new columns (\texttt{defl\_roots}, \texttt{defl\_retries}
for deflation and \texttt{engine}, \texttt{energy} for TAO) are appended
\emph{at the end of the row and only while their gate is on}, so a
gates-off run reproduces the baseline file byte by byte. The previously
hardcoded acceptance floor ($10^{-5}$) is now the environment knob
\texttt{KOBATHE\_LM\_ACCEPT\_FLOOR} with the same default, which
preserves the constant bit-exactly when unset and exposes it as the
tuner's sixth gene.

\section{The shear envelope as a spike detector}
\label{sec:drv-envelope}

The driver observes reversals through the sign of the prescribed drift
increment (the \texttt{since\_rev} counter of the original loop) and it
now maintains three scalars alongside.
%
\begin{description}
  \item[\(V_{\mathrm{exc}}\)] the running maximum of $|V|$ over the
        converged steps of the excursion in progress.
  \item[\(V_{\mathrm{env}}\)] the value of \(V_{\mathrm{exc}}\) frozen at
        the step where a reversal is detected. It is the force scale the
        column actually reached before unloading began.
  \item[\(V_{\mathrm{all}}\)] the historical maximum of $|V|$, used as
        the reference of the tuner's spike penalty.
\end{description}
%
A physical unloading branch keeps $|V|$ \emph{below} the pre-reversal
envelope. The spurious equilibria documented in
Cap.~\ref{ch:branch-selection-reversals} exceed it by factors of 3--5.
The detector therefore flags a step as spurious when it converged
\emph{and} $|V| > s\,V_{\mathrm{env}}$, with
$s=\texttt{KOBATHE\_LM\_DEFLATE\_SPIKE}=1.5$ by default. Requiring
convergence is deliberate. A non-converged step already reports its
failure in the \texttt{conv} column, and deflating it would conflate two
different pathologies (non-convergence and wrong-basin convergence) under
one remedy.

An offline sweep of the archived baseline shows where each pathology
lives on the $2{\times}2{\times}4$ mesh. Every reversal-window step with
$|V|/V_{\mathrm{env}} > 1.5$ (up to $2.5$ at the $\pm 200$\,mm reversal)
is a \emph{non-converged} step, and every converged window step sits at
or below the envelope ($0.6$--$0.96$), which is physical unloading. The
default detector consequently never fires on this baseline, a
conservative property worth having. The opt-in knob
\texttt{KOBATHE\_LM\_DEFLATE\_ON\_NOCONV} extends the detector to
non-converged spikes, where the multi-equilibrium manifests here as a
stagnated high-force acceptance, so the retry mechanism can engage the
real pathology of this mesh.

\section{Detect--restore--retry in the real driver}
\label{sec:drv-retry}

The retry loop follows Fig.~\ref{fig:ext-retry-flow} exactly, and the
driver adds the piece the solver alone cannot provide. The backend commit
is unconditional (\texttt{backend\_.accept()} runs even for a spurious
step), so by the time the spike is measurable the crack ratchet has
already advanced along the wrong branch. The driver therefore brackets
each reversal-window step with a \emph{pair} of checkpoints, the
analysis-side \texttt{nl.capture\_checkpoint()} (which copies the element
container with its full material history, Cap.~\ref{ch:extension-architecture})
and the solver-side \texttt{solver.capture\_state()} (solution plus
secant history). On detection the driver registers the spurious solution
as a deflation root, restores both checkpoints and re-solves the same
window, now repelled from the registered root. At most
\texttt{KOBATHE\_LM\_DEFLATE\_MAXRETRY} (default 2) retries run per step,
and the root list is cleared when the loop moves to the next target,
because the registered equilibria belong to the multi-equilibrium
manifold at that particular load level.

One integration detail matters for the auxiliary outputs. The commit path
fires the driver's step callback, which appends rows to the seven record
vectors behind \path{hysteresis.csv}, \path{control_state.csv} and their
siblings. A retry would therefore leave the rows of the discarded attempt
in those files. The checkpoint consequently also captures the seven
record-vector sizes and the restore truncates them back, so the auxiliary
CSVs only ever contain the accepted attempt of each step.

\section{The hybrid TAO step}
\label{sec:drv-tao}

Under \texttt{KOBATHE\_TAO\_REVERSAL} (reversal windows only) or
\texttt{KOBATHE\_TAO} (every step, diagnostic) the step becomes the
hybrid of Cap.~\ref{ch:energy-tao-continuation}. The LM pass proposes an
iterate, the driver snapshots it, restores the pre-step state (the LM
commit would otherwise contaminate the frozen material state that makes
the incremental energy honest), hands the iterate to
\texttt{TaoEnergyContinuation::set\_initial\_guess} and lets the
trust-region solve consolidate \emph{its} minimiser through the single
surviving commit. The accepted solution is copied back with
\texttt{solver.set\_solution()}, which also resets the LM secant history.
The energy backend added in this phase,
\texttt{NLAnalysisEnergyBackend}, evaluates
\[
  \Pi(\mathbf u)-\Pi(\mathbf u_{\mathrm{ref}})
  = \int_0^1 \mathbf R\bigl(\mathbf u_{\mathrm{ref}} + s\,\Delta\mathbf u\bigr)^{\!\top}
    \Delta\mathbf u \, ds ,
  \qquad \Delta\mathbf u = \mathbf u - \mathbf u_{\mathrm{ref}},
\]
with a three-point Gauss--Legendre rule over the existing residual seam,
so no material, element or model code changes. The value logged in the
\texttt{energy} column is computed \emph{before} the commit, while the
state is still frozen, and is therefore comparable across candidate
branches of the same step. Two guards from the risk analysis are
enforced. The gate is ignored with a warning unless the concrete profile
is \texttt{production\_stabilized}, since the hard crack closure of the
reference profile breaks the $C^1$ smoothness the trust-region model
assumes. And \texttt{KOBATHE\_TAO\_SYMCHECK} measures
$\lVert\mathbf K-\mathbf K^{\top}\rVert_F/\lVert\mathbf K\rVert_F$ once at
startup, falling back to the gradient-only \texttt{bqnls} type when the
penalty coupling asymmetrises the Hessian beyond $10^{-8}$ and the user
did not pin \texttt{KOBATHE\_TAO\_TYPE} explicitly
\cite{BalayPETScManual}.

\section{Window replay for the CA tuner}
\label{sec:drv-replay}

\texttt{KOBATHE\_CA=1} with
\texttt{KOBATHE\_LM\_REPLAY\_WINDOW="$k_0$:$m$"} switches the driver into
replay mode (Fig.~\ref{fig:drv-replay-flow}). The protocol runs normally
up to target $k_0-1$, a checkpoint captures the analysis, the solver, the
loop bookkeeping (direction, reversal counter, envelopes) and the record
sizes, and the Cultural Algorithm of
Cap.~\ref{ch:cultural-algorithm-module} maximises the window fitness
%
\[
  f(\mathbf g) = w_1\,\frac{n_{\mathrm{conv}}}{m}
   - w_2\,\max\!\Bigl(0,\ \frac{|V|_{\mathrm{peak}}}{V_{\mathrm{all}}}-1\Bigr)
   - w_3\,\overline{\imath},
  \qquad (w_1,w_2,w_3) = (1,\,1,\,0.2),
\]
%
where $\overline{\imath}$ is the mean iteration count normalised by
\texttt{max\_newton} and only the \emph{original} protocol targets are
scored, so a genome that subdivides reversals gains nothing from its
intermediate sub-steps. Each candidate evaluation decodes the genome into
the solver's mutators (\texttt{set\_config}, \texttt{set\_regularization}
from Phase~P1), restores the checkpoint, replays targets
$k_0,\dots,k_0{+}m{-}1$ and restores again. The eight genes and their
boxes are
%
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
gene & meaning & box & decoding \\
\midrule
$g_0$ & $\log_{10}\mu_0$              & $[-4,\ 0]$    & $\mu_0 = 10^{g_0}$ \\
$g_1$ & LM grow factor                & $[1.5,\ 10]$  & direct \\
$g_2$ & LM drop factor                & $[0.05,\ 0.9]$& direct \\
$g_3$ & stagnation cutoff             & $[4,\ 30]$    & rounded \\
$g_4$ & secant predictor max scale    & $[1,\ 8]$     & direct \\
$g_5$ & $\log_{10}$ acceptance floor  & $[-7,\ -3]$   & $10^{g_5}$ \\
$g_6$ & reversal subdivision          & $[1,\ 6]$     & rounded \\
$g_7$ & subdivision window            & $[1,\ 5]$     & rounded \\
\bottomrule
\end{tabular}
\end{center}
%
The latched material genes (\texttt{KOBATHE\_CLOSURE\_STIFF\_FRAC},
\texttt{KOBATHE\_ANCHOR\_CAP\_RATIO}) are read once into
\texttt{static const} storage and stay outside the genome. The campaign
script \path{scripts/kobathe_ca_tuner.sh} sweeps them by process instead
(phase~A), runs one seeded CA per value (phase~B) and launches a full
protocol with the winning genome (phase~C). The replay persists
\path{ca_history.csv} (generation, best, population mean, best genome)
and \path{ca_best_genome.json}, whose keys are literally the environment
variables that reproduce the winner, prints the same line to stderr and
stops without continuing the protocol. Determinism per seed is inherited
from the module's single-RNG design \cite{Reynolds1994Cultural}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    stage/.style={draw, rounded corners=2pt, fill=green!8,
                  font=\scriptsize\ttfamily, align=center, inner sep=4pt},
    art/.style={draw, rounded corners=2pt, fill=orange!10,
                font=\scriptsize\ttfamily, align=center, inner sep=4pt},
    flow/.style={-{Stealth[length=5pt]}, thick},
    back/.style={-{Stealth[length=4pt]}, dashed, gray}
  ]
  \node[stage] (prefix) at (0,0)
    {protocol targets $0 \dots k_0{-}1$\\(plain LM continuation)};
  \node[stage] (chk) at (5.2,0)
    {checkpoint\\nl + solver + loop vars\\+ record sizes};
  \node[stage] (decode) at (10.2,0)
    {decode genome\\set\_config / set\_regularization};
  \node[stage] (window) at (10.2,-2.1)
    {replay targets\\$k_0 \dots k_0{+}m{-}1$};
  \node[stage] (fit) at (5.2,-2.1)
    {fitness\\$w_1 \tfrac{n_{conv}}{m} - w_2\,\text{spike} - w_3\,\bar\imath$};
  \node[art] (persist) at (0,-2.1)
    {ca\_history.csv\\ca\_best\_genome.json\\stop (no continuation)};

  \draw[flow] (prefix) -- (chk);
  \draw[flow] (chk) -- (decode);
  \draw[flow] (decode) -- (window);
  \draw[flow] (window) -- (fit);
  \draw[back] (fit) -- node[above,font=\tiny\rmfamily]
    {restore, next candidate} (chk);
  \draw[flow] (fit) -- node[above,font=\tiny\rmfamily]
    {CA done} (persist);

  \draw[dashed, gray!70, rounded corners=3pt]
    ($(fit.west |- decode.north)+(-0.3,0.3)$)
      rectangle ($(decode.east |- window.south)+(0.3,-0.45)$);
  \node[font=\tiny\rmfamily, gray, anchor=north west]
    at ($(fit.west |- window.south)+(-0.25,-0.55)$)
    {CA objective, one call per candidate};
\end{tikzpicture}
\caption[Replay flow of the CA tuner in the driver.]{Replay flow of the
CA tuner inside the driver. The checkpoint captures the analysis (element
container with material history), the continuation solver (solution and
secant history), the loop bookkeeping and the sizes of the record
vectors, so every candidate evaluation starts from the identical state
and the auxiliary CSVs stay clean.}
\label{fig:drv-replay-flow}
\end{figure}

\section{Gate reference}
\label{sec:drv-gates}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
variable & default & effect \\
\midrule
\texttt{KOBATHE\_LM\_ACCEPT\_FLOOR}      & $10^{-5}$ & exposes the acceptance floor \\
\texttt{KOBATHE\_LM\_DEFLATE}            & off       & detect--restore--retry loop \\
\texttt{KOBATHE\_LM\_DEFLATE\_SPIKE}     & 1.5       & spike threshold over $V_{\mathrm{env}}$ \\
\texttt{KOBATHE\_LM\_DEFLATE\_MAXRETRY}  & 2         & retries per step \\
\texttt{KOBATHE\_LM\_DEFLATE\_ON\_NOCONV} & off      & also retry non-converged spikes \\
\texttt{KOBATHE\_LM\_DEFLATE\_RESID\_FACTOR} & 10    & residual guard of the kept deflated attempt \\
\texttt{KOBATHE\_LM\_DEFLATE\_POWER}     & 2         & deflation exponent $p$ \\
\texttt{KOBATHE\_LM\_DEFLATE\_SHIFT}     & 1         & deflation shift $\alpha$ \\
\texttt{KOBATHE\_LM\_DEFLATE\_TAUMAX}    & 50        & clamp of the deflation factor \\
\texttt{KOBATHE\_TAO\_REVERSAL}          & off       & hybrid TAO step in reversal windows \\
\texttt{KOBATHE\_TAO}                    & off       & hybrid TAO step everywhere \\
\texttt{KOBATHE\_TAO\_TYPE}              & bntr      & TAO algorithm \\
\texttt{KOBATHE\_TAO\_MAXIT}             & 200       & TAO iteration cap \\
\texttt{KOBATHE\_TAO\_GATOL}, \texttt{\_GRTOL} & $10^{-8}$ & gradient tolerances \\
\texttt{KOBATHE\_TAO\_ACCEPT\_FLOOR}     & $10^{-6}$ & residual acceptance for TAO \\
\texttt{KOBATHE\_TAO\_SYMCHECK}          & off       & one-shot Hessian symmetry probe \\
\texttt{KOBATHE\_CA}                     & off       & replay mode \\
\texttt{KOBATHE\_LM\_REPLAY\_WINDOW}     & ---       & window $k_0{:}m$ \\
\texttt{KOBATHE\_CA\_POP}, \texttt{\_GEN} & 8, 10    & population, generations \\
\texttt{KOBATHE\_CA\_SEED}               & 20260715  & RNG seed (reproducibility) \\
\texttt{KOBATHE\_CA\_TOPFRAC}            & 0.25      & belief-space acceptance fraction \\
\bottomrule
\end{tabular}
\end{center}

\section{Campaign evidence}
\label{sec:drv-evidence}

Evidence from the Phase-I campaigns on the $2{\times}2{\times}4$ mesh,
protocol $\pm$50/100/150/200\,mm, SPS~8, 196 targets, release build.

\paragraph{Gate G1, gates off.}
A fresh campaign with every gate off
(\path{data/output/g1_gateoff}) reproduces the archived baseline
\path{data/output/g0_baseline} \emph{byte by byte} in all eight CSV
outputs (LM hysteresis, quasi-static hysteresis, control state, crack
state, embedding gap, both rebar histories and the host probes). The
non-regression contract of Section~\ref{sec:drv-bitident} holds on the
real campaign, not only by construction.

\paragraph{Gate G1, deflation on (base detector).}
With \texttt{KOBATHE\_LM\_DEFLATE=1} and the default detector
(\path{data/output/p1_deflation}) the run completes the full protocol
with \texttt{noconv}=80, and the nine historical columns are identical
to the baseline. Zero retries fire, consistent with the offline sweep of
Section~\ref{sec:drv-envelope}: on this mesh no reversal-window step
both converges and spikes. The detector is conservative in the exact
sense intended, it never touches a step the plain continuation handles.

\paragraph{Gate G1, deflation on non-converged spikes.}
With the opt-in extension
(\texttt{KOBATHE\_LM\_DEFLATE\_ON\_NOCONV=1},
\path{data/output/p1_deflation_noconv}) the retry mechanism engages
exactly where the offline sweep predicted, four retries at steps 117 and
118 of the $+150$\,mm reversal, with roots registered, checkpoints
restored and the new CSV columns populated. The \emph{branch-selection}
outcome is negative on this mesh, and it is the most valuable
measurement of the campaign. Deflating the stagnated high-force root
sent the solver to \emph{neighbouring roots of even higher force}
($|V|=0.43$ and $0.28$\,MN during the retries of step~117 and 118,
against a $0.025$\,MN envelope), and committing the last attempt
degraded the rest of the run (\texttt{noconv} 80~$\to$~111, peak $|V|$
$0.048\to0.556$\,MN). The multi-equilibrium manifold near this reversal
is dense in spurious basins, and repulsion alone does not know
\emph{which} of the remaining basins is physical --- precisely the
argument for the energy criterion of Cap.~\ref{ch:energy-tao-continuation}.
Two consequences were adopted. The retry loop now carries a safety
fallback, on exhaustion with a result worse than the base attempt it
restores and reproduces the base attempt, so the experiment is
never-worse-than-baseline by construction. And deflation stays an
environment-gated experiment, not a promoted remedy
(Section~\ref{sec:drv-promotion}).

\paragraph{Gate G1, deflation with safety fallback.}
The definitive deflation campaign
(\path{data/output/p1_deflation_final}, residual guard active) exercises
seven retries across four steps (117, 180, 181, 182). Every deflated
re-solve landed either in a basin of \emph{higher} force (up to
$|V|=14.7$\,MN at step 182, against a $0.016$\,MN envelope) or in a
low-force junk state with an exploded residual ($|V|=0.0005$\,MN with
$\lVert\mathbf R\rVert=1.5\times10^{2}$ at step 180, caught by the
residual guard). The fallback restored the base attempt at all four
steps, and the run's nine historical columns are byte-identical to the
baseline with \texttt{noconv}=80. Gate G1 closes as follows. The
non-regression clause holds (gates off, byte-identical). The mechanism
clause holds (retries exercised in the real driver, roots registered,
checkpoint/restore and record truncation correct). The branch-selection
clause \emph{fails on this mesh}, and fails informatively: near these
reversals the equilibrium manifold is dense in spurious basins, and
repulsion from one root says nothing about the quality of the next one.
Branch selection needs a criterion of \emph{descent}, not of
\emph{avoidance}, which is what the energy hybrid provides.

\paragraph{Gate G2, hybrid TAO.}
With \texttt{KOBATHE\_TAO\_REVERSAL=1} (\path{data/output/p2_tao}) the
symmetry probe reports
$\lVert\mathbf K-\mathbf K^{\top}\rVert_F/\lVert\mathbf K\rVert_F =
1.2\times10^{-17}$, so the penalty coupling leaves the Hessian
numerically symmetric and BNTR keeps its Newton model. The campaign runs
24 hybrid steps (three per reversal window, eight reversals) and is the
strongest result of the three remedies.
%
\begin{itemize}
  \item Every TAO step reports \emph{negative} incremental energy, the
        trust region always descends from the LM proposal, which is the
        branch-selection semantics G2 asks for
        ($\Pi(\mathbf u_{\mathrm{tao}}) < \Pi(\mathbf u_{\mathrm{LM}})$
        at the same load, same reference).
  \item The dominant spike of the baseline (step 117, $+150$\,mm
        reversal, $|V|=0.0477$\,MN) drops to $0.0189$\,MN, below the
        $1.5\,V_{\mathrm{env}} = 0.037$\,MN bar, and the run's global
        peak falls from $0.0508$ to $0.0377$\,MN.
  \item \texttt{noconv} improves from 80 to 77. The healthier state the
        hybrid leaves behind lets the plain LM converge at step 119,
        which the baseline could not.
  \item The strict floor $\lVert\mathbf R\rVert \le 10^{-6}$ holds on
        6 of 24 TAO steps, all of them at the $\pm50$\,mm reversals
        (residuals $10^{-9}$). Once heavy cracking accumulates the
        incremental potential loses smoothness, BNTR saturates its
        200-iteration budget and stalls at
        $\lVert\mathbf R\rVert \sim 10^{-1}$, exactly the nonsmoothness
        risk anticipated in Cap.~\ref{ch:energy-tao-continuation}.
\end{itemize}
%
G2 is therefore met in its energy and envelope clauses and only
partially in its residual clause, concentrated in the heavily damaged
half of the protocol.

\paragraph{Gate G4, CA tuner.}
The campaign script ran phase~B on the $\pm150$\,mm reversal window
($k_0{=}116$, $m{=}6$, population~8, ten generations, seed fixed,
\path{data/output/p3_ca}). The window was chosen over the literal
``first reversal'' of the gate because the first reversal has no
regularized steps in the baseline, so the gate would be vacuous there;
the $+150$\,mm window carries four of the six baseline failures and the
dominant spike. Results.
%
\begin{itemize}
  \item The best-fitness history is non-decreasing
        ($0.256 \to 0.367 \to 0.373$ across ten generations, elitism),
        and the population mean rises from $-5.7$ to $+0.02$.
  \item The winning genome scores 3/6 converged in the window (baseline
        2/6) with an in-window peak of $0.0217$\,MN, \emph{below} the
        envelope itself, so the spike clause of G4
        ($|V|_{\mathrm{peak}} \le 1.5\,V_{\mathrm{env}} = 0.039$) holds
        with margin and the step-117 spike is suppressed entirely.
  \item Phase~C, the full 196-target protocol under the winning genome,
        drops \texttt{noconv} from 80 to 66 ($-17.5\%$) and the global
        peak from $0.0508$ to $0.0481$\,MN, at zero per-step overhead
        (the genome only re-parameterises the plain LM continuation).
  \item The winner is structurally interesting,
        $\mu_0 = 1.36\times10^{-2}$, grow $5.53$, drop $0.79$,
        stagnation cutoff 14, predictor scale $6.25$, floor
        $3.2\times10^{-5}$, and \emph{reversal subdivision 3 with window
        3}. Subdivision alone was known to worsen the reversal (it
        converges more firmly to the spurious branch); combined with a
        slow $\mu$ relaxation (drop $0.79$ against the default $0.3$)
        it helps, precisely the non-obvious coupling a tuner exists to
        find.
  \item Reproducibility is carried by construction (every stochastic
        draw flows from \texttt{KOBATHE\_CA\_SEED}, bit-identical reruns
        verified at module level by ctest
        \texttt{algorithms\_cultural\_algorithm}) and the artifacts
        record seed, window and genome.
\end{itemize}

\section{Promotion decision (Gate G5)}
\label{sec:drv-promotion}

The evidence matrix of the three remedies on the
$2{\times}2{\times}4$ campaign.

\begin{center}
\begin{tabular}{@{}lllll@{}}
\toprule
remedy & selects low branch & \texttt{noconv} & peak $|V|$ [MN] & overhead \\
\midrule
baseline G0            & ---            & 80 & 0.0508 & --- \\
deflation (retry)      & no (all basins worse) & 80 & 0.0508 & checkpoint per window step \\
hybrid TAO (reversals) & yes (energy descent)  & 77 & 0.0377 & $\le 200$ TAO its per window step \\
CA-tuned LM (full run) & indirectly            & 66 & 0.0481 & none at run time \\
\bottomrule
\end{tabular}
\end{center}

No remedy meets \emph{every} clause of its gate, so none is promoted to
a default, and all three stay environment-gated reproducible
experiments. Two dispositions follow from the matrix. The CA-tuned
genome is the cheapest sustained improvement, it re-parameterises the
continuation that already runs, and adopting it in a campaign is one
line of environment variables (recorded in
\path{ca_best_genome.json}). The hybrid TAO step is the only remedy
that \emph{selects} the branch rather than avoiding a root, it tames
the dominant spike below the envelope bar, and its open residual clause
is a solver-tuning problem (nonsmooth $\Pi$ under heavy cracking), not
a conceptual one. Deflation closes as a negative result with a complete
mechanism, kept for meshes where the spurious basin structure is
sparser. The workstream and this disposition are registered in
\path{src/validation/ValidationCampaignCatalog.hh}; any statement in
the thesis remains subject to the author's approval.
