% ==========================================================================
%  Chapter 57 — Ko–Bathe Plastic-Fracturing Concrete Model
% ==========================================================================

\chapter{Ko--Bathe Plastic-Fracturing Concrete Model}
\label{ch:ko-bathe-concrete}

This chapter documents the implementation of a two-dimensional plastic-fracturing
concrete model following the formulation proposed by Ko and
Bathe~\cite{KoBathe2026} for finite element analysis.  The model combines
three physical mechanisms---\emph{fracturing}, \emph{plasticity}, and
\emph{cracking}---in a strain-driven framework suitable for plane stress
problems.  The implementation is contained in a single header file,
\texttt{KoBatheConcrete.hh}, and satisfies all concept levels of the
\texttt{fall\_n} constitutive interface.

% ========================================================================
\section{Theoretical Background}
\label{sec:kb-theory}

The model belongs to the family of \emph{plastic-fracturing} constitutive
relations for concrete, which recognise that the nonlinear behaviour of
concrete under multi-axial loading arises from three distinct phenomena~\cite{KoBathe2026,BatheRamaswamy1979}:

\begin{enumerate}[label=(\roman*)]
  \item \textbf{Fracturing:} progressive degradation of the elastic tangent
        moduli as internal micro-cracking develops under compression.
  \item \textbf{Plasticity:} irrecoverable strains that accumulate once the
        failure envelope is reached.
  \item \textbf{Cracking:} discrete tensile fracture planes (smeared) that
        form when the maximum principal stress exceeds the tensile capacity.
\end{enumerate}

These three mechanisms are coupled through the octahedral stress invariants
and operate in sequence within each strain increment: compressive stiffness
degrades via fracturing; if the resulting stress state violates a yield
surface, plasticity activates; if the local tensile stress exceeds the crack
function, a smeared crack is initiated.

% ────────────────────────────────────────────────────────────────────────
\subsection{Octahedral Stress Invariants}
\label{ssec:kb-octahedral}

The model uses stress invariants as driving quantities for both
fracturing and cracking.  For a plane stress state
$\bsigma = (\sigma_{xx}, \sigma_{yy}, \tau_{xy})^T$ with $\sigma_{zz} = 0$,
the principal stresses are
\[
  \sigma_1 = \frac{\sigma_{xx}+\sigma_{yy}}{2}
           + \sqrt{\left(\frac{\sigma_{xx}-\sigma_{yy}}{2}\right)^2 + \tau_{xy}^2},
  \qquad
  \sigma_2 = \frac{\sigma_{xx}+\sigma_{yy}}{2}
           - \sqrt{\left(\frac{\sigma_{xx}-\sigma_{yy}}{2}\right)^2 + \tau_{xy}^2},
\]
with $\sigma_3 = 0$.  The octahedral normal stress and the deviatoric invariant read
\begin{align}
  \sigma_o &= \frac{\sigma_1 + \sigma_2}{3},
  \label{eq:kb-sigma-o}
  \\[4pt]
  \tau_o   &= \frac{\sqrt{2\,J_2}}{3}
             = \frac{1}{3}\,\sqrt{\sigma_{ij}^{\text{dev}}\,\sigma_{ij}^{\text{dev}}},
  \label{eq:kb-tau-o}
\end{align}
where $J_2 = (\sigma_1^2 + \sigma_2^2 - \sigma_1\sigma_2)/3$ is the second
deviatoric stress invariant for plane stress ($\sigma_3 = 0$).
The sign convention is engineering: compression gives $\sigma_o < 0$.

\emph{Note:} The paper's $\tau_o$ (Eq.~2b in~\cite{KoBathe2026}) is
$\sqrt{2J_2}/3$, which differs from the classical octahedral shear stress
$\tau_{\text{oct}} = \sqrt{2J_2/3}$ by a factor of $\sqrt{3}$.

% ────────────────────────────────────────────────────────────────────────
\subsection{Fracturing Mechanism}
\label{ssec:kb-fracturing}

The ascending non-linear branch of the stress--strain curve is captured
entirely by degrading the elastic moduli.  The secant bulk and shear moduli
are defined as \emph{rational} functions (Eqs.~10a--b in~\cite{KoBathe2026}):
\begin{align}
  K_s(\sigma_o) &= \frac{K_e}{1 + A\,(-\sigma_o/f_c')^{\,b-1}},
  \label{eq:kb-ks}
  \\[4pt]
  G_s(\tau_o)   &= \frac{G_e}{1 + c\,(\tau_o/f_c')^{\,d-1}},
  \label{eq:kb-gs}
\end{align}
where $K_e$ and $G_e$ are the initial elastic moduli, $f_c'$ is the
uniaxial compressive strength, and $A, b, c, d$ are empirically fitted
coefficients depending on $f_c'$ (see Section~\ref{ssec:kb-parameters}).
The formula~\eqref{eq:kb-ks} applies for $-\sigma_o/f_c' \le 2$; an
extended branch is defined for higher confinement.

The fracture history is driven by the \emph{maximum} octahedral invariants
ever reached:
\[
  \bar\sigma_o = \min_{t'\le t} \sigma_o(t'),
  \qquad
  \bar\tau_o   = \max_{t'\le t} \tau_o(t').
\]
During unloading $\sigma_o > \bar\sigma_o$, the moduli remain frozen at
their last-degraded values, representing the irreversibility of
micro-cracking.

The tangent moduli follow from differentiation (Eqs.~11a--b):
\begin{align}
  K_c &= \frac{K_e}{1 + A\,b\,(-\bar\sigma_o/f_c')^{\,b-1}},
  \label{eq:kb-kc}
  \\[4pt]
  G_c &= \frac{G_e}{1 + c\,d\,(\bar\tau_o/f_c')^{\,d-1}}.
  \label{eq:kb-gc}
\end{align}

\paragraph{Confining pressure $\sigma_{\text{id}}$.}
A confining pressure correction is introduced (Eq.~10b in the paper):
\begin{equation}
  \sigma_{\text{id}} = \frac{k\,f_c'\,(\tau_o/f_c')^{n}}{1 + l\,(-\sigma_o/f_c')^{m}},
  \label{eq:kb-sigma-id}
\end{equation}
which is subtracted from the volumetric stress contribution:
\begin{equation}
  \sigma_{ij} = 2\,G_s\,e^{\text{dev}}_{ij} + K_s\,e_{kk}\,\delta_{ij} - \sigma_{\text{id}}\,\delta_{ij}.
  \label{eq:kb-stress-with-sigma-id}
\end{equation}

Given $K$ and $G$, the plane stress constitutive matrix is
\[
  \mathbf{C}^{\text{ps}} = \frac{E}{1-\nu^2}
  \begin{bmatrix}
    1 & \nu & 0 \\ \nu & 1 & 0 \\ 0 & 0 & \tfrac{1-\nu}{2}
  \end{bmatrix},
  \qquad
  E = \frac{9KG}{3K+G},
  \quad
  \nu = \frac{3K-2G}{2(3K+G)}.
\]

% ────────────────────────────────────────────────────────────────────────
\subsection{Compression Coordinate System}
\label{ssec:kb-compression-coords}

Plasticity yield surfaces are defined not in principal-stress space but in a
\emph{compression coordinate} system.  Given the principal elastic strains
$(\varepsilon_1 \ge \varepsilon_2)$, define the compressive magnitudes (positive in
compression)
\[
  D_1 = \max(-\varepsilon_1,\, 0), \qquad D_2 = \max(-\varepsilon_2,\, 0),
\]
and the compression coordinates
\begin{align}
  \tilde e_1 &= D_2 - D_1 \ge 0 \qquad \text{(deviatoric compression)},
  \label{eq:kb-e1}
  \\[4pt]
  \tilde e_2 &= D_1 + D_2 \ge 0 \qquad \text{(volumetric compression)}.
  \label{eq:kb-e2}
\end{align}
For the plasticity yield functions, these are mapped to a set of three
elastic coordinates:
\[
  \hat{e}_1 = \frac{|\tilde e_2|}{\sqrt{3}}, \qquad
  \hat{e}_2 = \tilde e_1, \qquad
  \hat{e}_3 = \frac{\tilde e_2}{2},
\]
corresponding to hydrostatic, uniaxial, and biaxial loading paths
respectively.

% ────────────────────────────────────────────────────────────────────────
\subsection{Plasticity Yield Surfaces}
\label{ssec:kb-plasticity}

Three yield functions are defined in the compression coordinate space
(Eq.~15c--d in~\cite{KoBathe2026}):
\begin{align}
  f_1 &= |\hat{e}_1| - f_c'\,f_{p1}(\hat{e}_{p1}) \le 0, \label{eq:kb-f1} \\
  f_2 &= \hat{e}_2   - f_c'\,f_{p2}(\hat{e}_{p2}) \le 0, \label{eq:kb-f2} \\
  f_3 &= \hat{e}_3   - f_c'\,f_{p3}(\hat{e}_{p3}) \le 0, \label{eq:kb-f3}
\end{align}
where $\hat{e}_{pj}$ are effective plastic strains accumulated per surface
(Eq.~15b).  The failure curves (Eq.~15d) are:
\begin{align}
  f_{p1}(\hat{e}_{p1}) &= c_{p1}\left(1 + \frac{\hat{e}_{p1}-a}{a}\right)^{-s}\hat{e}_{p1},
  \label{eq:kb-fp1} \\
  f_{p2}(\hat{e}_{p2}) &= c_{p2}\exp\!\left[-\left(\frac{\hat{e}_{p2}}{c_2}\right)^{m_2}\right]\hat{e}_{p2},
  \label{eq:kb-fp2} \\
  f_{p3}(\hat{e}_{p3}) &= c_{p3}\exp\!\left[-\left(\frac{\hat{e}_{p3}}{c_3}\right)^{m_3}\right]\hat{e}_{p3},
  \label{eq:kb-fp3}
\end{align}
with fixed constants $c_{p1}=3$, $c_{p2}=36$, $c_{p3}=12$ and hydrostatic
parameters $a = 2.5\times 10^{-5}$, $s = 0.8$.

The Appendix~C coefficients $m_2, c_2, m_3, c_3$ are coupling parameters
that depend on the hydrostatic effective plastic strain $\hat{e}_{p1}$ 
(Eqs.~C.1--C.2 in the paper):
\begin{align}
  m_2 &= \frac{1}{\ln\!\left(\sqrt{2/3}\,\alpha_2 / (\beta_2\,\hat{\varepsilon}_{p2})\right)},
    &  c_2 &= (\beta_2\,\hat{\varepsilon}_{p2})\,m_2^{1/m_2}, \\
  \alpha_2 &= 1 + 0.575\,(e_{cp2})^{0.315}, &
  \beta_2  &= 1 + 8.4\,(e_{cp2})^{0.25}, \\
  e_{cp2}  &= \max(0,\; K_e\,\hat{e}_{p1} - 0.45), &
  e_{cp3}  &= \max(0,\; K_e\,\hat{e}_{p1} - 2.17),
\end{align}
with reference plastic strains $\hat{\varepsilon}_{p2} = 0.0012$ and
$\hat{\varepsilon}_{p3} = 0.0025$.

The return mapping uses a Newton iteration per active surface
(Eqs.~27--29):  for each $f_j > 0$, solve
\[
  f_j = \hat{e}_j - \Delta\lambda_j - f_c'\,f_{pj}(\hat{e}_{pj} + \sqrt{2/3}\,\Delta\lambda_j) = 0
\]
for $\Delta\lambda_j \ge 0$ using
\[
  \Delta\lambda_j^{(I+1)} = \Delta\lambda_j^{(I)} - \frac{f_j^{(I)}}{-1 - \sqrt{2/3}\,\partial(f_c' f_{pj})/\partial\hat{e}_{pj}}.
\]

% ────────────────────────────────────────────────────────────────────────
\subsection{Cracking Mechanism}
\label{ssec:kb-cracking}

Tensile cracking is governed by a Menetr\'ey--Willam crack function in the
octahedral stress plane (Eq.~16 in~\cite{KoBathe2026}):
\begin{equation}
  g(\sigma_o, \tau_o) = \tau_o - \frac{0.944\,f_c'}{w(e)}\,
    \left(t_p - \frac{\sigma_o}{f_c'}\right)^{0.724},
  \label{eq:kb-crack-fn}
\end{equation}
where $t_p = f_t / f_c'$ is the tensile-to-compressive strength ratio, and
$w(e)$ is an elliptic function of the eccentricity parameter:
\begin{align}
  w(e) &= \frac{4(1-e^2)\cos^2\theta + (2e-1)^2}
               {2(1-e^2)\cos\theta + (2e-1)\sqrt{4(1-e^2)\cos^2\theta + 5e^2-4e}},
  \label{eq:kb-w-eccentricity} \\
  e    &= 0.670551\,(t_p - \sigma_o/f_c')^{0.133}.
  \label{eq:kb-eccentricity}
\end{align}
A crack initiates when $g > 0$ \emph{and} the maximum principal stress is
tensile.

The model tracks up to two orthogonal smeared cracks in each material point.
The second crack, when formed, is forced to be perpendicular to the first.
Each crack possesses:
\begin{itemize}
  \item a fixed normal direction $\mathbf{n}$,
  \item an opening strain $\varepsilon_{nn}$ (projected normal strain),
  \item a maximum historical opening strain $\bar\varepsilon_{nn}$,
  \item a closing flag (re-contact when $\varepsilon_{nn} < 0$).
\end{itemize}

\paragraph{Cracked tangent (Eq.~21).}
After a crack forms, the tangent matrix is modified in the crack-local
coordinate system with stiffness retention parameters:
\begin{align}
  E_{nn} &= \eta_N\,\lambda_a & &\text{(normal, } \eta_N = 0.0001\text{)}, \label{eq:kb-eta-N} \\
  G_{ns} &= \eta_S\,G_c       & &\text{(shear, }  \eta_S = 0.1\text{)}.    \label{eq:kb-eta-S}
\end{align}

\paragraph{Tension softening (Eqs.~22--23).}
For the softening branch between peak and ultimate strains:
\begin{align}
  \varepsilon_{tp} &= \frac{f_c'\,t_p}{K_e + \tfrac{4}{3}\,G_e},
  \label{eq:kb-eps-tp} \\
  \varepsilon_{tu} &= \frac{2\,G_f}{f_c'\,t_p\,l_b},
  \label{eq:kb-eps-tu} \\
  C_{ts} &= -\frac{f_t}{\varepsilon_{tp} - \varepsilon_{tu}},
  \label{eq:kb-Cts}
\end{align}
where $G_f$ is the fracture energy (N/mm) and $l_b$ is the characteristic
element length (mm).  When $\varepsilon_{tp} \le \varepsilon_{nn} \le
\varepsilon_{tu}$, the normal stiffness $E_{nn}$ is replaced by $C_{ts}$.
When the crack closes ($\varepsilon_{nn}<0$), full compressive normal
stiffness is restored (\emph{re-contact}).

\subsection{Audit of the Current 3D Extension}
\label{ssec:kb-3d-audit}

The repository also contains a three-dimensional constitutive relation,
\texttt{KoBatheConcrete3D}, used by the multiscale continuum sub-models.
That class should \emph{not} be read as a literal transcription of the 2026
paper.  The paper formulates a generally applicable three-dimensional
framework~\cite{KoBathe2026}, but the current code path couples that theory to
additional numerical stabilization choices required by the FE\textsuperscript{2}
micro-problems.

\paragraph{Normative distinction.}
The present codebase therefore contains two scientifically distinct layers:
\begin{enumerate}
  \item the plane-stress implementation
    \texttt{KoBatheConcrete.hh}, which aims to follow the article directly;
  \item the three-dimensional implementation
    \texttt{KoBatheConcrete3D.hh}, which follows the same
    fracture/plasticity/crack sequence but adds FE-oriented stabilization in
    the cracked tangent.
\end{enumerate}

\paragraph{What is faithful to the paper in 3D.}
The current 3D relation still preserves the main ingredients of the article:
\begin{itemize}
  \item the same octahedral stress classification
    \((\sigma_o,\tau_o)\) as the driver of fracture loading/unloading;
  \item the same fracture-energy scaling
    \(\varepsilon_{tu} = 2G_f/(f_t l_b)\);
  \item the same scalar fracture-moduli laws
    \(K_s, G_s, K_c, G_c\);
  \item the same three compression-surface plastic update logic;
  \item the same Men\'etrey--Willam-type crack envelope for tensile cracking.
\end{itemize}

\paragraph{What is intentionally different.}
The cracked tangent in the current 3D implementation is a stabilized variant.
In the paper, Table~2 and Eq.~(21) prescribe the retention factors
\begin{equation}
  \eta_N = 10^{-4},
  \qquad
  \eta_S = 10^{-1},
  \label{eq:kb-paper-retention-3d}
\end{equation}
with a sharp open/close transition for the crack-normal stiffness.  In the
current multiscale production path, the default 3D profile is instead
\begin{equation}
  \eta_N^{\mathrm{prod}} = 0.20,
  \qquad
  \eta_S^{\mathrm{prod}} = 0.50,
  \qquad
  \alpha(\varepsilon_{nn}) =
  \frac{1}{2}\left(1 + \tanh\frac{\varepsilon_{nn}}{\delta_c}\right),
  \label{eq:kb-3d-stabilized-retention}
\end{equation}
with \(\delta_c = 10^{-5}\) by default.  Hence the returned normal stiffness
uses
\begin{equation}
  E_{nn} =
  (1-\alpha)\,\lambda_a + \alpha\,E_{nn}^{\mathrm{open}},
  \qquad
  E_{nn}^{\mathrm{open}} =
  \begin{cases}
    C_{ts}, & \varepsilon_{tp} \le \varepsilon_{nn} \le \varepsilon_{tu},\\
    \eta_N^{\mathrm{prod}}\lambda_a, & \text{otherwise},
  \end{cases}
\end{equation}
and the crack-related shear terms are scaled with
\begin{equation}
  G_{ns}^{\mathrm{eff}}
  =
  \bigl((1-\alpha)+\alpha\,\eta_S^{\mathrm{prod}}\bigr)\,G_{ns}.
\end{equation}

This modification is not claimed to be the unique or exact constitutive
representation of the paper in 3D.  It is a numerical regularization aimed at
reducing tangent discontinuities in the fully coupled FE\textsuperscript{2}
micro-problems.

\paragraph{Length scale.}
The article states in Section~2.1.5 and Fig.~7(b) that when \(l_b\) is not
provided globally, the local crack-band interpretation should use the
\emph{maximum side length} of the element, not the cube-root of its
volume~\cite{KoBathe2026}.  The current sub-model factory now follows that
interpretation:
\begin{equation}
  l_b^{\mathrm{local}}
  =
  10^3 \max(\Delta x, \Delta y, \Delta z)
  \quad \text{(mm)}.
  \label{eq:kb-local-lb-max-side}
\end{equation}

\begin{table}[htbp]
\centering
\caption{Current status of the Ko--Bathe implementations in the repository.}
\label{tab:kb-implementation-audit}
\begin{tabular}{@{}p{0.30\textwidth}p{0.28\textwidth}p{0.28\textwidth}@{}}
\toprule
\textbf{Feature} & \textbf{Paper / 2D-reference path} & \textbf{Current 3D FE\textsuperscript{2} path} \\
\midrule
Crack retention & $\eta_N=10^{-4}$, $\eta_S=10^{-1}$ & $\eta_N=0.20$, $\eta_S=0.50$ by default \\
Crack closure   & sharp re-contact switch & smooth \(\tanh\) blend by default \\
Length scale \(l_b\) & global crack-band width or local max side length & now local max side length when inferred \\
Factory parameter order & \texttt{fc, tp, Gf, lb} & corrected to match that order explicitly \\
Scientific status & article-faithful target & stabilized extension; not a strict paper transcription \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Implementation contract.}
To keep this distinction executable rather than merely descriptive, the code
now exposes explicit crack-stabilization profiles for the 3D relation:
\begin{itemize}
  \item \texttt{KoBathe3DCrackStabilization::paper\_reference()}, which
    reproduces the Table~2 retention factors and disables smooth closure; and
  \item \texttt{KoBathe3DCrackStabilization::stabilized\_default()}, which
    preserves the current production profile used in the multiscale runs.
\end{itemize}
Therefore the scientific debt is no longer hidden: the user can now choose
between a paper-like audit mode and the current FE\textsuperscript{2}
stabilized mode, and the documentation states explicitly that the default 3D
path is a \emph{stabilized implementation variant}.

\paragraph{Ordering of plasticity and crack kinematics.}
Table~1 and Fig.~8 of~\cite{KoBathe2026} place the crack-opening/closure
solution \emph{after} the elastic-fracturing / elasto-plastic update.  The
present audit confirmed that the repository still does \emph{not} implement a
fully article-faithful reordering of the entire 3D constitutive routine.
However, an important inconsistency was corrected in the present pass: when the
plastic return mapping updates the plastic strain inside a step, the stored
crack opening
\(\varepsilon_{nn}^{\mathrm{cr}}\)
and the closure flag are now refreshed from the \emph{final} elastic strain
state,
\begin{equation}
  \varepsilon_{nn}^{\mathrm{cr}}
  =
  \mathbf{n}^{\mathsf T}\,\boldsymbol{\varepsilon}^{e}_{t+\Delta t}\,\mathbf{n},
  \qquad
  \boldsymbol{\varepsilon}^{e}_{t+\Delta t}
  =
  \boldsymbol{\varepsilon}_{t+\Delta t}
  -
  \boldsymbol{\varepsilon}^{p}_{t+\Delta t},
\end{equation}
rather than from a transient pre-return trial strain.  This closes a concrete
implementation bug in the 3D extension: the committed crack opening is now
consistent with the committed plastic strain.  The broader algorithmic debt
remains explicit: the nucleation logic is still not a strict line-by-line
transcription of the article flow chart, and the documentation therefore does
not claim full 3D article-faithful equivalence.

\paragraph{Consistent tangent status.}
The article explicitly notes that the consistent stress--strain matrix is
important for quadratic convergence of the nonlinear solve
(\emph{cf.} the discussion around Eqs.~(30)--(31) and the implementation
remarks in Section~2.2 of~\cite{KoBathe2026}).  The repository now exposes that
choice explicitly at the 3D factory level, but the audit also showed that the
numerical consistent tangent should still be treated as \emph{experimental} in
the FE\textsuperscript{2} sub-models.  In the audited Case~5 first-cracked
benchmark, the default fracture-sec tangent converged, whereas the consistent
tangent variant reduced runtime but failed with two micro-solves.  Hence the
production FE\textsuperscript{2} profile continues to use the fracture-sec
tangent by default, while the consistent tangent is retained as an explicit
audit/experimental mode rather than being silently enabled.

% ────────────────────────────────────────────────────────────────────────
\subsection{Material Parameters}
\label{ssec:kb-parameters}

All coefficients are derived from a single primary parameter: the uniaxial
compressive strength $f_c'$ (in~MPa, positive).

\paragraph{Elastic moduli (Eq.~A.1).}
\begin{equation}
  K_e = 11000 + 3.2\,f_c' \;\text{(MPa)},
  \qquad
  G_e = 9224 + 136\,f_c' + 3296\!\times\!10^{-15}\,(f_c')^{8.273} \;\text{(MPa)}.
  \label{eq:kb-elastic-params}
\end{equation}

\paragraph{Fracture degradation (Eqs.~A.2--A.4).}
For $f_c' \le 31.7$~MPa:
\begin{equation}
  A = 0.516,\quad
  c = 3.573,\quad
  d = 2.12 + 0.0183\,f_c',\quad
  m = -2.415,\quad
  n = 1.0.
  \label{eq:kb-fracture-coeffs-low}
\end{equation}
For all $f_c'$:
\begin{equation}
  b = 2 + 1.81\!\times\!10^{-8}\,(f_c')^{4.461},\quad
  k = \frac{4}{1 + 1.087\,(\max(f_c'-15, 0.1))^{0.23}},\quad
  l = 0.222 + 0.01086\,f_c' - 0.000122\,(f_c')^2.
  \label{eq:kb-fracture-coeffs-all}
\end{equation}
For $f_c' > 31.7$~MPa, $A$, $c$, $d$, $m$, $n$ have extended formulas
(see Appendix~A of~\cite{KoBathe2026}).

\paragraph{Tension.}
The tensile strength ratio follows the fib~Model~Code approximation
\begin{equation}
  t_p = 0.3\,(f_c')^{-1/3},
  \quad\text{clamped to}\quad
  t_p \in [0.05,\, 0.10].
  \label{eq:kb-tp}
\end{equation}

\paragraph{Softening and crack retention.}
The fracture energy is taken as $G_f = 0.06\;\text{N/mm}$ (default), the
element length as $l_b = 100\;\text{mm}$ (default), and the stiffness
retention factors as $\eta_N = 0.0001$, $\eta_S = 0.1$.

Table~\ref{tab:kb-params} summarises the default parameter set for
$f_c' = 30\;\text{MPa}$.

\begin{table}[htb]
\centering
\caption{Default derived parameters for $f_c' = 30\;\text{MPa}$.}
\label{tab:kb-params}
\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Unit} \\
\midrule
$K_e$          & 11\,096  & MPa \\
$G_e$          & 13\,310  & MPa \\
$E$            & 28\,524  & MPa \\
$\nu$          & 0.072    & --- \\[\jot]
$A$            & 0.516    & --- \\
$b$            & 2.07     & --- \\
$c$            & 3.573    & --- \\
$d$            & 2.669    & --- \\[\jot]
$k$            & 1.32     & --- \\
$l$            & 0.438    & --- \\
$m$            & $-2.415$ & --- \\
$n$            & 1.0      & --- \\[\jot]
$t_p$          & 0.0965   & --- \\
$f_t = t_p f_c'$ & 2.90  & MPa \\
\bottomrule
\end{tabular}
\end{table}

% ========================================================================
\section{Implementation Architecture}
\label{sec:kb-architecture}

The implementation resides in a single header:
\begin{center}
\texttt{src/materials/constitutive\_models/non\_lineal/KoBatheConcrete.hh}
\end{center}
and consists of three C++ types.

% ────────────────────────────────────────────────────────────────────────
\subsection{KoBatheParameters}
\label{ssec:kb-params-class}

A plain aggregate containing all material coefficients derived from $f_c'$
at construction time.  Includes the elastic moduli, fracture degradation
coefficients, confining pressure parameters, tension parameters, and
plasticity failure curve constants.

\begin{lstlisting}
struct KoBatheParameters {
    double fc, Ke, Ge, Ee, nue;
    double A_coeff, b_coeff, c_coeff, d_coeff;
    double k_coeff, l_coeff, m_coeff, n_coeff;
    double tp, Gf, lb;
    static constexpr double cp1=3, cp2=36, cp3=12;
    static constexpr double eps_p2_ref=0.0012, eps_p3_ref=0.0025;
    static constexpr double hydro_a=2.5e-5, hydro_s=0.8;
    static constexpr double eta_N=0.0001, eta_S=0.1;

    explicit KoBatheParameters(double fc_MPa,
        double tp_ratio=0.0, double Gf_Nmm=0.06, double lb_mm=100.0);
};
\end{lstlisting}

% ────────────────────────────────────────────────────────────────────────
\subsection{KoBatheState}
\label{ssec:kb-state-class}

The internal history variable set tracks:

\begin{itemize}
  \item \textbf{Fracture history:} maximum octahedral invariants
        $\bar\sigma_o$ and $\bar\tau_o$.
  \item \textbf{Plastic strain:} $\bvarepsilon^p$ in Voigt format
        $(\varepsilon_{xx}^p, \varepsilon_{yy}^p, \gamma_{xy}^p)$.
  \item \textbf{Effective plastic strains:} $\hat{e}_{p1}, \hat{e}_{p2},
        \hat{e}_{p3}$ per yield surface.
  \item \textbf{Crack state:} number of cracks (0, 1, or 2), normal
        directions, opening strains, and closure flags.
  \item \textbf{Committed total strain:} for state tracking.
\end{itemize}

\noindent The state provides accessors \texttt{eps\_p()} and
\texttt{eps\_bar\_p()} for integration with the library's
\texttt{InternalFieldSnapshot} post-processing system.

% ────────────────────────────────────────────────────────────────────────
\subsection{KoBatheConcrete}
\label{ssec:kb-main-class}

The main constitutive class satisfying the \texttt{fall\_n} concept hierarchy:

\begin{itemize}
  \item \textbf{Level 1:} \texttt{ConstitutiveRelation}
    --- \texttt{compute\_response(k, alpha)}, \texttt{tangent(k, alpha)},
        \texttt{commit(alpha, k)}.
  \item \textbf{Level 2b:} \texttt{InelasticConstitutiveRelation}
    --- \texttt{compute\_response(k)}, \texttt{tangent(k)}, \texttt{update(k)},
        \texttt{internal\_state()}.
  \item \textbf{Level 3:} Combines Level~1 and Level~2b.
\end{itemize}

% ────────────────────────────────────────────────────────────────────────
\subsection{Algorithm Flow}
\label{ssec:kb-algorithm}

Algorithm~\ref{alg:kb-evaluate} summarises the evaluation procedure.

\begin{algorithm}[htb]
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\Input{Total strain $\bvarepsilon$, previous state $\alpha_n$}
\Output{Stress $\bsigma$, tangent $\mathbf{C}$, updated state $\alpha_{n+1}$}
$\bvarepsilon^e \leftarrow \bvarepsilon - \bvarepsilon^p_n$ \tcp*{Elastic strain}
$(K_s, G_s, K_c, G_c, \sigma_{\text{id}}) \leftarrow$ fracture moduli from $(\bar\sigma_{o,n}, \bar\tau_{o,n})$\;
$\bsigma \leftarrow \mathbf{C}^{\text{ps}}(K_s, G_s)\,\bvarepsilon^e - \sigma_{\text{id}}\,\boldsymbol{\delta}$ \tcp*{Eq.~\eqref{eq:kb-stress-with-sigma-id}}
$(\sigma_o, \tau_o) \leftarrow$ invariants of $\bsigma$ \tcp*{Eqs.~\eqref{eq:kb-sigma-o}--\eqref{eq:kb-tau-o}}
Update history: $\bar\sigma_o \leftarrow \min(\bar\sigma_{o,n},\sigma_o)$,
  $\bar\tau_o \leftarrow \max(\bar\tau_{o,n},\tau_o)$\;
Recompute moduli and stress with updated history\;
$\mathbf{C}_{\text{tan}} \leftarrow \mathbf{C}^{\text{ps}}(K_c, G_c)$\;
\If{max principal stress $> 0$ \textbf{and} $g(\sigma_o,\tau_o) > 0$}{
  Initiate new crack with normal along max principal direction\;
}
Update crack opening strains, closure flags\;
Apply cracking to $\bsigma$ and $\mathbf{C}_{\text{tan}}$ via $\eta_N, \eta_S$\;
$(\hat{e}_1,\hat{e}_2,\hat{e}_3) \leftarrow$ compression coordinates\;
\If{any $f_j > 0$ (yield violated)}{
  Newton return mapping per surface $\rightarrow$ update $\bvarepsilon^p, \hat{e}_{pj}$\;
  $\bvarepsilon^e \leftarrow \bvarepsilon - \bvarepsilon^p$\;
  $\bsigma \leftarrow \mathbf{C}_{\text{sec}}\,\bvarepsilon^e - \sigma_{\text{id}}\,\boldsymbol{\delta}$\;
}
\Return $(\bsigma, \mathbf{C}_{\text{tan}}, \alpha_{n+1})$\;
\caption{Ko--Bathe constitutive evaluation.}
\label{alg:kb-evaluate}
\end{algorithm}

% ========================================================================
\section{VTK Visualization}
\label{sec:kb-vtk}

The Ko--Bathe model is integrated with the library's VTK post-processing
infrastructure via the \texttt{InternalFieldSnapshot} mechanism.  The
following fields are exported at each integration (Gauss) point for
visualization in ParaView:

\begin{table}[htb]
\centering
\caption{VTK Gauss-point fields exported for the Ko--Bathe concrete model.}
\label{tab:kb-vtk-fields}
\begin{tabular}{@{}llp{7cm}@{}}
\toprule
\textbf{Field name} & \textbf{Components} & \textbf{Description} \\
\midrule
\texttt{qp\_stress\_voigt}      & $N_{\text{voigt}}$ & Cauchy stress tensor (standard) \\
\texttt{qp\_strain\_voigt}      & $N_{\text{voigt}}$ & Total strain tensor (standard) \\
\texttt{qp\_plastic\_strain\_voigt} & $N_{\text{voigt}}$ & Plastic strain tensor \\
\texttt{qp\_equivalent\_plastic\_strain} & 1 & $\sqrt{\hat{e}_{p1}^2+\hat{e}_{p2}^2+\hat{e}_{p3}^2}$ \\[\jot]
\texttt{qp\_num\_cracks}        & 1 & Number of active cracks (0, 1, 2) \\
\texttt{qp\_crack\_normal\_1}   & 3 & Normal to crack 1 (3D vector) \\
\texttt{qp\_crack\_normal\_2}   & 3 & Normal to crack 2 (3D vector) \\
\texttt{qp\_crack\_strain\_1}   & 1 & Opening strain across crack 1 \\
\texttt{qp\_crack\_strain\_2}   & 1 & Opening strain across crack 2 \\
\texttt{qp\_crack\_closed\_1}   & 1 & Closure flag: 1.0 = closed \\
\texttt{qp\_crack\_closed\_2}   & 1 & Closure flag: 1.0 = closed \\[\jot]
\texttt{qp\_sigma\_o\_max}      & 1 & Max octahedral normal stress \\
\texttt{qp\_tau\_o\_max}        & 1 & Max octahedral shear stress \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Oriented fracture planes.}
The crack normal vectors \texttt{qp\_crack\_normal\_1} and
\texttt{qp\_crack\_normal\_2} are stored as 3-component vectors
($n_x, n_y, 0$).  In ParaView, these can be displayed using the
\emph{Glyph} filter:
\begin{enumerate}
  \item Load the Gauss cloud \texttt{.vtu} file.
  \item Apply the Glyph filter with Glyph Type = Disc or Arrow.
  \item Set ``Orientation'' to \texttt{qp\_crack\_normal\_1}.
  \item Set ``Scale'' to \texttt{qp\_num\_cracks} or a constant.
  \item Set ``Masking Mode'' to threshold $\ge 1$ on
        \texttt{qp\_num\_cracks} to show only cracked points.
\end{enumerate}
This produces a visualization of oriented fracture planes at each
integration point where cracks have formed.

\paragraph{Implementation details.}
The snapshot path uses compile-time detection:
\begin{itemize}
  \item \texttt{KoBatheState::eps\_p()} $\to$ \texttt{plastic\_strain}
  \item \texttt{KoBatheState::eps\_bar\_p()} $\to$ \texttt{equivalent\_plastic\_strain}
  \item \texttt{KoBatheState::num\_cracks} $\to$ \texttt{InternalFieldSnapshot::num\_cracks}
  \item \texttt{KoBatheState::crack\_normals} $\to$ promoted to 3D vectors
  \item \texttt{KoBatheState::sigma\_o\_max}, \texttt{tau\_o\_max} $\to$ fracture history scalars
\end{itemize}

% ========================================================================
\section{Concept Satisfaction}
\label{sec:kb-concepts}

The implementation satisfies the three-level concept hierarchy at compile
time, verified via \texttt{static\_assert}:

\begin{lstlisting}
static_assert(ConstitutiveRelation<KoBatheConcrete>);
static_assert(InelasticConstitutiveRelation<KoBatheConcrete>);
static_assert(ExternallyStateDrivenConstitutiveRelation<KoBatheConcrete>);
\end{lstlisting}

% ========================================================================
\section{Test Suite}
\label{sec:kb-tests}

The test suite in \texttt{tests/test\_ko\_bathe\_concrete.cpp} comprises
eight groups (31 checks), all passing:

\begin{enumerate}
  \item \textbf{Compile-time concept checks} (3 static assertions).
  \item \textbf{Parameter derivation} for $f_c'=30$~MPa.
  \item \textbf{Elastic response} at small strain.
  \item \textbf{Uniaxial compression} curves for $f_c'= 21$ and $30$~MPa,
        exported to CSV.
  \item \textbf{Biaxial compression} (equibiaxial and proportional).
  \item \textbf{Tension cracking} with crack formation and softening.
  \item \textbf{Fracture moduli degradation} (stiffness reduction verification).
  \item \textbf{Commit/update cycle consistency} (Level~1 vs Level~2b agreement).
\end{enumerate}

\begin{figure}[htb]
\centering
\begin{tikzpicture}[scale=0.85]
  \draw[->] (0,0) -- (7.5,0) node[right] {$|\varepsilon|$};
  \draw[->] (0,0) -- (0,4.5) node[above] {$|\sigma|$};
  \draw[blue!70!black, thick, smooth] plot coordinates
    {(0,0) (0.5,1.2) (1.0,2.1) (1.5,2.8) (2.0,3.3)
     (2.5,3.6) (3.0,3.8) (3.5,3.85)};
  \draw[red!70!black, thick, smooth] plot coordinates
    {(3.5,3.85) (4.0,3.7) (4.5,3.4) (5.0,3.0) (5.5,2.5)
     (6.0,2.0) (6.5,1.5) (7.0,1.1)};
  \draw[dashed, gray] (0,3.85) -- (3.5,3.85) node[midway, above] {$f_c'$};
  \draw[dashed, gray] (3.5,0) -- (3.5,3.85);
  \node[blue!70!black] at (1.5,3.5) {\small fracturing};
  \node[red!70!black]  at (5.5,3.5) {\small plasticity};
  \draw[gray, thin] (0,0) -- (1.0,2.8) node[midway, left] {\small $E$};
\end{tikzpicture}
\caption{Schematic uniaxial compression response: nonlinear ascending
(fracturing) and post-peak descent (plasticity).}
\label{fig:kb-uniaxial}
\end{figure}

% ========================================================================
\section{Design Decisions}
\label{sec:kb-design-notes}

\begin{enumerate}
  \item \textbf{Single header.}  Consistent with the library's
        header-only philosophy.

  \item \textbf{Rational vs.\ exponential moduli.}  The paper's
        formulation~\eqref{eq:kb-ks}--\eqref{eq:kb-gs} uses rational
        degradation functions (not exponential), giving a better match to
        the experimental ascending branch.

  \item \textbf{Secant--tangent distinction.}  Secant moduli ($K_s, G_s$)
        for stress; tangent moduli ($K_c, G_c$) for the algorithmic
        tangent matrix.

  \item \textbf{Confining pressure $\sigma_{\text{id}}$.}  The stress
        includes a confining pressure correction (Eq.~\eqref{eq:kb-sigma-id})
        that accounts for triaxial effects even in the plane stress setting.

  \item \textbf{Plane stress only.}  Extension to 3D requires
        generalising the compression coordinates, yield surfaces, and
        crack normal tracking to three orthogonal planes.

  \item \textbf{VTK integration.}  The snapshot-based export provides
        crack normal vectors as 3D fields for direct ParaView Glyph rendering of
        oriented fracture planes at integration points.
\end{enumerate}

% ========================================================================
\section{3D Audit Against Ko--Bathe (2026)}
\label{sec:kb-3d-audit}

The current research frontier of the repository is no longer the 2D
\texttt{KoBatheConcrete} path alone but the 3D
\texttt{KoBatheConcrete3D} extension used in the FE\textsuperscript{2}
sub-models.  Because that 3D implementation is central to the current cyclic
validation campaign, it was re-audited directly against the algorithmic
description in Section~2.2 and Table~1 of Ko and
Bathe~\cite{KoBathe2026}.  The audit result is mixed: several concrete
implementation bugs were corrected, but some scientifically important
deviations from the article remain explicit.

\subsection{Corrections verified in code}

The following defects were verified and corrected in the current code base:
\begin{enumerate}
  \item \textbf{Fracture-energy parameter mapping.}  The material factory now
    maps the Ko--Bathe parameter tuple in the correct order
    \((f_c, t_p, G_f, l_b)\).  This removed a real bug in which fracture energy
    and characteristic length could be silently swapped when constructing the
    3D concrete model from the FE\textsuperscript{2} sub-model factories.
  \item \textbf{Local crack-band length-scale rule.}  When the user leaves
    \(l_b = 0\) and requests the local interpretation, the implementation now
    follows the article rule
    \[
      l_b
      =
      \max\!\left(
        \lVert \mathbf{x}_1-\mathbf{x}_2\rVert,\;
        \lVert \mathbf{x}_1-\mathbf{x}_3\rVert,\;
        \lVert \mathbf{x}_1-\mathbf{x}_4\rVert
      \right),
    \]
    rather than an ad hoc geometric surrogate.
  \item \textbf{Final-state crack kinematics.}  After plastic correction, the
    stored crack opening/closure state is now refreshed from the final elastic
    strain, not from the pre-return trial state.  This is important because the
    Ko--Bathe algorithm uses the final elastic state to classify crack opening
    and closure in the cracked branch of Table~1~\cite{KoBathe2026}.
\end{enumerate}

\subsection{Remaining deviations from the article}

The audit also exposed remaining scientific debts that have not yet been
closed in the current implementation.

\paragraph{No-flow classification is now explicit in the 3D implementation.}
The current 3D path now mirrors the top-level decision structure of
Table~1 in~\cite{KoBathe2026}.  After the elastic-fracture predictor, the code
classifies the branch using the trial octahedral invariants:
\begin{align}
  \text{compressive flow}
  &\iff
  {}^{t+\Delta t}\sigma_o^{\mathrm{tr}} < 0
  \;\wedge\;
  {}^{t+\Delta t}\tau_o^{\mathrm{tr}} > {}^{t}\tau_o^{\max},
  \label{eq:kb-3d-flow-classification}
  \\
  \text{no-flow}
  &\iff
  {}^{t+\Delta t}\sigma_o^{\mathrm{tr}} \ge 0
  \;\vee\;
  {}^{t+\Delta t}\tau_o^{\mathrm{tr}} \le {}^{t}\tau_o^{\max}.
  \label{eq:kb-3d-no-flow-classification}
\end{align}
Only the \emph{compressive-flow} branch is now allowed to enter the
compression-coordinate plastic return mapping.  In tensile states and in
compressive unloading, the effective plastic strains
\((\hat\varepsilon_{p1},\hat\varepsilon_{p2},\hat\varepsilon_{p3})\) remain
frozen, matching the article-level interpretation that those states belong to
the no-flow branch.  The implementation also stores the last classified branch
and the trial invariants in the constitutive state for auditability and
regression testing.

This closes a real implementation error.  Before the fix, the 3D code could
still enter the plastic-correction stage during unloading whenever the trial
compression coordinates happened to keep one of the three failure functions
numerically active.  The added regression surface now verifies that once a
mixed cracked-compressive state is unloading or returns to tension, the
effective plastic strains remain unchanged.

\paragraph{Eq.~(26b) is now represented explicitly in the 3D no-flow branch.}
The audited 3D implementation now carries an explicit tensorial update for the
no-flow branch.  Once the trial state has been classified as tensile or as
compressive unloading, the constitutive law is switched to
\begin{equation}
  C_{ijkl}^{\mathrm{NF}}
  =
  \begin{cases}
    C_{ijkl}^{E}(K_e,G_e),
      & {}^{t+\Delta t}\sigma_o^{\mathrm{tr}} \ge 0, \\
    C_{ijkl}^{E}(K_s,G_s),
      & {}^{t+\Delta t}\sigma_o^{\mathrm{tr}} < 0
        \;\wedge\;
        {}^{t+\Delta t}\tau_o^{\mathrm{tr}} \le {}^{t}\tau_o^{\max},
  \end{cases}
  \label{eq:kb-3d-no-flow-ce}
\end{equation}
which is then replaced by the cracked law \(C_{ijkl}^{C}(n)\) when active
cracks are present.  This is the article-consistent correction of a real
implementation defect: the previous code kept using the compressive
fracture/plasticity tangent \((K_c,G_c)\) even in tensile or unloading states.

The internal plastic strain-like history is then updated using the explicit
coupling form
\begin{equation}
  {}^{t+\Delta t}\!\varepsilon^{p}_{ij}
  =
  {}^{t}\!\varepsilon^{p}_{ij}
  -
  {}^{t+\Delta t}\!\varepsilon^{e}_{ij}(N)
  +
  {}^{t+\Delta t}\!\varepsilon^{e}_{ij}(0),
  \label{eq:kb-26b-audit}
\end{equation}
which is the explicit coupling term between fracture and plasticity in the
article's elastic-fracturing branch~\cite{KoBathe2026}.  Computationally, the
current 3D path evaluates
\begin{align}
  {}^{t+\Delta t}\sigma_{ij}(N)
  &\approx
  {}^{t}\sigma_{ij}(N)
  +
  C_{ijkl}^{\mathrm{NF}}
  \Bigl(
    {}^{t+\Delta t}\varepsilon_{kl}^{e}(0)
    -
    {}^{t}\varepsilon_{kl}^{e}(0)
  \Bigr),
  \\
  {}^{t+\Delta t}\varepsilon_{ij}^{e}(N)
  &=
  \bigl(C^{\mathrm{NF}}\bigr)^{\dagger}_{ijkl}
  \,
  {}^{t+\Delta t}\sigma_{kl}(N),
\end{align}
using a robust pseudo-inverse recovery of the elastic strain when the cracked
operator is nearly singular.  The effective plastic strains
\((\hat\varepsilon_{p1},\hat\varepsilon_{p2},\hat\varepsilon_{p3})\) remain
frozen in this branch, exactly as required by the top-level no-flow
classification.

\paragraph{Bounded crack-state stabilization loop in the spirit of Table~1.}
The article resolves crack-status changes through the inner \(m\)-loop of
Table~1, with the crack history updated separately from the material counter
\(k\).  The current 3D implementation now carries a bounded fixed-point
stabilization of the crack-status history rather than a single restart.  Let
\(\chi^{(m)}\) denote the cracked operator state at sub-iteration \(m\),
including the active crack count \(n^{(m)}\) and the open/closed flags.  The
code now performs
\begin{align}
  {}^{t+\Delta t}\sigma_{ij}^{(m)}
  &=
  \begin{cases}
    {}^{t}\sigma_{ij}(N)
    +
    C_{ijkl}^{\mathrm{NF}}\!\bigl(\chi^{(m)}\bigr)
    \Bigl(
      {}^{t+\Delta t}\varepsilon_{kl}^{e}(0)
      -
      {}^{t}\varepsilon_{kl}^{e}(0)
    \Bigr),
    & m = 0 \text{ and } \chi^{(0)} = \chi^{t},
    \\[0.5ex]
    C_{ijkl}^{\mathrm{NF}}\!\bigl(\chi^{(m)}\bigr)
    \,{}^{t+\Delta t}\varepsilon_{kl}^{e}(0),
    & \text{otherwise,}
  \end{cases}
  \\
  {}^{t+\Delta t}\varepsilon_{ij}^{p,(m+1)}
  &=
  {}^{t}\varepsilon_{ij}^{p}
  -
  {}^{t+\Delta t}\varepsilon_{ij}^{e,(m)}(N)
  +
  {}^{t+\Delta t}\varepsilon_{ij}^{e}(0),
\end{align}
then refreshes the crack openings/closures from the final elastic strain and
rebuilds \(C_{ijkl}^{\mathrm{NF}}\bigl(\chi^{(m+1)}\bigr)\).  The iteration is
accepted when
\[
  \chi^{(m+1)} = \chi^{(m)},
\]
or after a bounded number of constitutive sub-iterations
\(m_{\max}=6\) in the current production default.  The implementation records
the number of stabilization passes, the number of crack-state switches, and a
Boolean stabilized/not-stabilized flag in the constitutive diagnostics.  This
is materially closer to the article than the old single-restart
approximation, because it treats the crack operator as a small internal
fixed-point problem rather than as a one-shot reset.

\paragraph{Crack-count reduction is regularized rather than article-faithful.}
The article specifies that if the crack closes, or if the new orientation is
effectively the same as a previous crack, or if the tensile crack strain does
not exceed the historical maximum, the crack count is reduced and the material
can return to a lower-order cracked state or even to the non-cracked
branch~\cite{KoBathe2026}.  The current production implementation instead keeps
the crack object alive and regularizes the transition through explicit
retention factors \((\eta_N,\eta_S)\) and an optional smooth closure ramp.  The
default FE\textsuperscript{2} profile therefore remains intentionally more
robust than the literal article semantics, but it is also less
article-faithful.  Hence the remaining scientific deviation is now narrower:
the open/closed status is iterated in a Table~1-style inner loop, but the
article's explicit crack-count reduction and return to lower-order cracked
states are still regularized rather than transcribed literally.

\subsection{Interpretation}

The practical consequence of this audit is that the current 3D Ko--Bathe path
should be understood as an \emph{article-informed stabilized extension}, not as
a line-by-line transcription of Table~1.  That is acceptable for the present
FE\textsuperscript{2} stabilization phase, but it must remain explicit in the
documentation because the unresolved gaps above directly affect hysteresis,
crack closure, and the interpretation of any ``consistent tangent'' claim.

% ========================================================================
\section{Possible Extensions}
\label{sec:kb-extensions}

\begin{itemize}
  \item \textbf{Full 3D generalisation}: extend compression coordinates
        to all three principal directions; add a third crack plane.
  \item \textbf{Consistent tangent for plasticity}: derive the full
        algorithmic tangent including the plastic correction.
  \item \textbf{Rate effects}: temperature and strain-rate dependence through
        modification of the fracture degradation functions.
  \item \textbf{Mesh-adjusted softening}: use actual element sizes from
        the mesh for the characteristic length $l_b$.
\end{itemize}
