% ==========================================================================
%  Chapter 71 -- Persistent Nonlinear Sub-Model Evolver and L-Shaped
%                Building Multi-Scale Seismic Analysis
% ==========================================================================

\chapter{Persistent Nonlinear Sub-Model Evolution and L-Shaped Building
         Multi-Scale Seismic Analysis}
\label{ch:nl-sub-model-evolver}


% =========================================================================
\section{Introduction and Motivation}
\label{sec:nlevo-intro}

The multiscale seismic analysis framework developed in
Chapter~\ref{ch:multiscale_seismic} provides the infrastructure for
transitioning from a fibre-section beam model to three-dimensional
continuum sub-models at critical elements.  The single-shot solver
\texttt{SubModelSolver} (Chapter~\ref{ch:multiscale_seismic},
\S\,Sub-model solve) and its reinforced variant
(Chapter~\ref{ch:reinforced-submodel}) construct a fresh
\texttt{Model<>} at each call, which
means every solve starts from a \emph{virgin material state}: the
internal variables---crack normals, plastic strains, damage
indices---are discarded after each invocation.

For an earthquake time-history analysis spanning hundreds of time steps,
this approach fails to capture the progressive degradation of the
concrete.  The Ko--Bathe plastic-fracturing model
(Chapter~\ref{ch:ko-bathe-3d}) tracks up to three crack planes per
integration point, storing:
\begin{itemize}
  \item crack normals $\mathbf{n}_1, \mathbf{n}_2, \mathbf{n}_3$,
  \item crack opening strains
        $\varepsilon_{\mathrm{cr},1}, \varepsilon_{\mathrm{cr},2},
         \varepsilon_{\mathrm{cr},3}$,
  \item open/closed status flags,
  \item fracture-history invariants such as the octahedral normal and shear
        envelopes \(\sigma_{o,\max}\) and \(\tau_{o,\max}\).
\end{itemize}
Resetting these variables at each step would allow the same crack to
form repeatedly without accumulating irrecoverable damage, yielding
a fictitiously resilient response.

This chapter presents the \texttt{NonlinearSubModelEvolver}, a
persistent nonlinear continuum sub-model driver that solves the
sub-model boundary value problem \emph{incrementally}, preserving
material history across all earthquake time steps via direct
PETSc~\texttt{SNES} management.  It is demonstrated through a 4-storey
L-shaped reinforced-concrete building subjected to three-component
earthquake excitation from the 2011 T\=ohoku event (station MYG004)
at full intensity (no scaling), with triquadratic Hex27 elements and
detailed crack evolution tracking.


% =========================================================================
\section{Mathematical Framework}
\label{sec:nlevo-math}

% -------------------------------------------------------------------------
\subsection{Section Kinematics from Beam Elements}
\label{ssec:nlevo-section-kin}

A Timoshenko beam element in three dimensions carries six degrees of
freedom per node: translations $(u_x, u_y, u_z)$ and rotations
$(\theta_x, \theta_y, \theta_z)$ about the local axes.  For a given
element $e$ with nodal DOF vector $\mathbf{q}_e \in \mathbb{R}^{12}$
(two-node element) or $\mathbb{R}^{18}$ (three-node), the section
kinematics at a parametric coordinate $\xi \in [-1, +1]$ along the
beam axis are obtained by interpolation:
\begin{align}
  \mathbf{u}_0(\xi) &= \mathbf{N}_u(\xi)\,\mathbf{q}_e,
  \label{eq:nlevo-u0}
  \\
  \boldsymbol{\theta}(\xi) &= \mathbf{N}_\theta(\xi)\,\mathbf{q}_e,
  \label{eq:nlevo-theta}
\end{align}
where $\mathbf{N}_u$ and $\mathbf{N}_\theta$ are the standard
Lagrangian shape function matrices for translations and rotations,
respectively.  The section deformation measures are then:
\begin{alignat}{2}
  \varepsilon_{zz} &= \frac{\partial u_z}{\partial s},
  &\qquad& \text{(axial strain)}
  \label{eq:nlevo-eps-zz}
  \\
  \kappa_x &= \frac{\partial \theta_x}{\partial s},
  &\qquad& \text{(curvature about local $x$)}
  \label{eq:nlevo-kappa-x}
  \\
  \kappa_y &= \frac{\partial \theta_y}{\partial s},
  &\qquad& \text{(curvature about local $y$)}
  \label{eq:nlevo-kappa-y}
  \\
  \gamma_{xz} &= \frac{\partial u_x}{\partial s} - \theta_y,
  &\qquad& \text{(transverse shear $xz$)}
  \label{eq:nlevo-gamma-xz}
  \\
  \gamma_{yz} &= \frac{\partial u_y}{\partial s} + \theta_x,
  &\qquad& \text{(transverse shear $yz$)}
  \label{eq:nlevo-gamma-yz}
\end{alignat}
where $s$ is the arc-length coordinate and the Jacobian
$J = \partial s / \partial \xi$ converts parametric to physical
derivatives.

The complete section state at each end ($\xi = -1$ for face~A,
$\xi = +1$ for face~B) is collected in a \texttt{SectionKinematics}
structure:
\begin{equation}
  \mathcal{K}
  = \bigl\{
      \mathbf{u}_0,\;\boldsymbol{\theta},\;
      \varepsilon_{zz},\;
      \kappa_x,\;\kappa_y,\;
      \gamma_{xz},\;\gamma_{yz},\;
      E,\;G,\;\nu
    \bigr\}.
  \label{eq:nlevo-section-kin}
\end{equation}
The elastic moduli $(E, G, \nu)$ are attached for computing the
Poisson contraction contribution in the displacement reconstruction.


% -------------------------------------------------------------------------
\subsection{Kinematic Reconstruction: From Beam to 3D Continuum}
\label{ssec:nlevo-reconstruction}

The kinematic reconstruction maps the six-DOF section state
$\mathcal{K}$ at a face (A or B) to three-component displacement
boundary conditions on every node of the corresponding sub-model face.
Let the local coordinate system of the cross-section be
$(\hat{\mathbf{e}}_x, \hat{\mathbf{e}}_y, \hat{\mathbf{e}}_z)$, with
$\hat{\mathbf{e}}_z$ along the beam axis.  For a node at local
position $(x_\ell, y_\ell)$ relative to the section centroid, the
reconstructed 3D displacement is:

\begin{equation}
  \boxed{
  \mathbf{u}^{\mathrm{3D}}(x_\ell, y_\ell)
  = \underbrace{\mathbf{u}_0\vphantom{\Big|}}_{\text{centroid}}
  + \underbrace{\boldsymbol{\theta}
    \times \mathbf{r}\vphantom{\Big|}}_{\text{rotation}}
  \underbrace{{} - \nu\,\varepsilon_{zz}\,
    \mathbf{r}_\perp\vphantom{\Big|}}_{\text{Poisson}}
  }
  \label{eq:nlevo-recon}
\end{equation}
where:
\begin{align}
  \mathbf{r} &= x_\ell\,\hat{\mathbf{e}}_x
               + y_\ell\,\hat{\mathbf{e}}_y,
  \\
  \mathbf{r}_\perp &= x_\ell\,\hat{\mathbf{e}}_x
                     + y_\ell\,\hat{\mathbf{e}}_y.
\end{align}

Expanding the cross product
$\boldsymbol{\theta} \times \mathbf{r}$ componentwise:
\begin{align}
  u_x(x_\ell, y_\ell)
    &= u_{x,0} - \theta_z\,y_\ell - \nu\,\varepsilon_{zz}\,x_\ell,
  \label{eq:nlevo-ux}
  \\
  u_y(x_\ell, y_\ell)
    &= u_{y,0} + \theta_z\,x_\ell - \nu\,\varepsilon_{zz}\,y_\ell,
  \label{eq:nlevo-uy}
  \\
  u_z(x_\ell, y_\ell)
    &= u_{z,0} + \theta_x\,y_\ell - \theta_y\,x_\ell.
  \label{eq:nlevo-uz}
\end{align}

Equations~\eqref{eq:nlevo-ux}--\eqref{eq:nlevo-uz} encode three
physical mechanisms:
\begin{enumerate}
  \item \textbf{Rigid-body translation} $\mathbf{u}_0$: bodily
        displacement of the entire section.
  \item \textbf{Rotation-induced displacement}
        $\boldsymbol{\theta} \times \mathbf{r}$: the plane-sections
        hypothesis for bending (axial distribution linear in
        $x_\ell, y_\ell$) and torsion ($\theta_z$ produces in-plane
        displacements).
  \item \textbf{Poisson contraction}
        $-\nu\,\varepsilon_{zz}\,\mathbf{r}_\perp$: lateral contraction
        (or expansion) of the cross-section due to axial strain, scaled
        by the effective Poisson ratio~$\nu$.
\end{enumerate}

The reconstruction is evaluated at face~A ($\xi = -1$, the MinZ face
of the prismatic domain) and face~B ($\xi = +1$, the MaxZ face),
yielding Dirichlet boundary conditions for all nodes on those faces.


% -------------------------------------------------------------------------
\subsection{Incremental Application of Boundary Conditions}
\label{ssec:nlevo-incremental-bc}

The evolver operates in two distinct modes, depending on whether the
sub-model has been solved before.

\paragraph{First step: incremental loading from virgin state.}
At the transition instant (the first time the sub-model is activated),
the material is virgin and the displacement field is zero.  The
target boundary displacements $\mathbf{u}^{\mathrm{target}}_{\mathrm{imp}}$
from~\eqref{eq:nlevo-recon} may be large relative to the elastic
limit of the concrete.  Applying them in a single step would
produce a severe initial residual and likely cause Newton divergence.

Instead, the target is applied in $N_{\mathrm{sub}}$ increments
(default $N_{\mathrm{sub}} = 15$):
\begin{equation}
  \mathbf{u}_{\mathrm{imp}}^{(k)}
  = \frac{k}{N_{\mathrm{sub}}}\,
    \mathbf{u}^{\mathrm{target}}_{\mathrm{imp}},
  \qquad k = 1, 2, \ldots, N_{\mathrm{sub}}.
  \label{eq:nlevo-incremental}
\end{equation}
At each sub-step, a Newton iteration is performed via
\texttt{SNESSolve} from the current converged displacement.  On
convergence, the material state is committed (crack normals, plastic
strains, damage are frozen); on divergence, the state is reverted
to the previous converged snapshot and the analysis is flagged.

\paragraph{Subsequent steps: direct Newton from current state.}
For all following earthquake time steps, the boundary condition
increment between consecutive steps is small (of order $\Delta t$
times the ground velocity), so the previously converged displacement
$\mathbf{U}^{(n)}$ is an excellent initial guess for
$\mathbf{U}^{(n+1)}$.  No sub-stepping is required:
\begin{equation}
  \mathbf{U}^{(n+1)}
  = \text{SNESSolve}\bigl(\mathbf{U}^{(n)}\bigr)
  \quad\text{with updated}~
  \mathbf{u}_{\mathrm{imp}}^{(n+1)}.
  \label{eq:nlevo-subsequent}
\end{equation}

\emph{This is the central design distinction from
\texttt{NonlinearAnalysis::solve\_incremental()}, which resets
$\mathbf{U}$ to zero before every solve.}  By retaining the converged
state, the evolver benefits from quadratic convergence from an
excellent initial guess, and---crucially---the internal material
variables evolve continuously.


% -------------------------------------------------------------------------
\subsection{Residual and Tangent Stiffness Formulation}
\label{ssec:nlevo-residual}

The static equilibrium sub-problem at each step reads:
find $\mathbf{U} \in \mathbb{R}^{n_{\mathrm{free}}}$ such that
\begin{equation}
  \mathbf{R}(\mathbf{U})
  = \mathbf{f}_{\mathrm{int}}(\mathbf{U})
  - \mathbf{f}_{\mathrm{ext}}
  = \mathbf{0},
  \label{eq:nlevo-equilibrium}
\end{equation}
where the internal force vector is assembled element-by-element.  The
external force vector $\mathbf{f}_{\mathrm{ext}}$ is identically zero
for a sub-model driven purely by displacement boundary conditions
(no applied tractions or body forces).

The total displacement at any node is:
\begin{equation}
  \mathbf{u}_{\mathrm{total}}
  = \underbrace{\mathbf{u}_{\mathrm{free}}}_{\text{from SNES vector } \mathbf{U}}
  + \underbrace{\mathbf{u}_{\mathrm{imp}}}_{\text{imposed\_solution()}}.
  \label{eq:nlevo-total-disp}
\end{equation}
The free DOFs $\mathbf{u}_{\mathrm{free}}$ are determined by Newton
iteration; the imposed DOFs $\mathbf{u}_{\mathrm{imp}}$ are prescribed
from the kinematic reconstruction.  In the PETSc implementation, this
decomposition is handled via \texttt{DMGlobalToLocal} (which maps the
free-DOF global vector to the full local vector, placing zeros at
constrained DOFs) and a subsequent \texttt{VecAXPY} to add the
imposed solution.

\paragraph{Residual callback.}
At each Newton iterate with trial $\mathbf{U}^{(i)}$:
\begin{enumerate}
  \item Compute the local total displacement:
        $\mathbf{u}_{\mathrm{loc}} = \mathrm{G2L}(\mathbf{U}^{(i)})
         + \mathbf{u}_{\mathrm{imp}}$.
  \item Assemble element internal forces into a local vector:
        $\mathbf{f}_{\mathrm{int,loc}}
         = \sum_e \mathbf{f}_{\mathrm{int},e}(\mathbf{u}_{\mathrm{loc}})$.
  \item Map back to the global (free-DOF) space:
        $\mathbf{R} = \mathrm{L2G}(\mathbf{f}_{\mathrm{int,loc}})
         - \mathbf{f}_{\mathrm{ext}}$.
\end{enumerate}

\paragraph{Tangent callback.}
The consistent tangent stiffness matrix is assembled analogously:
\begin{equation}
  \mathbf{K}_T
  = \frac{\partial \mathbf{R}}{\partial \mathbf{U}}
  = \sum_e \mathbf{K}_{T,e}(\mathbf{u}_{\mathrm{loc}}),
  \label{eq:nlevo-tangent}
\end{equation}
where each element tangent $\mathbf{K}_{T,e}$ is computed from the
constitutive tangent operator at the current strain state, accounting
for cracked/crushed zones in the Ko--Bathe model.

The Newton update is:
\begin{equation}
  \mathbf{U}^{(i+1)}
  = \mathbf{U}^{(i)}
  - \mathbf{K}_T^{-1}\,\mathbf{R}(\mathbf{U}^{(i)}).
  \label{eq:nlevo-newton}
\end{equation}
The PETSc SNES framework manages convergence monitoring, line-search
strategies, and tolerance checks automatically.  The solver is
configured with:
\begin{itemize}
  \item Direct LU factorisation (\texttt{-pc\_type lu}),
  \item Maximum 50 iterations (\texttt{-snes\_max\_it 50}),
  \item Basic line search (\texttt{-snes\_linesearch\_type basic}).
\end{itemize}


% -------------------------------------------------------------------------
\subsection{Material State Management}
\label{ssec:nlevo-material-state}

After each converged Newton solve, the material state is
\emph{committed}:
\begin{equation}
  \boldsymbol{\sigma}^{(n+1)}_{\mathrm{committed}},\;
  \mathbf{q}^{(n+1)}_{\mathrm{committed}}
  \leftarrow
  \boldsymbol{\sigma}^{(n+1)}_{\mathrm{trial}},\;
  \mathbf{q}^{(n+1)}_{\mathrm{trial}},
  \label{eq:nlevo-commit}
\end{equation}
where $\mathbf{q}$ contains all internal variables (crack normals,
opening strains, plastic strains, damage, and loading/unloading
flags).  This is done by calling
\texttt{elem.commit\_material\_state(u\_local)} for every element, which
evaluates the constitutive response at the converged displacement and
freezes the internal state.

If Newton fails to converge, the material state is \emph{reverted}:
\begin{equation}
  \boldsymbol{\sigma}^{(n+1)}_{\mathrm{trial}},\;
  \mathbf{q}^{(n+1)}_{\mathrm{trial}}
  \leftarrow
  \boldsymbol{\sigma}^{(n)}_{\mathrm{committed}},\;
  \mathbf{q}^{(n)}_{\mathrm{committed}},
  \label{eq:nlevo-revert}
\end{equation}
which restores all Gauss-point states to the last converged snapshot.
The displacement vector $\mathbf{U}$ retains the last converged
values so that the next step can attempt to solve from that baseline.


% -------------------------------------------------------------------------
\subsection{Post-Setup Boundary Condition Update}
\label{ssec:nlevo-bc-update}

The PETSc DMPlex framework applies Dirichlet constraints by
eliminating constrained DOFs from the global vector during
\texttt{model->setup()}.  Thereafter, the method
\texttt{constrain\_node()} cannot be called again to modify
constrained values.

The evolver bypasses this limitation by writing directly into the
model's \texttt{imposed\_solution()} local vector.  For each
constrained node~$i$ with new displacement target
$\mathbf{u}_i^{\mathrm{new}}$:
\begin{enumerate}
  \item Retrieve the DMPlex point index:
        $p = \texttt{domain.node}(i).\texttt{sieve\_id}$.
  \item Query the PetscSection for the DOF offset:
        $\texttt{PetscSectionGetOffset}(section, p, \&offset)$.
  \item Write the three displacement components at
        $\texttt{arr}[offset], \texttt{arr}[offset{+}1],
         \texttt{arr}[offset{+}2]$.
\end{enumerate}

This low-level access is safe because the constraint topology
(which nodes are constrained) does not change---only the
\emph{values} are updated.  The DMPlex remains valid, and the
\texttt{DMGlobalToLocal} / \texttt{DMLocalToGlobal} mappings
continue to operate correctly.


% =========================================================================
\section{Ko--Bathe Concrete in the Multi-Scale Context}
\label{sec:nlevo-ko-bathe}

The three-dimensional Ko--Bathe plastic-fracturing model
(Chapter~\ref{ch:ko-bathe-3d}) governs the material response within
the sub-model.  In the multi-scale context, the model captures:

\begin{enumerate}
  \item \textbf{Crack initiation and propagation.}  When the maximum
        principal stress exceeds the tensile strength $f_t$
        (derived from $f'_c$ via $f_t \approx 0.33\sqrt{f'_c}$~MPa),
        a crack plane is inserted normal to the major principal stress.
        Up to three orthogonal cracks can form at each Gauss point.

  \item \textbf{Crack opening and closing.}  Under cyclic earthquake
        loading, cracks may open, close, and partially reopen.  The
        model tracks the open/closed status and applies a reduced
        shear retention factor $\beta_{\mathrm{cr}}$ for open cracks.

  \item \textbf{Compressive plasticity.}  Beyond the elastic limit in
        compression, the model uses a Drucker--Prager-type yield
        surface with hardening/softening.

  \item \textbf{Fracture-history accumulation.}  The current
        \texttt{KoBatheConcrete3D} implementation exposes persistent crack
        geometry, opening/closing state, and octahedral fracture-history
        invariants.  These quantities evolve monotonically enough to diagnose
        cracking and stiffness degradation even though the public
        \texttt{InternalFieldSnapshot} path does not currently provide a single
        scalar damage channel for this law.
\end{enumerate}

Because these internal variables persist across time steps in the
\texttt{NonlinearSubModelEvolver}, the cyclic degradation during the
earthquake is captured faithfully: a crack that opens at step $n$
remains in the material memory at step $n{+}1$, affecting the local
stiffness and stress redistribution.

\paragraph{Multi-scale RC treatment.}
The sub-model domain is meshed with homogeneous KoBathe concrete
elements (no explicit rebar).  This choice is justified within the
multi-scale framework:
\begin{itemize}
  \item The longitudinal and transverse reinforcement effects are
        captured at the \emph{structural scale} through the fibre
        sections (Kent--Park confined concrete + Menegotto--Pinto
        steel) that govern the beam--column response.
  \item The 3D sub-model zooms into the \emph{concrete matrix} to
        resolve crack patterns, stress redistribution around cracks,
        and damage localisation that the fibre model cannot capture.
  \item The boundary conditions imposed on the sub-model faces
        already encode the integrated effect of the reinforcement
        through the section kinematics extracted from the beam element.
\end{itemize}
This is consistent with the standard hierarchical multi-scale modelling
approach where different resolution levels capture different physical
phenomena.


% =========================================================================
\section{Triquadratic Hex27 Elements}
\label{sec:nlevo-hex27}

The sub-model domain used in this example employs triquadratic
Lagrangian hexahedral elements (Hex27), constructed via
\texttt{PrismaticDomainBuilder} with \texttt{HexOrder::Quadratic}
(Chapter~\ref{ch:ho-prismatic-domain}).  Each element has 27 nodes
(8 vertices + 12 edge midpoints + 6 face centres + 1 body centre)
and is integrated with a $3 \times 3 \times 3$ Gauss quadrature rule
(27 integration points per element).

Compared to the serendipity Hex20 elements (which lack face-centre
and body-centre nodes), Hex27 elements provide:
\begin{itemize}
  \item Complete triquadratic polynomial reproduction: all monomial
        terms up to $\xi^2\eta^2\zeta^2$ are represented exactly.
  \item Superior stress accuracy: the additional interior nodes
        eliminate spurious hourglass-like modes that can appear in
        Hex20 elements under bending-dominated deformations.
  \item Richer crack representation: with $4 \times 4 \times 8 = 128$
        elements and $27 \times 128 = 3{,}456$ Gauss points, the
        sub-model can resolve fine spatial variations in the crack
        pattern.
\end{itemize}

The prismatic mesh with $n_x = 4$, $n_y = 4$, $n_z = 8$ divisions
over a $0.40 \times 0.40 \times 3.20$~m column produces element
aspect ratios of $1{:}1{:}4$ (each element is
$0.10 \times 0.10 \times 0.40$~m), which is acceptable for the
primarily axial-plus-bending deformation mode of a column.


% =========================================================================
\section{Crack Evolution Tracking}
\label{sec:nlevo-crack-evolution}

A key output of the persistent sub-model analysis is the temporal
evolution of cracking: how many Gauss points have cracked, the total
number of individual crack planes, the available fracture-history
envelopes, and the maximum crack opening---all as functions of the
earthquake time.

\subsection{CrackSummary Aggregation}

After each solved step, the evolver collects
\texttt{InternalFieldSnapshot} data from every integration point.
A \texttt{CrackSummary} structure aggregates the statistics:
\begin{equation}
  \texttt{CrackSummary} =
  \bigl\{
    n_{\mathrm{GP, cracked}},\;
    n_{\mathrm{cracks,total}},\;
    \chi_d,\;
    d_{\max},\;
    \sigma_{o,\min},\;
    \tau_{o,\max},\;
    \varepsilon_{\mathrm{cr,max}}
  \bigr\},
  \label{eq:nlevo-crack-summary}
\end{equation}
where:
\begin{itemize}
  \item $n_{\mathrm{GP, cracked}}$ = number of Gauss points with
        at least one crack,
  \item $n_{\mathrm{cracks,total}}$ = total crack planes across all
        cracked Gauss points (each GP can contribute 1--3),
  \item \(\chi_d \in \{0,1\}\) = availability flag for a constitutive-law
        scalar damage channel in the snapshot path,
  \item $d_{\max}$ = maximum scalar damage index among all GPs only when
        \(\chi_d = 1\); otherwise the field is written as \texttt{NaN},
  \item $\sigma_{o,\min}$ = most compressive octahedral normal envelope among
        cracked points,
  \item $\tau_{o,\max}$ = maximum octahedral shear envelope among cracked
        points,
  \item $\varepsilon_{\mathrm{cr,max}}$ = maximum crack opening strain.
\end{itemize}

For the present \texttt{KoBatheConcrete3D} path, the scientifically honest
interpretation is therefore:
\begin{equation}
  \chi_d = 0
  \quad\Longrightarrow\quad
  \text{use }
  \{n_{\mathrm{GP, cracked}}, n_{\mathrm{cracks,total}},
    \sigma_{o,\min}, \tau_{o,\max}, \varepsilon_{\mathrm{cr,max}}\}
  \text{ as the fracture indicators.}
\end{equation}

\subsection{CSV Output Format}

At each evolution step, a row is written to
\texttt{crack\_evolution.csv}:
\begin{verbatim}
time,total_cracked_gps,total_cracks,damage_scalar_available,
peak_damage_scalar,most_compressive_sigma_o_max,max_tau_o_max,max_opening,...
\end{verbatim}
This allows post-processing scripts to plot:
\begin{itemize}
  \item Crack count vs.\ time: visualising the onset and progression
        of cracking with respect to ground motion intensity.
  \item Fracture-history evolution: monitoring
        \(\sigma_{o,\min}\), \(\tau_{o,\max}\), and
        \(\varepsilon_{\mathrm{cr,max}}\) whenever no scalar damage channel is
        available.
  \item Per-sub-model comparison: comparing crack evolution across
        different column elements to identify the most vulnerable
        structural member.
\end{itemize}


\subsection{Crack Plane Visualisation}

In addition to the scalar statistics, for each integration point with
$n_{\mathrm{crack}} \geq 1$, a detailed \texttt{CrackRecord} is stored
containing:
\begin{itemize}
  \item Global coordinates $\mathbf{x}_{\mathrm{GP}}$ of the Gauss point.
  \item Crack normals $\mathbf{n}_1, \mathbf{n}_2, \mathbf{n}_3
        \in \mathbb{R}^3$.
  \item Crack opening strains $\varepsilon_{\mathrm{cr},1},
        \varepsilon_{\mathrm{cr},2}, \varepsilon_{\mathrm{cr},3}$.
  \item Open/closed status flags.
  \item Damage index $d$.
\end{itemize}

Each active crack is exported as a small \texttt{VTK\_QUAD} cell
centred at the Gauss point and oriented perpendicular to the
crack normal.  Two tangent vectors $\mathbf{t}_1, \mathbf{t}_2$ are
computed as:
\begin{equation}
  \mathbf{t}_1 = \frac{\mathbf{n}_k \times \mathbf{e}_{\mathrm{ref}}}
                      {\|\mathbf{n}_k \times \mathbf{e}_{\mathrm{ref}}\|},
  \qquad
  \mathbf{t}_2 = \mathbf{n}_k \times \mathbf{t}_1,
  \label{eq:nlevo-crack-tangents}
\end{equation}
where $\mathbf{e}_{\mathrm{ref}}$ is $\hat{\mathbf{e}}_x$ or
$\hat{\mathbf{e}}_y$ (whichever is less parallel to $\mathbf{n}_k$).
The quad corners are:
\begin{equation}
  \mathbf{x}_j = \mathbf{x}_{\mathrm{GP}}
  + \frac{h}{2}
    \bigl[(\pm 1)\,\mathbf{t}_1 + (\pm 1)\,\mathbf{t}_2\bigr],
  \qquad j = 1,\ldots,4,
  \label{eq:nlevo-crack-quad}
\end{equation}
where $h = 0.4\,\min(\Delta x, \Delta y, \Delta z)/2$ is a sizing
factor based on the element dimensions.  Cell data arrays carry the
opening strain, normal vector, and crack state (open${}=1$,
closed${}=0$), enabling ParaView visualisation via Glyph or Threshold
filters.

Three PVD time-series files are maintained per sub-model:
\begin{itemize}
  \item \texttt{nlsub\_<id>\_mesh.pvd} --- deformed mesh with
        displacement and material fields (stress, strain, damage).
  \item \texttt{nlsub\_<id>\_gauss.pvd} --- Gauss-point cloud.
  \item \texttt{nlsub\_<id>\_cracks.pvd} --- crack plane quads.
\end{itemize}


% =========================================================================
\section{Implementation}
\label{sec:nlevo-implementation}

% -------------------------------------------------------------------------
\subsection{Class Architecture}
\label{ssec:nlevo-class}

The \texttt{NonlinearSubModelEvolver} class manages:
\begin{enumerate}
  \item A persistent \texttt{Model<ThreeDimensionalMaterial,
        SmallStrain, 3>} instance (\texttt{HomogModel}) with
        KoBathe concrete at every element.
  \item PETSc SNES objects: \texttt{SNES snes\_}, \texttt{Vec U\_}
        (global free-DOF displacement), \texttt{Vec R\_} (residual),
        \texttt{Vec f\_ext\_} (external forces, identically zero),
        and \texttt{Mat J\_} (tangent stiffness).
  \item A \texttt{Context} struct holding pointers to the model and
        external force vector, passed to the static SNES callbacks.
  \item PVD writers for mesh, Gauss, and crack output.
  \item A vector of \texttt{CrackRecord} containing the latest crack
        data.
\end{enumerate}

The lifecycle proceeds as:
\begin{enumerate}
  \item \textbf{Construction}: store reference to
        \texttt{MultiscaleSubModel}, $f'_c$, output paths.
  \item \textbf{First \texttt{solve\_step()} call}: create model,
        apply constraints, setup, create SNES, incremental load.
  \item \textbf{Subsequent calls}: update
        \texttt{imposed\_solution()} via PetscSection offsets,
        \texttt{SNESSolve} from current $\mathbf{U}$.
  \item \textbf{After each solve}: collect crack data, write VTK.
  \item \textbf{Finalisation}: write PVD files.
\end{enumerate}

The class supports move semantics but prohibits copying, since the
PETSc objects are non-copyable resources.  The move constructor uses
\texttt{std::exchange} for all PETSc handles and rebinds the
\texttt{Context} pointer.


% -------------------------------------------------------------------------
\subsection{SNES Callbacks}
\label{ssec:nlevo-snes-callbacks}

Two static member functions serve as PETSc callbacks:

\paragraph{FormResidual.}
\begin{enumerate}
  \item Convert global $\mathbf{U}$ to local via
        \texttt{DMGlobalToLocal}.
  \item Add \texttt{imposed\_solution()} to obtain
        $\mathbf{u}_{\mathrm{total}}$.
  \item Loop over elements, calling
        \texttt{compute\_internal\_forces(u\_local, f\_local)}.
  \item Map local $\mathbf{f}_{\mathrm{int}}$ to global
        $\mathbf{R}$ via \texttt{DMLocalToGlobal(ADD\_VALUES)}.
  \item Subtract $\mathbf{f}_{\mathrm{ext}}$ (zero for sub-models).
\end{enumerate}

\paragraph{FormJacobian.}
\begin{enumerate}
  \item Zero the matrix $\mathbf{J}$.
  \item Compute local total displacement (same as above).
  \item Loop over elements, calling
        \texttt{inject\_tangent\_stiffness(u\_local, J)}.
  \item Assemble with \texttt{MatAssembly\{Begin,End\}}.
\end{enumerate}

These callbacks are identical in structure to
\texttt{NonlinearAnalysis::FormResidual/FormJacobian}
(Chapter~\ref{ch:analysis_model_architecture}), but are owned by the
evolver instance rather than the analysis wrapper, enabling the
critical feature of persistent $\mathbf{U}$.


% =========================================================================
\section{L-Shaped Building Example}
\label{sec:nlevo-lshaped}

% -------------------------------------------------------------------------
\subsection{Structural Configuration}
\label{ssec:nlevo-structure}

A 4-storey reinforced-concrete moment frame with an L-shaped floor
plan is modelled.  The plan grid coordinates are:

\begin{center}
\begin{tabular}{ll}
  \toprule
  Direction & Grid coordinates \\
  \midrule
  $x$       & $0,\; 5,\; 10,\; 15$~m \\
  $y$       & $0,\; 4,\; 8$~m \\
  \bottomrule
\end{tabular}
\end{center}

The L-shape is achieved by cutting out the bay at grid indices
$(i_x \in [2,3),\; i_y \in [1,2))$ above all storeys, removing the
upper-right quadrant of the full rectangular plan.  Each storey has a
height of 3.2~m; the total building height is 12.8~m.  The resulting
structural model contains 55 nodes, 44 column elements, and 60 beam
elements (104 structural elements total).

Section properties:

\begin{center}
\begin{tabular}{lcccc}
  \toprule
  Member & $b$ [m] & $h$ [m] & $f'_c$ [MPa]
         & Steel \\
  \midrule
  Columns & 0.40 & 0.40 & 28 & $f_y = 420$~MPa \\
  Beams   & 0.30 & 0.50 & 25 & $f_y = 420$~MPa \\
  \bottomrule
\end{tabular}
\end{center}

Fibre sections use Kent--Park confined concrete (for columns with
$s_t = 0.10$~m tie spacing) or unconfined concrete (for beams), and
Menegotto--Pinto steel with hardening ratio $b = 0.01$.

\paragraph{Mass and damping.}
Lumped mass is computed with $\rho = 2.4 \times 10^{-3}$~MN$\cdot$s$^2$/m$^4$.
Rayleigh damping targets $\xi = 5\%$ at $T_1 \approx 0.60$~s and
$T_3 \approx 0.15$~s.


% -------------------------------------------------------------------------
\subsection{Three-Component Ground Motion}
\label{ssec:nlevo-ground-motion}

The 2011 T\=ohoku earthquake record from K-NET station MYG004
(Tsukidate, Miyagi Prefecture) is applied simultaneously in three
components:

\begin{center}
\begin{tabular}{lcccc}
  \toprule
  Component & DOF & PGA [m/s$^2$] & PGA [$g$] & $t_{\mathrm{PGA}}$ [s] \\
  \midrule
  NS & $x$ & 26.79 & 2.731 & 70.60 \\
  EW & $y$ & 12.85 & 1.310 & 70.66 \\
  UD & $z$ & 18.87 & 1.923 & 70.31 \\
  \bottomrule
\end{tabular}
\end{center}

The records are trimmed to $[25, 300]$~s (skipping the pre-event
noise) and applied at \textbf{full intensity} (scale factor $= 1.0$,
no amplification or reduction) to capture the realistic severity of
this near-fault, high-intensity event.  With PGA exceeding 2.7$g$ in
the NS direction, the structure is subjected to extreme demands that
drive extensive cracking and damage in the sub-models.

Time integration uses the generalized-$\alpha$ method with spectral
radius $\rho_\infty = 0.9$ and a fixed time step
$\Delta t = 0.02$~s (matching the record sampling interval).


% -------------------------------------------------------------------------
\subsection{Phase 1: Global Dynamic Analysis with Transition Monitoring}
\label{ssec:nlevo-phase1}

A \texttt{DynamicAnalysis} solver is configured with a
\texttt{CompositeObserver} containing:
\begin{enumerate}
  \item \texttt{DamageTracker} --- monitors the maximum fibre strain
        in every element and maintains a ranked list of the top-10
        most-damaged elements.
  \item \texttt{FiberHysteresisRecorder} --- records
        $(\varepsilon, \sigma)$ histories for the 5 most-stressed
        fibres per material class (concrete and steel), enabling
        post-processing of hysteretic loops.
  \item \texttt{NodeRecorder} --- records displacement time histories
        at all roof-level nodes (3 DOFs each).
\end{enumerate}

A \texttt{TransitionDirector} with a
\texttt{MaxStrainDamageCriterion} halts Phase~1 the instant any steel
fibre anywhere in the structure exceeds its yield strain:
\begin{equation}
  \varepsilon_{\max,e} > \varepsilon_y
  = \frac{f_y}{E_s}
  = \frac{420}{200{,}000}
  = 0.0021.
  \label{eq:nlevo-yield-criterion}
\end{equation}

The transition report records the time of first yield, the critical
element index, and the damage metric.  Given the extreme ground
motion intensity (PGA~$>\,$2.7$g$), first yield is expected well
before the PGA arrival at $t \approx 45.6$~s (accounting for the
25~s trim offset from the original 70.6~s).


% -------------------------------------------------------------------------
\subsection{Phase 2: Multi-Scale Sub-Model Evolution}
\label{ssec:nlevo-phase2}

\paragraph{Critical element identification.}
After Phase~1 halts, the top-3 most-damaged beam elements
(with damage index $> 0.8$) are identified from the
\texttt{DamageTracker} ranking.  If fewer than 3 elements exceed
the threshold, the single critical element from the transition
report is used.

\paragraph{Sub-model construction.}
For each critical column, a \texttt{MultiscaleCoordinator}:
\begin{enumerate}
  \item Extracts section kinematics at both ends ($\xi = \pm 1$).
  \item Builds a prismatic domain via \texttt{PrismaticDomainBuilder}
        with $n_x = 4,\; n_y = 4,\; n_z = 8$ Hex27 elements.
  \item Computes displacement boundary conditions on the two end faces
        from~\eqref{eq:nlevo-ux}--\eqref{eq:nlevo-uz}.
\end{enumerate}

\paragraph{Sub-model evolution loop.}
A \texttt{NonlinearSubModelEvolver} is instantiated per critical
column with $f'_c = 28$~MPa.  The global dynamic analysis then
continues step-by-step.  At each step:
\begin{enumerate}
  \item The global solver advances one time step via
        \texttt{solver.step\_n(1)}.
  \item For each sub-model evolver:
    \begin{enumerate}
      \item Section kinematics are re-extracted from the (updated)
            beam element.
      \item \texttt{update\_kinematics(\ldots)} recomputes face BCs
            via~\eqref{eq:nlevo-recon}.
      \item \texttt{solve\_step(t)} performs Newton iteration from
            the previous converged state.
      \item \texttt{crack\_summary()} returns the aggregate statistics
            for the CSV log.
    \end{enumerate}
  \item Crack evolution data is written to CSV.
  \item Peak damage across the global model is updated.
  \item VTK snapshots are written every 50 steps (1.0~s intervals).
\end{enumerate}

This loop continues until the end of the 275~s analysis window
(corresponding to $t = 300$~s in the original record).

\paragraph{Output files.}
The analysis produces:
\begin{itemize}
  \item \texttt{yield\_state.vtm} --- global frame VTK at first yield.
  \item \texttt{evolution/frame\_NNNNNN.vtm} + \texttt{frame.pvd} ---
        global frame time series.
  \item \texttt{evolution/sub\_models/nlsub\_<id>\_\{mesh,gauss,cracks\}.pvd} ---
        per sub-model PVD series.
  \item \texttt{recorders/roof\_displacement.csv} --- nodal displacement
        histories at roof level.
  \item \texttt{recorders/fiber\_hysteresis\_\{concrete,steel\}.csv} ---
        fibre stress--strain histories.
  \item \texttt{recorders/crack\_evolution.csv} --- crack count, damage,
        and opening vs.\ time for each sub-model.
\end{itemize}


% =========================================================================
\section{Incremental Kinematics Coupling: Full Derivation}
\label{sec:nlevo-coupling-derivation}

This section presents the complete mathematical chain from beam DOFs
to sub-model Dirichlet boundary conditions.

\paragraph{Step 1: Element DOF extraction.}
For beam element $e$ with element DOF vector
$\mathbf{q}_e \in \mathbb{R}^{12}$ (2-node Timoshenko):
\begin{equation}
  \mathbf{q}_e
  = \bigl[
      u_{x,A},\; u_{y,A},\; u_{z,A},\;
      \theta_{x,A},\; \theta_{y,A},\; \theta_{z,A},\;
      u_{x,B},\; u_{y,B},\; u_{z,B},\;
      \theta_{x,B},\; \theta_{y,B},\; \theta_{z,B}
    \bigr]^T.
  \label{eq:nlevo-qe}
\end{equation}

\paragraph{Step 2: Interpolation at end sections.}
At $\xi = -1$ (face~A), the linear shape functions give:
\begin{align}
  \mathbf{u}_0^A &= (u_{x,A},\; u_{y,A},\; u_{z,A})^T, \\
  \boldsymbol{\theta}^A &= (\theta_{x,A},\; \theta_{y,A},\;
    \theta_{z,A})^T.
\end{align}
The section deformation measures at face~A require derivative
evaluation:
\begin{align}
  \varepsilon_{zz}^A
    &= \frac{1}{L}(u_{z,B} - u_{z,A}), \\
  \kappa_x^A
    &= \frac{1}{L}(\theta_{x,B} - \theta_{x,A}), \\
  \kappa_y^A
    &= \frac{1}{L}(\theta_{y,B} - \theta_{y,A}),
\end{align}
where $L$ is the element length.  Analogous expressions hold for
face~B at $\xi = +1$.

\paragraph{Step 3: Displacement reconstruction on sub-model face.}
Let the sub-model face have $n_f$ nodes at local coordinates
$\{(x_{\ell,j}, y_{\ell,j})\}_{j=1}^{n_f}$ relative to the section
centroid.  The $3 \times 1$ displacement vector at node $j$ is:
\begin{equation}
  \mathbf{u}_j^{\mathrm{face}}
  = \underbrace{
    \begin{pmatrix} u_{x,0} \\ u_{y,0} \\ u_{z,0} \end{pmatrix}
    }_{\text{centroid}}
  + \underbrace{
    \begin{pmatrix}
      -\theta_z\,y_{\ell,j} \\
       \theta_z\,x_{\ell,j} \\
       \theta_x\,y_{\ell,j} - \theta_y\,x_{\ell,j}
    \end{pmatrix}
    }_{\text{rotation}}
  + \underbrace{
    \begin{pmatrix}
      -\nu\,\varepsilon_{zz}\,x_{\ell,j} \\
      -\nu\,\varepsilon_{zz}\,y_{\ell,j} \\
      0
    \end{pmatrix}
    }_{\text{Poisson}}.
  \label{eq:nlevo-full-recon}
\end{equation}

\paragraph{Step 4: Time stepping.}
At each earthquake step $n$, the beam element DOFs
$\mathbf{q}_e^{(n)}$ are extracted from the global state vector.
The section kinematics $\mathcal{K}_A^{(n)}, \mathcal{K}_B^{(n)}$ are
computed, and the face displacements
$\{\mathbf{u}_j^{\mathrm{face}}\}_j$ are evaluated
via~\eqref{eq:nlevo-full-recon}.  These become the Dirichlet BCs for
the sub-model Newton solve at step $n$.

The \emph{increment} in the boundary displacement between steps
is:
\begin{equation}
  \Delta\mathbf{u}_j^{(n)}
  = \mathbf{u}_j^{\mathrm{face},(n)}
  - \mathbf{u}_j^{\mathrm{face},(n-1)},
  \label{eq:nlevo-delta-bc}
\end{equation}
which is typically small compared to the total displacement.  This
ensures that the Newton solver converges rapidly from the previous
state, usually in 2--4 iterations.

\paragraph{Step 5: Equilibrium solve.}
The sub-model solves:
\begin{equation}
  \mathbf{f}_{\mathrm{int}}
    \bigl(\mathbf{u}_{\mathrm{free}}^{(n)}
          + \mathbf{u}_{\mathrm{imp}}^{(n)}\bigr)
  = \mathbf{0}
  \label{eq:nlevo-submodel-eq}
\end{equation}
for $\mathbf{u}_{\mathrm{free}}^{(n)}$, with initial guess
$\mathbf{u}_{\mathrm{free}}^{(n-1)}$.  The constitutive response at
each Gauss point accounts for the full crack/damage history from
steps $1, \ldots, n{-}1$.


% =========================================================================
\section{Discussion}
\label{sec:nlevo-discussion}

\paragraph{Convergence behaviour.}
The first solve (incremental from virgin state with $N_{\mathrm{sub}}
= 15$ sub-steps) typically requires 5--10 Newton iterations per
sub-step due to the nonlinearity of crack initiation.  Subsequent
solves converge in 2--4 iterations thanks to the small BC increments
and the excellent initial guess from the previous converged state.

\paragraph{Computational cost.}
Each sub-model has $4 \times 4 \times 8 = 128$ Hex27 elements with
$27 \times 128 = 3{,}456$ Gauss points and
approximately $9 \times 9 \times 17 = 1{,}377$ nodes (accounting for
the Hex27 lattice).  With 3 DOFs per node, the free-DOF system
(after constraining face nodes) has approximately 3{,}500--3{,}800 DOFs.
A direct LU solve of this size is very fast: under 0.1~s per Newton
iteration.  For 3 sub-models with $\sim$5 iterations each per step,
the sub-model overhead is approximately 1.5~s per global step---a
modest cost relative to the global dynamic solve that includes
mass matrix assembly, Rayleigh damping, and the generalized-$\alpha$
time integrator.

\paragraph{Persistent state vs.\ single-shot.}
The persistent evolver produces fundamentally different results from
the single-shot \texttt{SubModelSolver}: once a crack forms at a Gauss
point, it persists and affects all subsequent steps.  Under cyclic
earthquake loading, this leads to progressive stiffness degradation,
stress redistribution to intact regions, and eventually crack
localisation.  The single-shot solver, by contrast, would produce the
same crack pattern at every step for the same boundary displacements,
regardless of prior loading history.

\paragraph{Full-scale earthquake effects.}
With the MYG004 record applied at full intensity (PGA $>$ 2.7$g$ NS),
the structure experiences extreme demands.  The sub-models are
expected to show:
\begin{itemize}
  \item Early onset of cracking (during the build-up phase before PGA).
  \item Rapid increase in cracked Gauss points during the intense
        shaking phase ($t \approx 45$--$95$~s after the trim offset).
  \item Potential crack saturation (many GPs reaching the 3-crack
        maximum) at peak demands.
  \item Partial recovery during the coda phase as the ground motion
        subsides, with cracks closing but damage persisting.
\end{itemize}

\paragraph{Limitations and extensions.}
\begin{itemize}
  \item The current sub-models use homogeneous concrete without
        explicit rebar.  For columns with high confinement or
        significant shear-reinforcement interaction, embedded truss
        elements (Chapter~\ref{ch:truss-element}) could be added via
        \texttt{MixedModel}.  This would require extending the evolver
        to handle both element types, with specialised crack data
        collection for the concrete elements.
  \item The kinematic coupling is one-way: the global beam model is
        not affected by the sub-model response.  A two-way coupling
        (where the sub-model feeds back homogenised stiffness to the
        beam) would require modifications to the global dynamic solver.
  \item Geometric nonlinearity (large rotations) is not considered in
        the reconstruction~\eqref{eq:nlevo-recon}.  For columns with
        extreme inter-storey drifts, a co-rotational formulation would
        be more appropriate.
\end{itemize}
