% =============================================================================
%  ch82_cyclic_validation.tex — Progressive cyclic validation of FE² pipeline
% =============================================================================

\chapter{Progressive Cyclic Validation of the FE² Pipeline}
\label{ch:cyclic-validation}

This chapter documents a progressive validation suite for the
nonlinear analysis and FE² multiscale pipeline.  Five cases of
increasing complexity are solved under the same cyclic displacement
protocol, using a \enquote{4-legged table} geometry as the base
structure.

%----------------------------------------------------------------------
\section{Motivation}
\label{sec:cyclic-motivation}

The seismic table analysis of Chapter~\ref{ch:table-multiscale}
demonstrated the full FE² pipeline but posed challenges in isolating
the source of convergence difficulties.  This validation suite
replaces the dynamic seismic excitation with a quasi-static
displacement-controlled cyclic protocol, guaranteeing that each
amplitude level is reached regardless of dynamic effects.  The
progressive structure isolates complexity layer by layer:

\begin{enumerate}
  \item \textbf{Case~1} — Single beam element with RC fiber sections,
    validating the constitutive response at the section level.
  \item \textbf{Case~2} — Single continuum column with KoBathe
    concrete and embedded rebar, validating the 3D constitutive model.
  \item \textbf{Case~3} — Full table structural model (4~beams +
    MITC16 shell), validating the mixed-element nonlinear solver.
  \item \textbf{Case~4} — Table with FE² sub-models in one direction
    (not yet implemented).
  \item \textbf{Case~5} — Table with FE² sub-models in two directions
    (not yet implemented).
\end{enumerate}


%----------------------------------------------------------------------
\section{Shared Geometry and Material Properties}
\label{sec:cyclic-geometry}

All cases share the same RC column section and material parameters,
matching the table geometry of Chapter~\ref{ch:table-multiscale}:

\begin{table}[htbp]
\centering
\caption{Shared geometric and material parameters.}
\label{tab:cyclic-params}
\begin{tabular}{lrl}
\toprule
\textbf{Parameter}    & \textbf{Value} & \textbf{Unit} \\
\midrule
Plan dimensions       & $5.0 \times 5.0$ & m \\
Column height $H$     & 3.2        & m \\
Column section        & $0.25 \times 0.25$ & m \\
Clear cover           & 0.03       & m \\
Bar diameter          & 16         & mm \\
$f'_c$                & 30.0       & MPa \\
$E_s$                 & 200\,000   & MPa \\
$f_y$                 & 420        & MPa \\
$b$ (hardening)       & 0.01       & --- \\
Slab thickness        & 0.20       & m \\
\bottomrule
\end{tabular}
\end{table}


%----------------------------------------------------------------------
\section{Cyclic Displacement Protocol}
\label{sec:cyclic-protocol}

The loading protocol is a triangular cyclic displacement with four
amplitude levels (Figure~\ref{fig:cyclic-protocol}):

\[
  \delta_k \in \{1, 2, 4, 8\} \times \delta_y
  \qquad \text{where } \delta_y \approx 0.01 \text{ m}
\]

Each amplitude level consists of three linear segments:
$0 \to +\delta_k$, $+\delta_k \to -\delta_k$, and
$-\delta_k \to 0$, giving 12~segments mapped onto the control
parameter $p \in [0, 1]$.  With 120~incremental steps and up to
6~levels of bisection, this provides 10~solver steps per segment.

The displacement protocol function is:
\begin{equation}
  d(p) = \begin{cases}
    f \cdot A_k        & \text{phase 0 (ascending)} \\
    A_k(1 - 2f)        & \text{phase 1 (descending)} \\
    -A_k(1 - f)        & \text{phase 2 (return)}
  \end{cases}
\end{equation}
where $A_k$ is the current amplitude, $f$ is the fractional position
within the current segment, and the segment index is derived from
$p \times 12$.


%----------------------------------------------------------------------
\section{Case 1: Single Cantilever Beam}
\label{sec:cyclic-case1}

A single vertical column from $z = 0$ (clamped) to $z = H$
(displacement-controlled in~$X$) is modelled with
\texttt{TimoshenkoBeamN<N>} for $N = 2, 3, 4$.  The RC fiber section
uses Kent--Park confined/unconfined concrete and Menegotto--Pinto
steel, identical to the table columns.

\begin{itemize}
  \item \textbf{Case~1a:} $N = 2$ (linear, 1 Gauss point)
  \item \textbf{Case~1b:} $N = 3$ (quadratic, 2 Gauss points)
  \item \textbf{Case~1c:} $N = 4$ (cubic, 3 Gauss points)
\end{itemize}

The hysteresis curves (drift vs.\ base shear) validate convergence
of the fiber-section response with increasing element order.


%----------------------------------------------------------------------
\section{Case 2: Single Continuum Column}
\label{sec:cyclic-case2}

A prismatic column ($0.25 \times 0.25 \times 3.2$~m) is meshed with
$2 \times 2 \times 12$ hexahedral elements plus 8~embedded truss
rebar bars.  Three element orders are compared:

\begin{itemize}
  \item \textbf{Case~2a:} Hex8 (trilinear)
  \item \textbf{Case~2b:} Hex20 (serendipity)
  \item \textbf{Case~2c:} Hex27 (triquadratic)
\end{itemize}

The continuum model uses the KoBathe concrete constitutive law
and Menegotto--Pinto embedded rebar, matching the FE² sub-model
material configuration.  The bottom face is fully clamped; the top
face is displacement-controlled in~$X$ with $u_y$ and $u_z$ free.


%----------------------------------------------------------------------
\section{Case 3: Full Table Structural Model}
\label{sec:cyclic-case3}

The complete 4-legged table is assembled with 4~\texttt{BeamElement}
columns and 1~MITC16 shell slab (20~nodes, 5~elements).  All four
base nodes are clamped; the four slab corner nodes are
displacement-controlled in~$X$ with a uniform drift.

This case validates the mixed beam--shell nonlinear solver under
cyclic loading, prior to introducing the multiscale sub-models.


%----------------------------------------------------------------------
\section{Comparative Analysis}
\label{sec:cyclic-comparison}

The hysteresis curves from all cases are plotted on a common
drift--base-shear axis.  Key observations:

\begin{enumerate}
  \item \textbf{Beam convergence (Case~1):} The three beam orders
    should produce similar hysteresis loops, with the quadratic and
    cubic elements converging faster through the nonlinear regime.

  \item \textbf{Continuum mesh sensitivity (Case~2):} The hex element
    order affects the crack-band width and post-peak softening.
    Hex20 and Hex27 capture more distributed cracking than Hex8.

  \item \textbf{Beam vs.\ continuum (Cases 1 vs.\ 2):} The beam
    fiber-section model produces a smoother hysteresis loop than the
    KoBathe concrete model, which exhibits localised cracking and
    potentially asymmetric degradation.

  \item \textbf{Structural model (Case~3):} The full table response
    should be approximately 4$\times$ the single-column shear
    (symmetric plan, uniform drift), modulated by the slab stiffness.
\end{enumerate}


%----------------------------------------------------------------------
\section{FE² Allocation Audit}
\label{sec:fe2-alloc-audit}

An audit of the PETSc vector allocations in the FE² pipeline
(\texttt{NonlinearSubModelEvolver}) was performed.  The key findings:

\begin{itemize}
  \item \texttt{FormResidual}: 2~\texttt{DMGetLocalVector} per SNES
    iteration (plus corresponding restores).
  \item \texttt{FormJacobian}: 1~\texttt{DMGetLocalVector} per SNES
    iteration.
  \item \texttt{compute\_homogenized\_tangent}: 2~\texttt{VecDuplicate}
    per call (for saving/restoring the solution).
  \item \textbf{No memory leak} was found in
    \texttt{section\_forces\_from\_reactions()} —
    \texttt{DMRestoreLocalVector} is correctly called.
  \item \textbf{Optimisation opportunity}: Pre-allocating working
    vectors as class members would eliminate $\sim$80 allocation/deallocation
    cycles per global step per sub-model.
\end{itemize}


%----------------------------------------------------------------------
\section{Implementation}
\label{sec:cyclic-implementation}

The validation driver is implemented in
\texttt{main\_table\_cyclic\_validation.cpp} with a CLI dispatch:

\begin{verbatim}
  ./fall_n_table_cyclic_validation --case 1a   # single case
  ./fall_n_table_cyclic_validation --case all   # all 7 cases
\end{verbatim}

Each case produces a \texttt{hysteresis.csv} file and VTK output
in the subdirectory
\texttt{data/output/cyclic\_validation/case\{id\}/}.
A Python plotting script (\texttt{scripts/plot\_cyclic\_validation.py})
generates comparative PDF figures.
