% ==========================================================================
%  Chapter 84 -- Execution record for multiscale stabilization v1
% ==========================================================================

\chapter{Execution Record for Multiscale Stabilization v1}
\label{ch:multiscale-stabilization-execution}

\section{Intent and Verification Protocol}
\label{sec:ms-stab-protocol}

This chapter records the actual execution of the multiscale stabilization plan
after the publication-oriented redesign documented in
Chapter~\ref{ch:multiscale-publication-audit}.  The guiding rule for this
stabilization cycle was strict: neither the previous code, nor the existing
tests, nor the earlier documentation were accepted as ground truth without a
fresh audit against the implemented path.

The execution protocol used in this cycle was:
\begin{enumerate}
  \item audit the current normative path directly in code;
  \item reproduce uncertainty or weakness with a dedicated test or diagnostic;
  \item implement the correction or hardening;
  \item rerun a focused regression gate;
  \item document both the improvement and the residual debt in the same
    \LaTeX{} document.
\end{enumerate}

The main files audited and modified during this cycle were:
\begin{itemize}
  \item \texttt{src/analysis/MultiscaleTypes.hh};
  \item \texttt{src/analysis/MultiscaleAnalysis.hh};
  \item \texttt{src/reconstruction/NonlinearSubModelEvolver.hh};
  \item \texttt{src/numerics/SparseSchurComplement.hh};
  \item \texttt{src/reconstruction/LocalModelAdapter.hh};
  \item \texttt{src/reconstruction/HomogenisationStrategy.hh};
  \item \texttt{src/utils/CyclicProtocol.hh};
  \item \texttt{main\_table\_cyclic\_validation.cpp};
  \item \texttt{scripts/plot\_cyclic\_validation.py};
  \item \texttt{tests/test\_multiscale\_analysis\_api.cpp};
  \item \texttt{tests/test\_evolver\_advanced.cpp};
  \item \texttt{tests/test\_condensation.cpp};
  \item \texttt{tests/test\_cyclic\_protocol.cpp};
  \item \texttt{scripts/ci\_multiscale\_stabilization.ps1};
  \item \texttt{.github/workflows/multiscale-stability.yml}.
\end{itemize}

\section{Phase 0: Truth Baseline}
\label{sec:ms-stab-phase0}

Table~\ref{tab:ms-stab-truth-baseline} summarises the baseline classification
used before any code change was accepted.

\begin{table}[htb]
\centering
\small
\begin{tabular}{p{0.27\textwidth}p{0.13\textwidth}p{0.43\textwidth}}
\toprule
Audited item & Status & Decision in this stabilization cycle \\
\midrule
Typed multiscale orchestration path &
verified &
Kept as the normative public path. \\
Type-erased \texttt{LocalModelHandle} path &
experimental &
Left in code, but explicitly removed from the normative public narrative until
exact erased checkpoint/restore semantics are implemented. \\
\texttt{HomogenisationStrategy} route &
legacy/internal &
Retained only as a comparison utility; no longer described as the production
FE\textsuperscript{2} extension mechanism. \\
Local tangent extraction in
\texttt{NonlinearSubModelEvolver} &
fragile &
Replaced by sparse linearized condensation with explicit validation-mode
fallback and explicit diagnostics. \\
Macro rollback on trial-step failure inside FE\textsuperscript{2} &
incorrect &
Explicit checkpoint restoration added to the failure path. \\
Physical regression matrix for pure load modes &
missing &
Added as dedicated solver-level regression tests. \\
Visible CI gate for the multiscale release surface &
missing &
Added as a documented local script and a repository workflow. \\
\bottomrule
\end{tabular}
\caption{Truth baseline for the stabilization cycle.}
\label{tab:ms-stab-truth-baseline}
\end{table}

\section{Phase 1: Numerical Hardening of the Local Response}
\label{sec:ms-stab-phase1}

\subsection{Initial hypothesis}

The main numerical fragility of the public multiscale path was the local
tangent extraction: a one-sided perturbation with a fixed user-provided
\(h_{\mathrm{pert}}\), silent zero columns on failed perturbations, and hidden
post-processing by symmetry projection and diagonal clamping.

\subsection{Observed evidence before the change}

The audit of \texttt{NonlinearSubModelEvolver.hh} showed that the previous
implementation:
\begin{itemize}
  \item used only the \(+\) perturbation;
  \item used the same perturbation size for every generalized component;
  \item left the tangent column at zero if the perturbation solve diverged;
  \item restored displacement vectors but did not explicitly resynchronise the
    model state vector in the tangent loop;
  \item reported only the final tangent, thereby hiding which columns were
    valid and whether regularisation had actually modified the matrix.
\end{itemize}

\subsection{Implemented change}

The production operator remains the boundary-reaction operator, but the
stabilization no longer treats finite differences as the primary tangent path.
The current normative implementation first assembles the sparse local tangent
operator at the converged micro state, including the embedded-rebar penalty
terms, partitions it into free and constrained blocks, and condenses it
against the boundary kinematics.  With constrained increments
\(\delta \mathbf{u}_c = \mathbf{T}\,\delta \mathbf{q}\), the returned section
tangent is
\begin{equation}
  \mathbf{D}_{\mathrm{hom}}
  =
  \mathbf{T}^{\mathsf T}
  \mathbf{K}_{\mathrm{eff}}
  \mathbf{T},
  \qquad
  \mathbf{K}_{\mathrm{eff}}
  =
  \mathbf{K}_{cc}
  -
  \mathbf{K}_{cf}\mathbf{K}_{ff}^{-1}\mathbf{K}_{fc}.
\end{equation}
This is the first version of the multiscale operator in the repository that is
explicitly built from the linearised micro equilibrium equations rather than
from perturb-and-observe solves.  In the follow-up hardening pass, the
implementation was further upgraded from a dense prototype to a sparse solve of
the six required right-hand sides, which removed the need to materialise the
full dense Schur complement.

The present execution pass tightened that sparse route again by factoring the
Schur solve into a dedicated reusable utility and by adding a symbolic-pattern
workspace cache.  In practical terms, repeated calls to the local condensed
operator now reuse the symbolic factorisation pattern of the free block
\(\mathbf{K}_{ff}\) whenever the local topology is unchanged, while still
re-factorising the current numerical values at the converged state.  This is a
small but important architectural improvement: the code now makes the
performance boundary explicit instead of hiding symbolic-analysis cost inside
each operator evaluation.

Adaptive finite differences are retained only as a documented fallback.  When
the condensed operator cannot be formed---for example because a mixed element
does not expose explicit local tangent data or because the free block is
numerically singular---the implementation falls back to the adaptive scheme.
For each generalized component \(e_j\), the fallback uses
\begin{equation}
  h_j
  =
  \frac{\max\!\left(h_{\mathrm{pert}} L,\;
  10^{-3}\max(|q_j|, h_{\mathrm{pert}}L)\right)}{L},
\end{equation}
where \(L\) is the macro beam characteristic length and \(q_j\) is the
corresponding relative end degree of freedom.  The preferred tangent estimate
is
\begin{equation}
  \mathbf{D}_{\mathrm{hom}}^{(:,j)}
  \approx
  \frac{\mathbf{s}(\mathbf{e}+h_j \mathbf{e}_j)
       -\mathbf{s}(\mathbf{e}-h_j \mathbf{e}_j)}{2h_j},
\end{equation}
and the implementation falls back to one-sided differentiation only when one of
the two perturbation solves fails.

The present pass extended that fallback story into a runtime audit mode.  The
local solver can now request
\texttt{ValidateCondensationAgainstAdaptiveFiniteDifference}, which first
builds the sparse condensed tangent and then computes an adaptive
finite-difference reference on the same converged state.  The implementation
reports the global operator gap
\begin{equation}
  \eta_{\mathbf{D}}
  =
  \frac{\lVert \mathbf{D}_{\mathrm{cond}}
         - \mathbf{D}_{\mathrm{fd}} \rVert_F}
       {\max\!\left(1,
           \lVert \mathbf{D}_{\mathrm{cond}} \rVert_F,
           \lVert \mathbf{D}_{\mathrm{fd}} \rVert_F\right)}
\end{equation}
and the worst per-column diagnostic
\begin{equation}
  \eta_{\max} = \max_j \eta_j.
\end{equation}
The automatic acceptance gate currently uses \(\eta_{\mathbf{D}}\), while
\(\eta_{\max}\) is retained as a visible diagnostic only.  That choice was
made after the audit showed that raw columnwise Euclidean gaps can be dominated
by the mixed units of section operators (forces and moments in the same
column), which makes them useful as warnings but not yet mature enough to act
as the sole production gate.

The public response object now records:
\begin{itemize}
  \item the tangent computation mode requested;
  \item the tangent scheme used;
  \item the status of the condensed operator attempt;
  \item the runtime validation status of the condensed operator;
  \item the configured validation tolerance;
  \item the measured global and per-column validation gaps;
  \item the sparse condensed solve residual when that path is attempted;
  \item the perturbation size used for each column;
  \item whether each column is valid;
  \item whether each column used the central stencil;
  \item the number of failed perturbation solves;
  \item whether regularisation actually modified the returned tangent.
\end{itemize}

\subsection{Resulting semantics}

The stabilization deliberately stopped calling every non-perfect response
``successful''.  The public status contract is now:
\begin{description}
  \item[\texttt{Ok}] the condensed operator was assembled successfully, all
    tangent columns are valid, no actual regularisation was applied, and any
    requested runtime validation either passed or was not requested;
  \item[\texttt{Degraded}] a usable response exists, but at least one
    fallback to finite differences, actual tangent regularisation, or
    unavailable runtime validation reference occurred;
  \item[\texttt{SolveFailed}] the response is not usable as a section operator;
  \item[\texttt{NotReady}/\texttt{InvalidOperator}] explicit non-solution
    states.
\end{description}

\subsection{Internal architectural hardening}

The numerical hardening was paired with an internal architectural split of the
local solver path.  Instead of leaving boundary conditions, persistent state
management, and boundary-reaction homogenisation interleaved inside
\texttt{NonlinearSubModelEvolver}, the cycle extracted three dedicated
components:
\begin{itemize}
  \item \texttt{PersistentLocalStateOps}: synchronisation, commit, and revert
    of the persistent continuum state;
  \item \texttt{LocalBoundaryConditionApplicator}: update of section-end
    kinematics and writing of the imposed boundary vector;
  \item \texttt{BoundaryReactionHomogenizer}: section forces from reactions,
    condensed tangent extraction, finite-difference fallback, diagnostics, and
    response classification.
\end{itemize}

This split was intentionally implemented as lightweight header-only helpers
composed through pointers and templates, so the hot path still avoids virtual
dispatch and heap churn.  Architecturally, this matters because it creates a
single explicit seam where the current sparse condensation / validation
fallback pair can later be upgraded again without changing the public
\texttt{SectionHomogenizedResponse} contract or the typed local model API.

The same cycle also extended \texttt{FEM\_Element} with the explicit local
linearisation API already present in \texttt{StructuralElement}.  This closes
an architectural gap in the mixed-element path: the condensed local operator
now sees the element DOF map and the element tangent even when the micro model
is stored behind \texttt{MultiElementPolicy} type erasure.

The same cycle also hardened the semantics of operator validity.  A local
response is now classified as \texttt{InvalidOperator} if any tangent column is
missing entirely.  One-sided fallback with a complete operator remains
\texttt{Degraded}; only a fully central, fully valid, non-regularised operator
is reported as \texttt{Ok}.  This change is small in code volume, but
scientifically important because it prevents an incomplete section operator
from being treated as merely ``low quality'' and then entering the FE\textsuperscript{2}
loop as if it were usable.

\subsection{Verification added in the current pass}

This execution pass added three focused checks on top of the previous
stabilization matrix:
\begin{itemize}
  \item \texttt{test\_sparse\_schur\_operator\_matches\_dense\_reference}:
    verifies that the reusable sparse Schur utility reproduces the dense
    reference contraction on a hand-auditable block system;
  \item \texttt{test\_sparse\_schur\_workspace\_reuses\_symbolic\_pattern}:
    verifies that repeated calls with unchanged sparsity reuse the symbolic
    factorisation pattern instead of analysing it again;
  \item \texttt{test\_condensed\_operator\_reuses\_symbolic\_pattern}:
    verifies the same reuse property through the full
    \texttt{NonlinearSubModelEvolver} path, not only in the low-level utility.
  \item \texttt{test\_validated\_condensed\_operator\_accepts\_elastic\_case}:
    verifies that the runtime audit mode keeps the sparse condensed tangent on
    the production path when the measured global operator gap stays below the
    configured tolerance.
  \item \texttt{test\_validated\_condensed\_operator\_rejects\_with\_strict\_tolerance}:
    verifies that the same runtime audit mode falls back explicitly to adaptive
    finite differences and reports \texttt{ValidationRejected} when the
    tolerance is made artificially strict.
\end{itemize}

The purpose of this extra layer is architectural as much as numerical.  The
code now proves, through tests, that the reusable sparse Schur module is not a
hidden refactor-only detail: its semantics propagate all the way to the public
\texttt{SectionHomogenizedResponse}.

After this pass, the full local release gate was rerun again through
\texttt{scripts/ci\_multiscale\_stabilization.ps1 -SkipConfigure}.  The gate
completed successfully, including the heavy multiscale example builds and the
\LaTeX{} document compilation, so the sparse Schur workspace refactor is not
only locally correct in isolation but also stable on the current release
surface.

\section{Phase 2: Protocol Hardening for Macro--Micro Iteration}
\label{sec:ms-stab-phase2}

\subsection{Initial hypothesis}

The checkpointed FE\textsuperscript{2} loop added in the redesign was close to
publication-ready, but the failure semantics were still too optimistic and too
opaque for a release gate.

\subsection{Errors confirmed during audit}

Two issues were confirmed in the code:
\begin{enumerate}
  \item on a macro trial-step failure inside
    \texttt{IteratedTwoWayFE2}, the macro checkpoint was not restored
    explicitly before returning;
  \item \texttt{CouplingIterationReport} did not expose the termination reason,
    rollback execution, regularisation presence, or failed coupling sites.
\end{enumerate}

\subsection{Implemented change}

The orchestrator now records:
\begin{itemize}
  \item \texttt{CouplingTerminationReason};
  \item whether rollback was performed;
  \item whether relaxation changed the current iterate;
  \item whether regularisation was detected among accepted local responses;
  \item the set of hard-failure sites;
  \item a count of regularised submodels.
\end{itemize}

Most importantly, the macro checkpoint is now restored explicitly on macro-step
failure before the FE\textsuperscript{2} step returns.  This closes a real
semantic gap that could otherwise leave a dirty trial state in memory after a
reported failure.

\section{Phase 3: Physical Regression Matrix}
\label{sec:ms-stab-phase3}

\subsection{Added regression coverage}

The stabilization cycle introduced a small deterministic matrix of physical
reference cases using the real \texttt{NonlinearSubModelEvolver} path:
\begin{itemize}
  \item pure axial loading;
  \item pure bending about \(y\);
  \item pure bending about \(z\);
  \item pure shear along \(y\);
  \item pure shear along \(z\);
  \item pure torsion.
\end{itemize}

The tests check:
\begin{itemize}
  \item convergence of the local sub-model;
  \item availability of the homogenised response;
  \item positivity of the expected tangent diagonal;
  \item completion of the tangent diagnostics;
  \item use of \texttt{LinearizedCondensation} and the
    \texttt{forces\_consistent\_with\_tangent} flag in the elastic cases;
  \item dominance or mode-consistent prominence of the expected resultant;
  \item linearised consistency in the elastic axial case;
  \item boundary-reaction versus volume-average agreement in the elastic axial
    case;
  \item axial macro--micro power consistency.
\end{itemize}

\subsection{Observed numerical results}

The focused axial consistency test produced the following values on the local
execution used for this chapter:
\begin{align}
  \text{linearization\_error} &= 7.85096\times 10^{-5}, \\
  \text{operator\_gap} &= 2.84408\times 10^{-5}, \\
  \text{power\_gap} &= 2.84408\times 10^{-9}.
\end{align}
These are sufficiently small for the intended stabilization gate and, more
importantly, provide a reproducible numerical baseline for future refactors.

After the scientific-hardening follow-up, the local gate also compares the
production condensed tangent with a deliberately forced adaptive
finite-difference tangent on the same elastic mixed-mode state.  The measured
comparison on the audited execution was
\begin{align}
  \text{tangent\_gap} &= 4.60221\times 10^{-2}, \\
  \text{force\_gap} &= 0.
\end{align}
This result is intentionally not presented as exact equality.  Its role in the
stabilization gate is to show that the production condensed operator and the
validation fallback remain close on a coupled elastic state while still making
their non-zero discrepancy explicit.

The new runtime audit mode made that comparison part of the API itself rather
than a test-only post-mortem.  On the same elastic mixed-mode state, the
audited runtime validation metrics were
\begin{align}
  \eta_{\mathbf{D}} &= 4.60221\times 10^{-2}, \\
  \eta_{\max} &= 5.82414\times 10^{-1}.
\end{align}
With a validation tolerance of \(7.5\times 10^{-2}\), the runtime audit keeps
the sparse condensed operator on the production path and reports
\texttt{TangentValidationStatus::Accepted}.  With a deliberately strict
tolerance of \(10^{-8}\), the same case falls back to adaptive finite
differences and reports
\texttt{CondensedTangentStatus::ValidationRejected}.  This pair of tests is
important because it closes the scientific-observability gap: the code now
proves both branches of the runtime decision, not just the existence of a
numerical discrepancy.

\subsection{Executed regression targets}

The following focused executables were rebuilt and executed locally after the
runtime-validation extension of the stabilization changes:
\begin{itemize}
  \item \texttt{fall\_n\_multiscale\_api\_test.exe}: 43 passed, 0 failed;
  \item \texttt{fall\_n\_micro\_solve\_executor\_test.exe}: 6 passed, 0 failed;
  \item \texttt{fall\_n\_evolver\_advanced\_test.exe}: 123 passed, 0 failed;
  \item \texttt{fall\_n\_heterogeneity\_test.exe}: 26 passed, 0 failed;
  \item \texttt{fall\_n\_beam\_test.exe}: 21 passed, 0 failed;
  \item \texttt{fall\_n\_steppable\_solver\_test.exe}: 48 passed, 0 failed;
  \item \texttt{fall\_n\_steppable\_dynamic\_test.exe}: 32 passed, 0 failed.
\end{itemize}

\section{Phase 4: Release Hardening and CI Surface}
\label{sec:ms-stab-phase4}

\subsection{Local reproducible gate}

The repository now contains
\texttt{scripts/ci\_multiscale\_stabilization.ps1}, which configures the build
directory if needed, builds the multiscale release gate, executes the key
regression tests, and compiles the PDF.  In the current stabilized form, the
script builds each target individually rather than issuing a single monolithic
\texttt{ninja} invocation; that change was introduced specifically to expose
which heavy target is consuming or blocking the local gate.  The script covers:
\begin{itemize}
  \item \texttt{fall\_n\_multiscale\_api\_test};
  \item \texttt{fall\_n\_micro\_solve\_executor\_test};
  \item \texttt{fall\_n\_evolver\_advanced\_test};
  \item \texttt{fall\_n\_beam\_test};
  \item \texttt{fall\_n\_steppable\_solver\_test};
  \item \texttt{fall\_n\_steppable\_dynamic\_test};
  \item \texttt{fall\_n\_lshaped\_multiscale};
  \item \texttt{fall\_n\_lshaped\_multiscale\_16};
  \item \texttt{fall\_n\_table\_multiscale};
  \item \texttt{fall\_n\_table\_cyclic\_validation};
  \item \texttt{latexmk -pdf}.
\end{itemize}

During this execution cycle, the script itself exposed two practical issues:
\begin{enumerate}
  \item the initial \texttt{param()} placement was invalid for PowerShell and
    had to be corrected;
  \item the document step required a defensive fallback to
    \texttt{pdflatex} on Windows when \texttt{latexmk} propagated a stale
    toolchain/runtime failure.
\end{enumerate}
After those corrections, the equivalent gate commands were executed explicitly
in the audit environment to verify the release surface.

For this follow-up execution cycle, the focused regression executables and the
\LaTeX{} document were revalidated again after the condensed tangent
implementation.  A fresh rebuild of the heavy example targets was also
attempted, including separate launches for the main multiscale examples, and
the target-by-target gate ultimately completed successfully.  In the audited
environment the full local stabilization script rebuilt
\texttt{fall\_n\_lshaped\_multiscale},
\texttt{fall\_n\_lshaped\_multiscale\_16},
\texttt{fall\_n\_table\_multiscale}, and
\texttt{fall\_n\_table\_cyclic\_validation}, reran the regression executables,
and regenerated \texttt{main.pdf}.  The wall-clock cost is still high, but the
important point for release readiness is that the gate now finishes end-to-end
instead of being left as an unverified partial build.

\subsection{Repository workflow}

The repository now also contains the workflow
\texttt{.github/workflows/multiscale-stability.yml}.  Its purpose is not to
claim that hosted CI is already production-proven, but to freeze the intended
release gate in repository form:
\begin{enumerate}
  \item configure an MSYS2 UCRT64 environment on Windows;
  \item install the numerical dependencies used by the project;
  \item configure CMake with Ninja;
  \item build the regression executables one by one;
  \item build the heavy multiscale examples one by one;
  \item run the focused regression executables;
  \item compile the \LaTeX{} document.
\end{enumerate}

\section{Post-Stabilization Merge Audit with \texttt{origin/master}}
\label{sec:ms-stab-merge-audit}

Once the stabilization gate was green locally, the branch was not merged
blindly with \texttt{origin/master}.  Instead, the integration was treated as
another audit phase: every conflicting file was reviewed under the same rule
used throughout this stabilization cycle, namely that older or semantically
weaker code is not accepted just because it came from the remote default
branch.

\subsection{Audit hypothesis}

The initial hypothesis was that the merge could reintroduce one of three
classes of defects:
\begin{enumerate}
  \item semantic regression, where the remote side reintroduced placeholder or
    legacy code that weakens the now-documented public API;
  \item architectural regression, where a low-level header would be coupled
    again to non-normative concerns;
  \item numerical regression, where a geometry or constitutive alias would
    silently change the behaviour of multiscale builds.
\end{enumerate}

\subsection{Conflict classification}

Table~\ref{tab:ms-stab-merge-conflicts} records the files that required
explicit judgment during the merge.

\begin{table}[htb]
\centering
\small
\begin{tabular}{p{0.28\textwidth}p{0.16\textwidth}p{0.40\textwidth}}
\toprule
File & Resolution & Rationale \\
\midrule
\texttt{src/analysis/Analysis.hh} &
kept stabilized branch version &
the remote side only carried a placeholder ``analysis concept'' comment,
whereas the stabilized branch already exposed the complete analysis aggregator
surface (\texttt{DynamicAnalysis}, \texttt{AnalysisDirector},
\texttt{TransitionDirector}, observers). \\
\texttt{src/elements/element\_geometry/LagrangeElement.hh} &
kept stabilized branch version &
the remote side appended stale duplicated geometry/member code after the class
definition.  Accepting it would have reintroduced ambiguity and maintainability
debt. \\
\texttt{src/geometry/Cell.hh} &
kept stabilized branch version &
the remote side tried to embed VTK-specific cell metadata directly in the core
geometry descriptor.  That coupling was rejected because VTK translation is
already owned by the post-processing layer. \\
\texttt{src/materials/MaterialPolicy.hh} &
kept stabilized branch version &
the remote side only preserved historical comments and omitted the explicit
constitutive-space aliases now required by the stabilized public API. \\
\texttt{src/materials/Strain.hh} &
kept stabilized branch version &
the stabilized branch already uses \texttt{set\_components()}, which is the
current normative vocabulary. \\
\texttt{src/materials/VoigtVector.hh} &
merged with compatibility shim &
the stabilized branch kept the stronger \texttt{components()} /
\texttt{set\_components()} semantics, but the legacy \texttt{set\_vector()}
entry point was reintroduced as a thin alias for backward compatibility. \\
\texttt{src/mesh/gmsh/GmshDomainBuilder.hh} &
kept stabilized branch version &
the remote side attempted to map HEX8 elements through a lower-order geometry
and under-integrated rule, which would have been a genuine numerical
regression. \\
\texttt{src/model/Model.hh} &
kept stabilized branch version &
the extra remote \texttt{friend class Analysis;} declaration was not required
by the current architecture and would only preserve dead coupling. \\
\bottomrule
\end{tabular}
\caption{Conflict-by-conflict resolution during the merge with \texttt{origin/master}.}
\label{tab:ms-stab-merge-conflicts}
\end{table}

\subsection{Compatibility retained intentionally}

Not every remote change was rejected.  The merge kept small, safe compatibility
improvements that do not weaken the stabilized architecture:
\begin{itemize}
  \item \texttt{VoigtVector} now exposes a compatibility
    \texttt{set\_vector(...)} alias that forwards to
    \texttt{set\_components(...)}; this preserves old call sites without making
    the legacy name normative again;
  \item the non-conflicting edits in \texttt{ReadGmsh.hh} and \texttt{DoF.hh}
    were audited and accepted because they do not alter the multiscale
    semantics or the release gate.
\end{itemize}

\subsection{Merge-preserved dependency surface}

The purpose of the merge was not to maximize line retention; it was to preserve
the dependency surface that the stabilization cycle had already validated.  The
resulting local architecture around the merged headers is summarised in
Figure~\ref{fig:ms-stab-merge-class-diagram}.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners, align=center, minimum width=3.1cm, minimum height=1.0cm},
  arrow/.style={-{Latex[length=2.2mm]}, thick}
]
  \node[box] (analysis) {\texttt{Analysis.hh}\\aggregator surface};
  \node[box, below left=1.7cm and 2.0cm of analysis] (solver) {\texttt{DynamicAnalysis.hh}\\\texttt{NLAnalysis.hh}};
  \node[box, below right=1.7cm and 2.0cm of analysis] (director) {\texttt{AnalysisDirector.hh}\\\texttt{TransitionDirector.hh}};
  \node[box, below=3.5cm of analysis] (model) {\texttt{Model.hh}};
  \node[box, left=4.2cm of model] (material) {\texttt{MaterialPolicy.hh}\\\texttt{Strain.hh}\\\texttt{VoigtVector.hh}};
  \node[box, right=4.2cm of model] (mesh) {\texttt{ReadGmsh.hh}\\\texttt{GmshDomainBuilder.hh}\\\texttt{LagrangeElement.hh}};

  \draw[arrow] (analysis) -- (solver);
  \draw[arrow] (analysis) -- (director);
  \draw[arrow] (solver) -- (model);
  \draw[arrow] (director) -- (model);
  \draw[arrow] (material) -- (model);
  \draw[arrow] (mesh) -- (model);
\end{tikzpicture}
\caption{Merged header surface intentionally preserved after the stabilization audit.}
\label{fig:ms-stab-merge-class-diagram}
\end{figure}

\subsection{Decision}

The merge decision was therefore conditional rather than automatic: the branch
was considered mergeable with \texttt{origin/master} only after the stabilized
surface survived conflict resolution \emph{and} the regression gate was
revalidated.  In other words, the merge itself became part of the release
readiness proof.

\subsection{Post-merge gate evidence}

After the conflict audit, the local release gate was rerun through
\texttt{scripts/ci\_multiscale\_stabilization.ps1 -SkipConfigure}.  The
post-merge execution completed successfully and covered the full public release
surface already defined in this chapter:
\begin{itemize}
  \item regression executables:
    \texttt{fall\_n\_multiscale\_api\_test},
    \texttt{fall\_n\_micro\_solve\_executor\_test},
    \texttt{fall\_n\_evolver\_advanced\_test},
    \texttt{fall\_n\_beam\_test},
    \texttt{fall\_n\_steppable\_solver\_test}, and
    \texttt{fall\_n\_steppable\_dynamic\_test};
  \item heavy example builds:
    \texttt{fall\_n\_lshaped\_multiscale},
    \texttt{fall\_n\_lshaped\_multiscale\_16},
    \texttt{fall\_n\_table\_multiscale}, and
    \texttt{fall\_n\_table\_cyclic\_validation};
  \item document compilation:
    \texttt{latexmk -pdf}, producing the updated \texttt{main.pdf}.
\end{itemize}

The important technical result is not merely that the code compiled, but that
the merge did not alter the observed multiscale semantics:
\begin{enumerate}
  \item the FE\textsuperscript{2} API regressions still passed with zero
    failures;
  \item the advanced local-evolver matrix still passed after the merge,
    including the condensed-versus-forced-FD scientific comparison;
  \item the heavy multiscale examples still linked successfully from the merged
    header surface;
  \item the \LaTeX{} document still compiled, while preserving the explicit
    warning ledger for pre-existing bibliography/reference debt.
\end{enumerate}

\section{Follow-up: Cyclic Validation Extension and Build Audit}
\label{sec:ms-stab-cyclic-followup}

\subsection{Intent}

After the core multiscale stabilization and merge audit were completed, the
next objective was to stress the current implementation in a way that is both
scientifically revealing and operationally honest:
\begin{enumerate}
  \item extend the cyclic validation protocol so that cracking and hysteresis
    become visible at the table scale;
  \item improve observability of the FE\textsuperscript{2} cases without
    depending on debugger-only output;
  \item audit the true compile-time cost of the current validation executable
    rather than assuming that the earlier PCH/CI work had already solved it.
\end{enumerate}

\subsection{Architecture of the current validation driver}

Figure~\ref{fig:ms-stab-cyclic-driver} summarises the current
\texttt{main\_table\_cyclic\_validation.cpp} execution surface after the
follow-up changes.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners, align=center, minimum width=3.3cm, minimum height=1.0cm},
  arrow/.style={-{Latex[length=2.2mm]}, thick}
]
  \node[box] (cli) {\texttt{main\_table\_cyclic\_validation.cpp}\\CLI + protocol configuration};
  \node[box, below left=1.8cm and 2.3cm of cli] (case3) {Case 3\\\texttt{NonlinearAnalysis}\\beam + shell table};
  \node[box, below right=1.8cm and 2.3cm of cli] (fe2) {Cases 4/5\\\texttt{MultiscaleAnalysis}\\table + local sub-models};
  \node[box, below=1.8cm of case3] (rec3) {\texttt{global\_history.csv}\\fiber hysteresis\\VTM/PVD output};
  \node[box, below=1.8cm of fe2] (rec45) {\texttt{global\_history.csv}\\\texttt{crack\_evolution.csv}\\\texttt{solver\_diagnostics.csv}};
  \node[box, below=2.0cm of $(rec3)!0.5!(rec45)$] (plot) {\texttt{plot\_cyclic\_validation.py}\\comparisons + recorder plots};

  \draw[arrow] (cli) -- (case3);
  \draw[arrow] (cli) -- (fe2);
  \draw[arrow] (case3) -- (rec3);
  \draw[arrow] (fe2) -- (rec45);
  \draw[arrow] (rec3) -- (plot);
  \draw[arrow] (rec45) -- (plot);
\end{tikzpicture}
\caption{Current cyclic validation driver surface after the stabilization follow-up.}
\label{fig:ms-stab-cyclic-driver}
\end{figure}

The important architectural point is that the validation driver no longer only
emits a global hysteresis loop.  The FE\textsuperscript{2} cases now persist
step-level local response status, tangent scheme, condensed-operator status,
local SNES data, crack counts, and crack openings.  This matters because the
current scientific bottleneck is not a simple binary ``converged/failed''
question; it is a mixture of cost, cracking onset, and operator availability.

\subsection{Implemented changes}

The follow-up changes touched three areas:
\begin{itemize}
  \item \textbf{Protocol extension.}  The cyclic protocol utility now exposes
    both \texttt{legacy20} and \texttt{extended50}, with dedicated unit tests
    for each preset.
  \item \textbf{Driver observability.}  The table cyclic validation driver now
    writes \texttt{solver\_diagnostics.csv} for the FE\textsuperscript{2}
    cases, supports CLI overrides for local-ramp parameters and SNES tolerances,
    and exposes a manual FE\textsuperscript{2} step cap that preserves the
    protocol parametrization instead of rescaling it.
  \item \textbf{Plotting support.}  The plotting script was extended so that,
    when the CSV files are present, it can produce not only the classical
    hysteresis overlays but also recorder-based plots for global damage, crack
    evolution, and solver diagnostics.
\end{itemize}

\subsection{Observed numerical evidence}

Two short but decisive runs were used as the current evidence baseline.

\paragraph{Case 3, structural-only table, extended protocol.}
The structural table completed the full \texttt{extended50} protocol in the
audited coarse run (one solver step per segment).  The recorder values at the
positive peaks were:
\begin{center}
\begin{tabular}{rcc}
\toprule
Peak drift & Base shear (MN) & Peak structural damage \\
\midrule
\(+2.5\)\,mm & 0.5678 & 0.0217 \\
\(+5.0\)\,mm & 1.1355 & 0.0435 \\
\(+10.0\)\,mm & 2.2527 & 0.1236 \\
\(+20.0\)\,mm & 4.5003 & 0.2650--0.2777 \\
\(+35.0\)\,mm & 7.8742 & 0.4857 \\
\(+50.0\)\,mm & 11.2483 & 0.6934 \\
\bottomrule
\end{tabular}
\end{center}
The macro solver needed multiple bisections in the \(\pm 10\)\,mm cycle, so
the extended protocol is already non-trivial even before FE\textsuperscript{2}
feedback is introduced.

\paragraph{Case 4, one-way FE\textsuperscript{2}, first physical point.}
The first physical point of the same protocol, namely the step to
\(+2.5\)\,mm drift, required approximately 579~s in the audited environment.
The recorder values were:
\begin{itemize}
  \item macro base shear \(= -0.5678\)~MN;
  \item structural peak damage \(= 0.0217028\);
  \item total cracked Gauss points \(= 360\);
  \item total cracks \(= 1080\);
  \item peak crack opening \(= 9.75991\times 10^{-4}\).
\end{itemize}
The accompanying \texttt{solver\_diagnostics.csv} also showed that all four
sub-models:
\begin{itemize}
  \item finished \texttt{FirstSolveRamp} successfully;
  \item reported positive SNES convergence reasons;
  \item used \texttt{LinearizedCondensation};
  \item reported \texttt{CondensedTangentStatus::Success};
  \item did not require tangent regularization.
\end{itemize}

This observation is scientifically useful because it rules out a tempting but
incorrect diagnosis.  The one-way FE\textsuperscript{2} path is not currently
limited by immediate operator unavailability at the first cracked point.  It is
limited by \emph{cost} and by a \emph{response mismatch}: the continuum
sub-models accumulate many cracks at the same global drift level where the
structural-only model still reports relatively modest macro damage.

\subsection{Executor caveat}

One explicit caveat emerged from the audit and is now part of the documented
truth baseline: the cyclic FE\textsuperscript{2} driver must not currently be
run with \texttt{OpenMPExecutor}.  In the audited environment, that path caused
PETSc wrong-state failures associated with interleaved norm begin/end calls.
The code now therefore hardwires the cyclic FE\textsuperscript{2} validation
driver to \texttt{SerialExecutor}.  This is not a theoretical objection to
threaded micro execution in general; it is a concrete statement that this
specific PETSc-backed path is not yet validated for threaded execution.

\subsection{Compile-time evidence}

The compile-time audit also remained deliberately empirical.  After the earlier
PCH enablement and the move to parallel per-target builds in the CI script and
workflow, the heavy validation target still showed large rebuild costs in the
current environment:
\begin{itemize}
  \item rebuild of \texttt{fall\_n\_table\_cyclic\_validation} after the
    follow-up instrumentation changes: approximately 889~s;
  \item rebuild after the protocol-preserving FE\textsuperscript{2} step-cap
    correction: approximately 947~s.
\end{itemize}

This is important because it keeps the compile-time narrative honest.  PCH and
parallel target builds improved the gate and reduced redundant work, but they
did \emph{not} solve the underlying translation-unit size problem of
\texttt{main\_table\_cyclic\_validation.cpp}.  The next credible compile-time
improvement remains the one already identified during the audit: split this
driver into multiple translation units so that rebuilds of the FE\textsuperscript{2}
cases do not force recompilation of the entire validation suite.

\subsection{Second pass: recorder semantics, driver split, and adaptive continuation}

The next execution pass addressed three coupled problems that had become clear
only after the first exploratory FE\textsuperscript{2} cyclic runs:
\begin{enumerate}
  \item the cyclic recorders were semantically dishonest for constitutive laws
    that do not expose a scalar damage channel;
  \item the validation executable still hid too much compile-time cost behind a
    monolithic entrypoint;
  \item the first load reversal and the first \(+5\)\,mm excursion required
    adaptive continuation in the local sub-models, and the existing adaptive
    solver still contained an exact-budget classification bug.
\end{enumerate}

\paragraph{Recorder semantic correction.}
The audit of \texttt{KoBatheConcrete3D}, \texttt{Material.hh},
\texttt{FEM\_Element.hh}, and the cyclic FE\textsuperscript{2} driver confirmed
that the earlier recorder path silently mapped ``no scalar damage field
available'' to zero.  The stabilized execution now keeps that distinction
explicit:
\begin{equation}
  \texttt{damage\_scalar\_available} = 0
  \quad\Longrightarrow\quad
  \texttt{peak\_submodel\_damage\_scalar} = \texttt{NaN}.
\end{equation}
For the present cyclic FE\textsuperscript{2} path, the trustworthy fracture
channels are therefore the crack counts, crack opening envelope, and the
octahedral invariants \(\sigma_{o,\min}\) and \(\tau_{o,\max}\).  This matters
scientifically because it prevents a heavily cracked sub-model from being
misread as an undamaged one purely because the constitutive snapshot did not
expose a scalar damage index.

\paragraph{Driver split and measured compile-time improvement.}
The validation executable was then split into a thin CLI entrypoint
(\texttt{main\_table\_cyclic\_validation.cpp}), a lightweight public driver API
(\texttt{TableCyclicValidationAPI.hh}), a structural implementation TU
(\texttt{TableCyclicValidationStructural.cpp}), and the FE\textsuperscript{2}
implementation TU (\texttt{TableCyclicValidationFE2.cpp}).  The resulting
translation-unit structure is shown in Figure~\ref{fig:ms-stab-cyclic-split}.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners, align=center, minimum width=3.3cm, minimum height=0.95cm},
  arrow/.style={-{Latex[length=2.2mm]}, thick}
]
  \node[box] (main) {\texttt{main\_table\_cyclic\_validation.cpp}\\thin CLI entrypoint};
  \node[box, below=1.6cm of main] (api) {\texttt{TableCyclicValidationAPI.hh}\\lightweight driver surface};
  \node[box, below left=1.8cm and 2.6cm of api] (struct) {\texttt{TableCyclicValidationStructural.cpp}\\Cases 0--3};
  \node[box, below right=1.8cm and 2.6cm of api] (fe2) {\texttt{TableCyclicValidationFE2.cpp}\\Cases 4--5};
  \node[box, below=1.8cm of fe2] (support) {\texttt{TableCyclicValidationSupport.hh}\\heavy internal helpers};

  \draw[arrow] (main) -- (api);
  \draw[arrow] (api) -- (struct);
  \draw[arrow] (api) -- (fe2);
  \draw[arrow] (fe2) -- (support);
  \draw[arrow] (struct) -- (support);
\end{tikzpicture}
\caption{Translation-unit split introduced in the second cyclic-validation stabilization pass.}
\label{fig:ms-stab-cyclic-split}
\end{figure}

The compile-time audit was then repeated after regenerating the corrupted
Ninja dependency database (\texttt{build/.ninja\_deps}), because the broken
depfile state had been forcing spurious rebuilds of every object in the target.
With valid dependency tracking restored, the measured incremental rebuild costs
after the TU split alone were:
\begin{itemize}
  \item touching \texttt{main\_table\_cyclic\_validation.cpp}: 13.667~s;
  \item touching \texttt{TableCyclicValidationFE2.cpp}: 444.167~s;
  \item touching \texttt{TableCyclicValidationStructural.cpp}: 745.102~s.
\end{itemize}

A third pass then introduced a dedicated target-local precompiled header,
\texttt{TableCyclicValidationPch.hh}, instead of the earlier project-wide
\texttt{header\_files.hh} umbrella.  Applying that PCH to every translation
unit lowered the heavy incremental rebuilds to 338.699~s
(\texttt{FE2.cpp}) and 577.183~s (\texttt{Structural.cpp}), but it also
inflated the thin CLI rebuild to 50.365~s.  The final configuration therefore
keeps the dedicated PCH for the heavy validation implementation TUs while
marking \texttt{main\_table\_cyclic\_validation.cpp} with
\texttt{SKIP\_PRECOMPILE\_HEADERS}.  The resulting final measured costs are:
\begin{itemize}
  \item touching \texttt{main\_table\_cyclic\_validation.cpp}: 19.288~s;
  \item touching \texttt{TableCyclicValidationFE2.cpp}: 332.621~s;
  \item touching \texttt{TableCyclicValidationStructural.cpp}:
    577.183~s in the immediately preceding all-PCH measurement.
\end{itemize}
This is still expensive, but it is a materially better frontier than the
post-split/no-PCH state, and it preserves the fast feedback of the thin CLI
layer.

\paragraph{Adaptive continuation in the cyclic FE\textsuperscript{2} path.}
The first exploratory one-way FE\textsuperscript{2} run with an
8-substep \texttt{FirstSolveRamp} reached \(+2.5\)\,mm but still failed at the
first reversal \(-2.5\)\,mm.  The reason was not the initial ramp; it was the
fact that the subsequent solve path still attempted a full post-cracking step
without adaptive continuation.  The driver was therefore extended with
explicit controls for:
\begin{itemize}
  \item enabling adaptive sub-stepping from the first post-ramp step;
  \item choosing the arc-length activation threshold;
  \item choosing the adaptive substep and bisection budgets.
\end{itemize}
With adaptive continuation enabled from the start, the one-way cyclic
FE\textsuperscript{2} path completed the first reversal and the return to zero.
The \(-2.5\)\,mm reversal required exactly two adaptive substeps in every
sub-model; the zero-return step also required two adaptive substeps in every
sub-model.

\paragraph{Exact-budget bug and scientific consequence.}
The first attempt at the \(+5\)\,mm excursion then exposed a more subtle issue:
one sub-model reached \(\texttt{achieved\_fraction}=1.0\) at the same time it
hit the hardcoded 30-substep budget, but the solver still returned failure
because the budget check was executed after the successful update and did not
distinguish ``budget exhausted before completion'' from ``budget exhausted
exactly at completion''.  That was a real correctness defect, not merely a
tuning issue.

The fix had two parts:
\begin{enumerate}
  \item classify the adaptive step as failed only when the budget is exhausted
    \emph{and} the progress fraction remains \(< 1\);
  \item expose the adaptive substep/bisection budgets as explicit evolver and
    CLI configuration rather than hidden constants.
\end{enumerate}

This correction was locked by a dedicated regression test,
\texttt{test\_adaptive\_substepping\_accepts\_exact\_budget\_completion},
which constructs a case that converges in exactly two adaptive substeps and
verifies that the solver remains classified as converged.

\paragraph{Observed cyclic FE\textsuperscript{2} result after the fix.}
After the exact-budget bug was fixed, the same truncated
\texttt{extended50} one-way FE\textsuperscript{2} campaign completed the first
four protocol points:
\[
  +2.5\text{ mm},\quad -2.5\text{ mm},\quad 0,\quad +5.0\text{ mm}.
\]
The \(+5\)\,mm point required substantial local continuation:
the four sub-models completed \((21,23)\), \((26,28)\), \((30,32)\), and
\((15,15)\) adaptive substeps/bisections, respectively.  The resulting
global/crack recorders reported:
\begin{itemize}
  \item base shear \(V = -1.13555\)~MN;
  \item peak structural damage \(= 0.0434864\);
  \item 426 cracked Gauss points and 1268 crack planes;
  \item maximum crack opening \(= 1.82279\times 10^{-3}\);
  \item \(\sigma_{o,\min} = -2.23149\) and
    \(\tau_{o,\max} = 12.2076\).
\end{itemize}

Scientifically, this is a meaningful milestone: the current one-way
FE\textsuperscript{2} validation now exhibits both the first loop reversal and
the first higher-amplitude excursion with explicit crack-history observability.
The runtime is still very high, but the failure mode is no longer hidden and
the recorders no longer lie about the local damage semantics.

\paragraph{Execution profile and interruption-safe FE\textsuperscript{2} runs.}
The next stabilization pass focused on making the long
\texttt{extended50} cyclic FE\textsuperscript{2} campaign both cheaper and
more observable.  Three implementation changes were introduced together:
\begin{enumerate}
  \item a dedicated execution profile, \texttt{fe2\_crack50}, was added to the
    driver API/CLI to bundle the empirically useful settings for exploratory
    one-way FE\textsuperscript{2} runs
    (\(1\) step/segment, 8-substep ramp, adaptive continuation from the first
    reversal, enlarged adaptive budgets, and disabled VTK output);
  \item when sub-model VTK is disabled, \texttt{NonlinearSubModelEvolver} now
    caches only a \texttt{CrackSummary} instead of rebuilding the detailed
    crack-plane geometry that is needed solely for VTK export;
  \item the FE\textsuperscript{2} \texttt{hysteresis.csv} file is now streamed
    incrementally, step by step, in the same spirit as
    \texttt{global\_history.csv} and \texttt{solver\_diagnostics.csv}.
\end{enumerate}

This combination does not magically solve the runtime problem, but it changes
the failure semantics in an important way: partial long runs are now
diagnostically useful instead of losing the global hysteresis curve at the
point of interruption.

\paragraph{Observed frontier after the execution-profile update.}
A fresh one-way Case~4 run was launched with
\texttt{--protocol extended50 --fe2-profile crack50 --max-steps 5}.  The run
was allowed to execute for two hours in the audited environment.  It still did
not complete the \(-5\)\,mm reversal within that wall-clock budget, so the
runtime debt remains open.  However, the streamed recorders confirmed again
that the stabilized FE\textsuperscript{2} path completed the first four
protocol points through \(+5\)\,mm without losing any hysteresis or crack
history data.  The new summary-only plotting mode in
\texttt{scripts/plot\_cyclic\_validation.py} also confirmed, from the same
output directory, the broader status of the validation campaign:
\begin{itemize}
  \item Case~3 currently reaches the full \(50\)\,mm
    \texttt{extended50} envelope with structural damage
    \(= 0.69342\) and \(\max |V| = 11248.308\)~kN;
  \item Case~4 currently reaches \(+5\)\,mm with streamed FE\textsuperscript{2}
    hysteresis, \(\max |V| = 1135.55\)~kN, peak structural damage
    \(= 0.04349\), 1268 crack planes, and a maximum crack opening of
    \(1.8228\)~mm.
\end{itemize}
Scientifically, that is a better state than before this pass: the validation
campaign now has a clear full-amplitude structural reference and an honest
partial FE\textsuperscript{2} frontier, both reproducible from the same driver.

\section{Phase 5: Residual Debt Ledger}
\label{sec:ms-stab-phase5}

The stabilization cycle deliberately left some debt explicit rather than
masking it under a ``green'' release summary.  Table
\ref{tab:ms-stab-residual-debt} records the items that remain open after this
execution.

\begin{table}[htb]
\centering
\small
\begin{tabular}{p{0.28\textwidth}p{0.14\textwidth}p{0.40\textwidth}}
\toprule
Residual item & Severity & Current disposition \\
\midrule
Sparse condensed tangent still depends on explicit element-level local
linearisation support and keeps a finite-difference fallback &
medium/high &
Mitigated further in this cycle by extracting the sparse Schur solve into a
reusable workspace with symbolic-pattern caching; the fallback remains
necessary for unsupported local elements or singular free blocks. \\
Validation against forced adaptive finite differences shows a non-zero but
bounded tangent gap on mixed elastic states &
medium/high &
Documented explicitly through the new scientific regression; the current state
is honest and useful, but it is not yet a proof of exact algebraic identity
between the condensed operator and the perturbation fallback. \\
\texttt{NonlinearSubModelEvolver} still owns several responsibilities &
medium &
Improved in this cycle by extracting boundary-condition, state, and
homogenisation helpers; solver-core, diagnostics, and output concerns still
need separation. \\
Threaded cyclic FE\textsuperscript{2} validation path &
medium/high &
Documented explicitly as non-validated: the PETSc-backed cyclic driver must
currently use \texttt{SerialExecutor} because the threaded path triggered
wrong-state failures in the audited environment. \\
Compile time of \texttt{fall\_n\_table\_cyclic\_validation} &
medium/high &
Partially mitigated.  The final measured state uses a dedicated validation PCH
for the heavy TUs and skips it for the thin CLI entrypoint, yielding
approximately 19.3~s for \texttt{main\_table\_cyclic\_validation.cpp},
332.6~s for \texttt{TableCyclicValidationFE2.cpp}, and 577.2~s for the
structural TU in the audited environment. \\
One-way cyclic FE\textsuperscript{2} runtime beyond the first crack reversal &
medium/high &
Mitigated further by the \texttt{fe2\_crack50} execution profile, VTK-free
crack-summary caching, and streamed hysteresis output.  Even so, a two-hour
audited run still reached only the first four protocol points through
\(+5\)\,mm; the \(-5\)\,mm reversal remains beyond the current wall-clock
budget. \\
MPI-distributed micro execution &
medium &
Communicator ownership is explicit, but the distributed execution engine is not
yet implemented. \\
Type-erased local-model path &
medium &
Left available only as experimental code until erased checkpoint/restore is
completed. \\
Legacy documentation warnings unrelated to this stabilization &
low/medium &
Inventoried below; not silently fixed in this cycle because they pre-existed
and are broader than the multiscale module. \\
\bottomrule
\end{tabular}
\caption{Residual debt after the stabilization execution.}
\label{tab:ms-stab-residual-debt}
\end{table}

\section{Documentation Warning Ledger}
\label{sec:ms-stab-doc-ledger}

The PDF compilation succeeded during this cycle, but the log still reports
pre-existing warnings outside the immediate stabilization work:
\begin{itemize}
  \item undefined references;
  \item multiply-defined labels;
  \item several hyperref string warnings associated with older chapters.
\end{itemize}

These warnings were intentionally not hidden or silently normalised.  The
stabilization decision was to keep the document compiling, add the current
chapter as an execution record, and leave the broader document cleanup as a
separate documentation-quality task.

\section{Closure of the Stabilization Cycle}
\label{sec:ms-stab-close}

The multiscale stabilization cycle v1 closed the most important gap between
``public API redesign'' and ``release-worthy behaviour'':
\begin{enumerate}
  \item the local tangent path now exposes its own uncertainty instead of
    silently zeroing columns;
  \item incomplete local section operators are now classified explicitly as
    invalid rather than being treated as weak-but-usable responses;
  \item the FE\textsuperscript{2} protocol reports why it terminated and
    whether it rolled back;
  \item the physical path now has deterministic axial, flexural, shear, and
    torsional regression coverage;
  \item the release gate exists both as a local script and as a repository
    workflow;
  \item the debt that remains is explicit, bounded, and technically
    understandable.
\end{enumerate}
