% ==========================================================================
%  Chapter 90 -- FE^2 Staggered Chain Results (Plan v2 §Fase 4 + §5.3/5.4)
% ==========================================================================

\chapter{FE\(^2\) Staggered Chain Results}
\label{ch:fe2-staggered-chain-results}

\paragraph{Normative status.}
This chapter documents the four-stage FE\(^2\) chain frozen in
\href{run:src/validation/ReducedRCMultiscaleValidationStartCatalog.hh}{%
\texttt{ReducedRCMultiscaleValidationStartCatalog.hh}}:
\(\texttt{one\_way\_replay} \to \texttt{local\_site\_batch} \to
\texttt{elastic\_fe2\_smoke} \to \texttt{enriched\_fe2\_guarded\_smoke}\)
and its multi-site extensions (Plan~v2 \S5.3, \S5.4).

\section{Four-stage chain (\S Fase 4)}

Each stage is materialised as a standalone Eigen-only driver that
consumes a structural cyclic CSV (e.g. \path{hysteresis.csv} produced by
\texttt{main\_reduced\_rc\_xfem\_reference\_benchmark}) and emits a
versioned JSON manifest. The drivers are:

\begin{itemize}
  \item \texttt{main\_multiscale\_one\_way\_replay} -- 4A planner;
        emits \texttt{multiscale\_replay\_plan.json} and
        \texttt{multiscale\_runtime\_policy.json}.
  \item \texttt{main\_local\_site\_batch} -- 4B batch plan;
        emits \texttt{multiscale\_local\_site\_batch\_plan.json}.
  \item \texttt{main\_elastic\_fe2\_smoke} -- 4C elastic plumbing;
        synthesises a per-site \texttt{UpscalingResult} with diagonal
        \(D_{\mathrm{hom}}=\operatorname{diag}(\mathrm{EA},\mathrm{EI})\)
        and asserts the Cap.~79 (0.03,\,6) gate.
  \item \texttt{main\_xfem\_fe2\_guarded\_one\_step} -- 4D enriched
        guarded one-step; runs the
        \texttt{EnrichmentActivationProbe} per site and emits a
        damage-degraded \texttt{UpscalingResult} with
        \(\|R\|_F = 0.005 + 0.020\,d\).
\end{itemize}

The shared CSV reader and multi-site demand-decay generator live in
\href{run:src/validation/MultiscaleReplayDriverHelpers.hh}{%
\texttt{MultiscaleReplayDriverHelpers.hh}}. End-to-end on the canonical
200~mm small-strain cyclic CSV with \texttt{--num-sites 3}: 4A
\texttt{selected=3}, 4B \texttt{batches=3}, 4C \texttt{overall\_pass=1},
4D \texttt{overall\_pass=1, max\_guarded\_frob<0.030}.

\section{Multi-site replay (\S5.3)}

The 4A driver accepts \texttt{--num-sites N},
\texttt{--site-z-list}, and \texttt{--site-scale-list}. The default
ladder \(z/L = (0.02, 0.10, 0.30, 0.55, 0.85)\) with demand scales
\((1.00, 0.70, 0.40, 0.20, 0.10)\) reproduces a cantilever-style
column. Drivers 4B/4C/4D inherit the same CLI.

The dedicated regression test
\href{run:tests/test_multiscale_fase5_multi_site_chain_smoke.cpp}{%
\texttt{test\_multiscale\_fase5\_multi\_site\_chain\_smoke.cpp}}
walks the full 4A\(\to\)4D chain in memory for three cantilever sites
and asserts: candidate=3, selected=3, monotonically decreasing
activation score, activated\_count=3, passed\_gate\_count=3,
\(\max\|R\|_F<0.030\).

\section{Cyclic two-way large-amplitude smoke (\S5.4)}

A 5-cycle ramped \(50\to300\)~mm protocol (144 samples) is replayed
across the same three-site ladder. The smoke test
\href{run:tests/test_multiscale_fase5_cyclic_two_way_smoke.cpp}{%
\texttt{test\_multiscale\_fase5\_cyclic\_two\_way\_smoke.cpp}} asserts
that damage and Frobenius residual grow monotonically across amplitude
prefixes \(\{24,48,72,96,120,144\}\) and that the Cap.~79 gate is
honoured at every prefix.

\paragraph{Honest scientific status.} \texttt{synthetic\_chain\_no\_real\_xfem\_local\_solve}.
The four standalone drivers exercise the FE\(^2\) plumbing with a
deterministic damage-driven proxy for the local solve, not the real
XFEM small-strain micro-step. The latter remains the responsibility of
\texttt{main\_reduced\_rc\_xfem\_reference\_benchmark} (Cap.~89) and
sits behind the heavy reference benchmark gate.

\section{Schur-vs-bordered diagnostic (\S4-bis)}

The diagnostic comparator
\href{run:src/numerics/SparseSchurComplement.hh}{%
\texttt{SparseSchurComplement.hh}} is exercised in
\texttt{test\_schur\_vs\_bordered\_dhom} (analytical pre-fissure case)
to verify that the Schur condensation and the bordered mixed-control
return identical \(D_{\mathrm{hom}}\) to within 1\%. Post-fissure
discrepancy is documented but not gated (Schur is diagnostic only).
