% ==========================================================================
%  Chapter 91 -- Cross-Model Equivalence Matrix (Plan v2 §Fase 5.1/5.2)
% ==========================================================================

\chapter{Cross-Model Equivalence Matrix}
\label{ch:cross-model-equivalence-matrix}

\paragraph{Normative status.}
This chapter records the cross-model identification program between
the reduced-RC structural column (beam fibre with Lobatto/GL/GR
quadrature) and the continuum / XFEM column references. It uses the
frozen matrix
\(
  \mathfrak{B}_{\text{phase3}} = N\times\{\mathrm{GL},\mathrm{Lobatto},
  \mathrm{GR}_{\text{left}},\mathrm{GR}_{\text{right}}\}
\)
declared in
\href{run:src/validation/ReducedRCColumnStructuralMatrixCatalog.hh}{%
\texttt{ReducedRCColumnStructuralMatrixCatalog.hh}}.

\section{Closure observable}

The \(g^{\star}\) observable (nearest active fixed-side station, see
Cap.~88) and the moment--curvature norm \(\eta_{M}\) defined in
\href{run:src/validation/ReducedRCColumnMomentCurvatureClosure.hh}{%
\texttt{ReducedRCColumnMomentCurvatureClosure.hh}} are the canonical
discrepancy metrics. The frozen reference cell \((N=4,\,\mathrm{Lobatto})\)
satisfies \(\eta_{M} \lesssim 1.8\times 10^{-3}\).

\section{Status (Plan v2 \S5.1, \S5.2)}

\textbf{Scoped-deferred} pending \texttt{LIBS FULL} integration. The
36-cell sweep is gated on the heavy
\texttt{main\_reduced\_rc\_xfem\_reference\_benchmark} which depends on
PETSc/SNES + reduced-RC section solver. The audit trail for this
deferral lives in
\path{data/output/validation_reboot/audit_phase5_plan_v2.json}
(\(\S5.3,\,\S5.4\) closed; \(\S5.1,\,\S5.2\) scoped-deferred).

\section{Reduced-driven equivalence already in place}

Independently of the heavy 36-cell matrix, the FE\(^2\) chain
(Cap.~90) constitutes a constructive cross-model bridge:

\begin{itemize}
  \item The 4C elastic smoke verifies that the homogenised
        \(D_{\mathrm{hom}}\) returned by the bordered mixed-control
        condensation matches the analytical sectional values
        \(\mathrm{EA}, \mathrm{EI}\) of the beam fibre to \(\sim 0.5\%\).
  \item The 4D guarded smoke verifies that, at the activation step,
        the enriched local response degrades \(D_{\mathrm{hom}}\)
        consistently with the structural damage indicator carried by
        the macro section.
\end{itemize}

The heavy 36-cell M--\(\phi\) sweep, when it lands in a dedicated
research branch, is expected to confirm the same behaviour over the
full \((N,\,q)\) grid. Its deferral does not block the FE\(^2\)
plumbing nor the staging 16-storey driver (Cap.~92).

\paragraph{Honest scientific status.}
\texttt{scoped\_deferred\_to\_branch} for the 36-cell matrix;
\texttt{constructive\_chain\_evidence} for the FE\(^2\) plumbing.
