% ==========================================================================
%  Chapter 93 -- Ko-Bathe Analytical Tangent (Plan v2 §Fase 8, optional)
% ==========================================================================

\chapter{Ko--Bathe Analytical Tangent (Closed with Runtime Evidence)}
\label{ch:ko-bathe-analytical-tangent}

\paragraph{Normative status.}
This chapter was originally declared an \emph{optional,
scoped-deferred growth path} by Plan~v2 \S Fase~8 (reachable only
if the XFEM-led closure of Cap.~89 failed). The
\texttt{KoBatheConcrete} accessor \texttt{tangent(eps)} was already
implemented as a closed-form 2D plane-stress derivative; what was
missing was an explicit runtime check that the accessor truly
\emph{is} an analytical derivative of \texttt{compute\_response(eps)}
rather than a secant approximation. That gap is now closed by
ctest~\#25 \texttt{ko\_bathe\_analytical\_tangent}
(\path{tests/test_ko_bathe_analytical_tangent.cpp}). Status:
\texttt{closed\_with\_runtime\_evidence}.

\section{Verification procedure}

For each strain state $\boldsymbol{\varepsilon}_0$ on a
representative 2D plane-stress grid the test computes
\[
  \mathbf{C}_{\text{FD}}\bigl(\boldsymbol{\varepsilon}_0\bigr)_{:j}
  = \frac{\boldsymbol{\sigma}\bigl(\boldsymbol{\varepsilon}_0+h\,\mathbf{e}_j\bigr)
        - \boldsymbol{\sigma}\bigl(\boldsymbol{\varepsilon}_0-h\,\mathbf{e}_j\bigr)}{2h},
  \qquad j\in\{1,2,3\},
\]
from \emph{independent copies} of the model (no committed-state
drift), then compares against the analytical
$\mathbf{C}(\boldsymbol{\varepsilon}_0)
  = \texttt{KoBatheConcrete::tangent}(\boldsymbol{\varepsilon}_0)$
in Frobenius-relative norm:
\[
  \epsilon^{\text{rel}}_F
   = \frac{\bigl\lVert\mathbf{C}-\mathbf{C}_{\text{FD}}\bigr\rVert_F}
          {\max\!\bigl(\bigl\lVert\mathbf{C}_{\text{FD}}\bigr\rVert_F,\,\epsilon_{\text{floor}}\bigr)}.
\]
Closure gates: $\epsilon^{\text{rel}}_F\le 5\%$ on the elastic
grid ($f_c=30$~MPa, $|\boldsymbol{\varepsilon}|\le 5\!\times\!10^{-5}$,
$h=10^{-9}$) and $\epsilon^{\text{rel}}_F\le 10\%$ on the
entering-nonlinear grid ($f_c=21$~MPa,
$|\boldsymbol{\varepsilon}|\le 3\!\times\!10^{-4}$, $h=10^{-8}$).
Both gates pass with $\ge 90\%$ of grid points within tolerance and
no blow-up (i.e.\ no point exhibits a runaway secant).

\section{Original deferral rationale (historical)}

The Fase~8 work was \emph{not} blocking because:

\begin{itemize}
  \item Cap.~89 closes the XFEM small-strain gate at
        \(\mathrm{NZ}\in\{4,5,6,8\}\) up to 300~mm cyclic amplitude.
  \item Cap.~89 closes the corotational gate at 200/300~mm.
  \item TL/UL residuals are recorded with a documented narrow
        peak-ratio overshoot at 200~mm and a larger overshoot at
        300~mm; this is honest research evidence, not a blocker for
        the FE\(^2\) chain (Cap.~90) which uses small-strain.
\end{itemize}

Ko--Bathe is still \texttt{research\_guardrail\_not\_promoted} per
\href{run:src/validation/ReducedRCLocalModelPromotionCatalog.hh}{%
\texttt{ReducedRCLocalModelPromotionCatalog.hh}}; the
analytical-tangent runtime check verifies its
\emph{constitutive-level} consistency, not its multiscale
promotion.

\section{Scope (when fully activated)}

\begin{enumerate}
  \item The closed-form 2D plane-stress tangent
        \(\partial \boldsymbol{\sigma}/\partial \boldsymbol{\varepsilon}\)
        is now verified by ctest~\#25.
  \item Validation against the experimental dataset
        \texttt{ko\_bathe\_3d\_uniaxial\_fc\{21,30\}.csv} is
        covered by \texttt{ko\_bathe\_concrete} and
        \texttt{ko\_bathe\_concrete\_3d}.
  \item Promotion of Ko--Bathe in the catalog from
        \texttt{research\_guardrail\_not\_promoted} to
        \texttt{promoted\_control} remains contingent on a future
        Phase~6 dynamic claim and is intentionally not made here.
\end{enumerate}

The audit JSON
\path{data/output/validation_reboot/audit_phase8_plan_v2.json}
records the present
\texttt{closed\_with\_runtime\_evidence} status and the residual
activation conditions for catalog promotion.

\paragraph{Honest scientific status.}
\texttt{closed\_with\_runtime\_evidence} (constitutive-level);
multiscale promotion of Ko--Bathe is intentionally still
\texttt{research\_guardrail\_not\_promoted}.
