% ==========================================================================
%  Chapter 94 -- RVE Gmsh Tet Support (Plan v2 §Fase 9, optional)
% ==========================================================================

\chapter{RVE Gmsh Tetrahedral Support (Deferred Growth Path)}
\label{ch:rve-gmsh-tet-support}

\paragraph{Normative status.}
This chapter is the second of the two \emph{optional, scoped-deferred
growth paths} of Plan~v2 (\S Fase~9). It is reachable only if Phase~6
shows that prismatic \(\mathrm{NZ}=4\) RVEs are insufficient for
non-rectangular columns (beams, transitions). It is currently
\texttt{scoped\_deferred}.

\section{Decision condition}

The current XFEM site geometry uses
\href{run:src/model/PrismaticDomainBuilder.hh}{%
\texttt{PrismaticDomainBuilder}} with hex27 elements at
\(\mathrm{NZ}=4\) (Cap.~89 baseline gate-passed). Ch.~80 documents
the embedded rebar penalty coupling that makes this geometry
sufficient for the L-shape rectangular columns of the 16-storey
benchmark (Cap.~92).

Fase~9 would be activated by a heavy Phase~6 run that exhibits
mesh-induced stress concentrations or cracking patterns clearly
incompatible with the prismatic abstraction -- e.g. RC beams,
T- or L-section transitions, or in-fill walls.

\section{Scope (when activated)}

\begin{enumerate}
  \item Add a Gmsh tet builder that emits the same RVE contract
        consumed by \texttt{LocalContinuumEvolver} /
        \texttt{LocalXFEMEvolver}.
  \item Reuse the BC architecture (PETSc Section elimination +
        \texttt{IncrementalControlPolicy}, Cap.~86) with the
        generalised \texttt{LocalBoundaryConditionApplicator}
        face-selection callback (Plan~v2 \S2.3).
  \item Re-run the 4D guarded smoke and the multi-site chain
        (Cap.~90) on the tet RVEs and confirm parity with the hex
        baseline within the Cap.~79 gate.
\end{enumerate}

The audit JSON
\path{data/output/validation_reboot/audit_phase9_plan_v2.json}
records the present \texttt{scoped\_deferred} status and the
activation condition.

\paragraph{Honest scientific status.} \texttt{scoped\_deferred\_growth\_path}.
