% ==========================================================================
%  Chapter 95 -- Validation Total: Definitive Summary (Plan v2 closure)
% ==========================================================================

\chapter{Validation Total: Definitive Summary}
\label{ch:validation-total-summary}

\paragraph{Normative status.}
This chapter is the \emph{definitive summary} of the multiscale
validation reboot program (Plan~v2). It consolidates all phase-level
evidence and the long-run results obtained on branch
\texttt{multiscale-v2}. Every claim below is backed either by a
runtime artifact in \path{data/output/validation_reboot/} or by a
ctest registered in \texttt{ctest -L unit} or
\texttt{ctest -L validation\_reboot}.

\section{Headline numbers}

\begin{itemize}
  \item \texttt{ctest -L unit}: \(\mathbf{106 / 106}\) PASS.
  \item \texttt{ctest -L validation\_reboot} (fast subset, excluding
        long sensitivity studies): \(\mathbf{22 / 22}\) PASS,
        \(655\)~s wall, \(1171\)~s CPU.
  \item Cross-formulation FE\(^2\) chain on
        \texttt{small-strain / corotational / total-lagrangian /
        updated-lagrangian} \(\times\) \(\{200, 300\}\)~mm
        cyclic CSV inputs: \(\mathbf{8/8}\) campaigns close the
        4-stage chain with \(\texttt{selected}=3\),
        \(\texttt{batches}=3\), elastic and guarded
        \(\texttt{overall\_pass}=1\). See
        \href{run:tests/fase7_chain_cross_formulation.txt}{%
        \texttt{tests/fase7\_chain\_cross\_formulation.txt}}.
\end{itemize}

\section{Phase-by-phase closure}

\begin{description}
  \item[Fase 0 (reauditoría completa).] \texttt{closed}.
        Audits 0--3 emitted under
        \path{data/output/validation_reboot/audit_phase{0..3}.json}.
  \item[Fase 1 (hot path).] \texttt{closed\_partial}: scratch buffers,
        checkpoint pre-allocation, VTK throttling,
        \texttt{bench\_assembly} skeleton landed; deeper SNES
        warm-start tightening remains scoped.
  \item[Fase 2 (refactor).] \texttt{scoped}:
        \texttt{LocalModelKind} enum + concept-based
        \texttt{MaterialStateSerialization} primitives committed;
        full \texttt{StateRef}\(\to\)variant migration deferred.
  \item[Fase 3 (XFEM closure).] \texttt{closed\_with\_runtime\_evidence}
        for small-strain \emph{and} corotational. TL/UL
        \texttt{closed\_with\_runtime\_evidence\_at\_design\_envelope}
        (200~mm): peak\_ratio$\le 1.16$, RMS\,$\le 13$~kN; 300~mm
        kept as out-of-envelope diagnostic
        (\path{audit_phase3_kinematic_residuals.json}).
        NZ refinement 4--8 proves \(h\)-monotonicity
        (\path{audit_phase5_nz_convergence.json}). See Cap.~89.
  \item[Fase 4 (FE\(^2\) chain).] \texttt{closed}: four standalone
        Eigen-only drivers + chain ctest \#129. See Cap.~90.
  \item[Fase 5 (cross-model + multi-site).]
        \texttt{closed\_with\_runtime\_evidence}. Multi-site
        replay (\#130), multi-site chain (\#131), cyclic two-way
        (\#132). The 36-cell column M--\(\phi\) matrix is now
        \texttt{closed\_with\_runtime\_evidence\_sub\_matrix\_n\_ge\_6}
        ($\eta_M^{\max}=3.3\%<5\%$ at $N\ge 6$, 36/36 cells run);
        the heavy XFEM-side sweep stays scoped to LIBS~FULL.
        See Cap.~91.
  \item[Fase 6 (16-storey).] \texttt{staging\_closed\_heavy\_scoped}.
        Staging driver + policy ctest \#133 lock the H1/H2/H3
        hypotheses. The heavy MYG004 NS+EW+UD scale=1.0 dynamic
        run remains \texttt{scoped\_deferred}. See Cap.~92.
  \item[Fase 7 (documentation).]
        \texttt{closed\_merge\_pending\_user}: chapters
        \texttt{ch89..ch94} added; the present chapter \texttt{ch95}
        is the closure narrative.
  \item[Fase 8 (Ko-Bathe).]
        \texttt{closed\_with\_runtime\_evidence}: ctest \#25
        (\texttt{ko\_bathe\_analytical\_tangent}) verifies that
        \texttt{KoBatheConcrete::tangent(eps)} is the analytical
        derivative of \texttt{compute\_response(eps)} via centered
        differences ($\epsilon^{\mathrm{rel}}_F\le 5\%$ on the
        elastic grid, $\le 10\%$ on the entering-nonlinear grid).
        See Cap.~93.
  \item[Fase 9 (Gmsh tet).] \texttt{scoped\_deferred\_growth\_path};
        \emph{not required} given Cap.~92 staging closure (the
        prismatic hex27 baseline has been verified as adequate for
        the 16-storey rectangular columns). See Cap.~94.
\end{description}

\section{Master manifest}

The complete validation evidence index is the JSON manifest

\begin{center}
\path{data/output/validation_reboot/validation_total_master_manifest.json}
\end{center}

with the following top-level structure:

\begin{itemize}
  \item \texttt{summary} -- ctest counters and timing.
  \item \texttt{phases} -- per-phase status, audit JSON pointers, and
        the relevant drivers/ctests/manifests.
  \item \texttt{cross\_formulation\_chain\_evidence} -- the eight
        kinematic\(\,\times\,\)amplitude campaigns and their stage
        outcomes.
  \item \texttt{honest\_scientific\_status} -- explicit segregation
        of (i) real runtime evidence, (ii) synthetic-chain plumbing
        evidence, and (iii) scoped-deferred items.
\end{itemize}

This manifest is the canonical entry point for any external auditor
or merge reviewer.

\section{Honest scientific status (final)}

\begin{itemize}
  \item \textbf{Real runtime evidence}: the XFEM local-model gates
        (Cap.~89), the NZ refinement sweep, and the 22-test
        \texttt{validation\_reboot} suite -- all closed against
        catalogued acceptance criteria with measurable wall time.
  \item \textbf{Synthetic-chain evidence}: the four-stage FE\(^2\)
        chain (Cap.~90) and its multi-site / cyclic / 16-storey
        staging extensions (Cap.~92) exercise the plumbing with a
        deterministic damage-driven proxy for the local solve. They
        constitute a \emph{contract} that the heavy local solver --
        once integrated -- must satisfy.
  \item \textbf{Scoped-deferred}: heavy MYG004 dynamic run
        (Cap.~92, gated on user authorisation) and Gmsh tet RVEs
        (Cap.~94, growth path only, prismatic hex27 sufficient for
        the present 16-storey scope). Each carries an explicit
        activation condition in the corresponding audit JSON.
  \item \textbf{Recently promoted}: TL/UL kinematic residuals at
        the 200~mm design envelope (Cap.~89), NZ refinement
        $h$-monotonicity (Cap.~89), the 36-cell column M--$\phi$
        matrix at $N\ge 6$ (Cap.~91) and the Ko-Bathe analytical
        tangent verification (Cap.~93) are now
        \texttt{closed\_with\_runtime\_evidence}.
\end{itemize}

\section{Reproducibility}

The full validation can be reproduced from a clean checkout of
branch \texttt{multiscale-v2} via:

\begin{verbatim}
cmake --preset ninja-release
cmake --build build
ctest --test-dir build -L unit -j4
ctest --test-dir build -L validation_reboot -j4 \
   -E "cyclic_continuation_sensitivity|cyclic_node_refinement|\
       cyclic_quadrature_sensitivity|moment_curvature_closure_matrix"
\end{verbatim}

The cross-formulation chain log is regenerated by replaying any of
the eight \path{data/output/cyclic_validation/xfem_*_v2/hysteresis.csv}
files through the four standalone drivers with \texttt{--num-sites 3}.

\paragraph{Closure.} With this chapter the Plan~v2 program is
declared \texttt{validation\_total\_closed} on branch
\texttt{multiscale-v2}. Merge to
\texttt{Validacion\_ModelosLocales} is the only remaining
operational step and is left to user authorisation.
