% ==========================================================================
%  Chapter 96 -- Stage A: One-Way FE^2 Column (cyclic 200 mm)
% ==========================================================================

\chapter{Stage~A: One-Way FE\textsuperscript{2} on the Cyclic 200~mm RC Column}
\label{ch:fe2-column-one-way}

\paragraph{Purpose.}
This chapter closes the first of three sequential stages that validate
the FE\textsuperscript{2} multiscale coupling on the same RC column already
validated in mono-scale (Cap.~89, small-strain $200/300$~mm). Stage~A
exercises the \emph{one-way} chain: macro Timoshenko-fibre column $\to$
synthesised local response, with \emph{no} feedback into the macro
tangent. Stage~B (Cap.~97) then enables two-way feedback. Only when both
close, Stage~C (Cap.~98) escalates to the full dynamic 16-storey case.

\section{Setup}

\begin{itemize}
  \item Macro element (proxy): \texttt{TimoshenkoBeam3D} fibre, $N=4$
        per column, section $200\times200$~mm; replayed via the
        Cap.~89 cyclic envelopes.
  \item Cyclic input: any of the eight canonical CSVs
        \path{data/output/cyclic_validation/xfem_*_v2/global_xfem_newton_progress.csv}.
  \item Activation criterion (injected): \texttt{FirstInelasticFiberCriterion}
        in \path{src/analysis/FirstInelasticFiberCriterion.hh}, defined as
        \[
          \texttt{trigger}\;=\;
          (|\sigma_{\text{rebar}}|\ge f_y) \;\lor\;
          (|\kappa|\,c_{\text{section}} \ge \varepsilon_y) \;\lor\;
          (d \ge d_{\text{floor}}).
        \]
  \item Local upscaling: damage-degraded diagonal section tangent
        $D_{\text{hom}} = (1-d)\,\mathrm{diag}(EA, EI)$, identical
        construction to the Fase~4D guarded smoke driver, ensuring
        cross-stage comparability.
\end{itemize}

\section{Driver and tests}

\begin{itemize}
  \item Driver: \path{main_fe2_column_cyclic_one_way.cpp} (Eigen-only,
        no PETSc, no VTK).
  \item Unit test of the criterion: \texttt{first\_inelastic\_fiber\_criterion}
        (\path{tests/test_first_inelastic_fiber_criterion.cpp}).
  \item Smoke test: \texttt{fe2\_column\_cyclic\_one\_way\_smoke}
        (\path{tests/test_fe2_column_cyclic_one_way_smoke.cpp}).
  \item Both ctests label \texttt{unit}; PASS in $<1$~s wall.
\end{itemize}

\section{Cross-formulation evidence}

The driver was run with \texttt{--num-sites 3} on every kinematic
formulation closed in Cap.~89:

\begin{center}
\begin{tabular}{lll}
\hline
Campaign & Selected & Overall pass \\
\hline
\texttt{xfem\_small\_strain\_200mm\_v2}       & 3 & true \\
\texttt{xfem\_corotational\_200mm\_v2}        & 3 & true \\
\texttt{xfem\_total\_lagrangian\_200mm\_v2}   & 3 & true \\
\texttt{xfem\_updated\_lagrangian\_200mm\_v2} & 3 & true \\
\texttt{xfem\_small\_strain\_300mm\_v2}       & 3 & true \\
\texttt{xfem\_corotational\_300mm\_v2}        & 3 & true \\
\texttt{xfem\_total\_lagrangian\_300mm\_v2}   & 3 & true \\
\texttt{xfem\_updated\_lagrangian\_300mm\_v2} & 3 & true \\
\hline
\end{tabular}
\end{center}

\noindent
The criterion fires within the first cycle (drift $\approx 6$~mm,
\texttt{trigger\_reason = rebar\_yield}) for every campaign, confirming
that the injected predicate is consistent with the rebar-yield surface
already exposed by the Cap.~89 envelopes.

\section{Honest scientific status}

\texttt{synthetic\_one\_way\_real\_macro\_history}: the macro CSVs are
real Cap.~89 evidence, while the local upscaling is the damage-degraded
diagonal surrogate inherited from Fase~4D. The moment-envelope gate is
marked N/A on these inputs because the canonical CSV does not record
the section moment column; the gate becomes active when the driver is
fed an extended history that includes \texttt{moment\_y\_MN\_m}. The
real one-way bridge with the heavy XFEM local solve lives in
\path{main_lshaped_multiscale.cpp} (LIBS~FULL).

\section{Audit JSON and manifest}

\path{data/output/validation_reboot/audit_stage_a_fe2_column_one_way.json}
captures the per-campaign pass status, criterion knobs and honest
caveats. It is the canonical entry point for Stage~B reviewers.
