% ==========================================================================
%  Chapter 97 -- Stage B: Two-Way FE^2 Column (cyclic 200 mm)
% ==========================================================================

\chapter{Stage~B: Two-Way FE\textsuperscript{2} on the Cyclic 200~mm RC Column}
\label{ch:fe2-column-two-way}

\paragraph{Purpose.}
Stage~B keeps the same column, the same cyclic envelope and the same
\texttt{FirstInelasticFiberCriterion} as Stage~A (Cap.~96) but enables
\emph{two-way} staggered feedback between the macro Timoshenko-fibre
section and the local condensed tangent. The chapter documents the
synthetic-chain closure and the prerequisites left for the LIBS~FULL
real-solver path.

\section{Driver and staggered loop}

\begin{itemize}
  \item Driver: \path{main_fe2_column_cyclic_two_way.cpp}
        (Eigen-only).
  \item Staggered iteration parameters (defaults aligned with
        \path{main_lshaped_multiscale.cpp}):
        \begin{itemize}
          \item \texttt{MAX\_STAGGERED\_ITER}~$= 4$
          \item \texttt{STAGGERED\_TOL}~$= 0.05$
          \item \texttt{STAGGERED\_RELAX}~$= 0.7$
        \end{itemize}
  \item Convergence-rate gate over selected sites:
        \texttt{min\_converged\_fraction}~$\ge 0.95$.
  \item Smoke test: \texttt{fe2\_column\_cyclic\_two\_way\_smoke}
        (\path{tests/test_fe2_column_cyclic_two_way_smoke.cpp}); PASS
        in $<0.1$~s wall.
\end{itemize}

\section{Cross-formulation evidence}

\begin{center}
\begin{tabular}{llll}
\hline
Campaign & Selected & Triggered & Converged \\
\hline
\texttt{xfem\_small\_strain\_200mm\_v2}       & 3 & 3 & 3 \\
\texttt{xfem\_corotational\_200mm\_v2}        & 3 & 3 & 3 \\
\texttt{xfem\_total\_lagrangian\_200mm\_v2}   & 3 & 3 & 3 \\
\texttt{xfem\_updated\_lagrangian\_200mm\_v2} & 3 & 3 & 3 \\
\texttt{xfem\_small\_strain\_300mm\_v2}       & 3 & 3 & 3 \\
\texttt{xfem\_corotational\_300mm\_v2}        & 3 & 3 & 3 \\
\texttt{xfem\_total\_lagrangian\_300mm\_v2}   & 3 & 3 & 3 \\
\texttt{xfem\_updated\_lagrangian\_300mm\_v2} & 3 & 3 & 3 \\
\hline
\end{tabular}
\end{center}

\noindent
$\texttt{converged\_fraction} = 1.00$ in all eight campaigns, comfortably
above the $0.95$ gate.

\section{Honest scientific status}

\texttt{synthetic\_two\_way\_real\_macro\_history}: the staggered loop
relaxes between the macro $\mathrm{diag}(EA, EI)$ tangent and a
damage-degraded local target with the same scaling as Stage~A. It
exercises the convergence/relaxation plumbing but is \emph{not}
equivalent to the real bordered-mixed-control bridge, which lives in
\path{main_lshaped_multiscale.cpp} (LIBS~FULL).

\section{Audit JSON and prerequisites for Stage~C}

\path{data/output/validation_reboot/audit_stage_b_fe2_column_two_way.json}
captures the per-campaign trigger / convergence counts and the
prerequisites for Stage~C:

\begin{itemize}
  \item \texttt{lshaped\_multiscale\_16storey\_staging\_policy\_smoke}
        (ctest \#133) remains green on the present branch.
  \item The \texttt{FirstInelasticFiberCriterion} is available and ready
        to be substituted for the synthetic damage-only trigger in
        \path{main_lshaped_multiscale_16storey.cpp} when the heavy
        MYG004 dynamic run is authorised.
\end{itemize}

\section{Closure}

With Stage~A and Stage~B closed, the multiscale program owns a
runtime-evidenced one-way \emph{and} two-way chain on the 200~mm
cyclic column for every kinematic formulation closed in Cap.~89. The
remaining step is the dynamic 16-storey escalation (Cap.~98), which is
gated on user authorisation due to compute cost.
