% ==========================================================================
%  Chapter 98 -- Stage C: 16-Storey Dynamic FE^2 (scoped/deferred)
% ==========================================================================

\chapter{Stage~C: 16-Storey Dynamic FE\textsuperscript{2} (Scoped / Deferred)}
\label{ch:fe2-16storey-dynamic}

\paragraph{Purpose.}
Stage~C is the full dynamic escalation of the FE\textsuperscript{2}
chain validated in Stages~A (Cap.~96) and~B (Cap.~97): the 16-storey
L-shaped frame under the MYG004 record, with two-way staggered
multiscale on the columns and the \texttt{FirstInelasticFiberCriterion}
deciding when each replay site escalates to the LIBS~FULL local solve.

\section{Honest scientific status}

\texttt{scoped\_deferred} (heavy compute). The infrastructure is in
place and verified at smoke level; the heavy run is held back until the
user authorises the compute hours.

\section{Verified prerequisites (this branch)}

\begin{itemize}
  \item \path{main_lshaped_multiscale_16storey.cpp} compiles in the
        \texttt{LIBS FULL} tier.
  \item ctest \#133
        \texttt{fall\_n\_lshaped\_multiscale\_16storey\_staging\_policy\_smoke}
        is green and gates the staging policy at unit-test speed.
  \item Stages~A and~B both close on the same column with the same
        criterion, ratifying the activation predicate that Stage~C
        will reuse without modification.
  \item The two staggered-loop constants
        (\texttt{MAX\_STAGGERED\_ITER}~$= 4$,
         \texttt{STAGGERED\_RELAX}~$= 0.7$) are shared verbatim between
        Stage~B and \path{main_lshaped_multiscale.cpp}, so the
        algorithmic surface is already cross-validated.
\end{itemize}

\section{Authorisation gate}

The heavy MYG004 dynamic run will only be triggered when the user
explicitly authorises the compute. At that point:
\begin{enumerate}
  \item The criterion is wired into the staging policy of
        \path{main_lshaped_multiscale_16storey.cpp}.
  \item A new audit JSON \texttt{audit\_stage\_c\_16storey\_dynamic.json}
        will be emitted alongside Cap.~96 and Cap.~97 audits.
  \item This chapter will be promoted from
        \texttt{scoped\_deferred} to \texttt{closed\_with\_runtime\_evidence}.
\end{enumerate}

Until that gate is opened, Stage~C remains documented but not executed.
